---
title: "12_prepare_clinical_dat"
author: "Tobias Hoch and Daniel Schulz"
date: "2021-02-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction
In this script, we load the clinical data and clean and prepare the data for further analysis. Furthermore, we create two .csv files containing the clinical data for RNA and Protein TMA separately in order to attach them to the SingleCellExperiment in a later step. 

# Preparations

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message= FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Load libraries
```{r}
library(dplyr)
library(lubridate)
library(stringr)
```

## Load the data

```{r}
# load clinical data and TMA data to link images with patients
clinical_mat <- read.csv(file = "data/data_for_analysis/200323_TMA_256_clinical_data_March2020.csv",
                         sep=",",stringsAsFactors = FALSE)
clinical_mat <- clinical_mat[is.na(clinical_mat$SpotNr) == FALSE,]

spot_mat <- read.csv(file = "data/data_for_analysis/191119_TMA_spotmatrix.csv", sep= ";",stringsAsFactors = FALSE)

# remove missing value rows from spot_mat
spot_mat <- spot_mat[is.na(spot_mat$SpotNr) == FALSE,]
# in the spot_mat  BlockID "B2017.47224" has a space incorporated: "B2017.47224 ". We will rename this BlockID here.
spot_mat[which(spot_mat$BlockID == "B2017.47224 "),]$BlockID <- "B2017.47224"

# load image metadata as output from CellProfiler
image_mat_prot <- read.csv(file = "data/data_for_analysis/protein/Image.csv", stringsAsFactors = FALSE)
image_mat_rna <- read.csv(file = "data/data_for_analysis/rna/Image.csv", stringsAsFactors = FALSE)
```

# Data cleaning

## Assign clean column names

```{r clean up the column names}
colnames(clinical_mat)

clean_names <- c("SpotNr","PatientID","BlockID","IHC_T_score","Age_range","Gender",
                 "MM_location","Primary_melanoma_type","Cancer_Stage","Mutation",
                 "Nr_treatments_before_surgery","Name_last_drug_used","Last_sys_treatment_before_surgery",
                 "Last_prev_therapy","Date_surgery","Drug_after_surgery","Treatment_after_surgery","Adjuvant",
                 "Start_treatment","End_treatment","Reason_to_stop_treatment","Date_progression",
                 "Status_at_3m","Response_at_3m","Time_of_response_assesment","Response_at_6m",
                 "Response_at_12m","Last_PET","Date_death")

colnames(clinical_mat) <- clean_names

# empty but non NA date_progression is equal to no progression
clinical_mat <- data.frame(clinical_mat)
clinical_mat$relapse <- NA
clinical_mat[is.na(clinical_mat$Date_progression) & is.na(clinical_mat$Nr_treatments_before_surgery) == FALSE,]$relapse <- "untreated/lost"
clinical_mat[clinical_mat$Date_progression %in% c("") & is.na(clinical_mat$Nr_treatments_before_surgery) == FALSE,]$relapse <- "no relapse"
clinical_mat[is.na(clinical_mat$relapse) & is.na(clinical_mat$Nr_treatments_before_surgery) == FALSE,]$relapse <- "relapse"
clinical_mat[is.na(clinical_mat$Nr_treatments_before_surgery),]$relapse <- "control"
```

# Calculate survival data
Starting point is the beginning of the surgery, end point is death/progression or last PET

## Convert Characters to Date

```{r}
clinical_mat$Date_surgery <- as.Date(clinical_mat$Date_surgery, format = "%d.%m.%Y")
clinical_mat$Date_death <- as.Date(clinical_mat$Date_death, format = "%d.%m.%Y")
clinical_mat$Date_progression <- as.Date(clinical_mat$Date_progression, format = "%d.%m.%Y")
clinical_mat$Last_PET <- as.Date(clinical_mat$Last_PET, format = "%d.%m.%Y")
clinical_mat$Start_treatment <- as.Date(clinical_mat$Start_treatment, format = "%d.%m.%Y")
clinical_mat$End_treatment <- as.Date(clinical_mat$End_treatment, format = "%d.%m.%Y")
```

## Calculate time until last PET or death

```{r}
clinical_mat$Time_to_death_or_last_PET <- ifelse(is.na(clinical_mat$Date_death) == TRUE, # if no death, then
                                                   clinical_mat$Last_PET - clinical_mat$Start_treatment, # time to last PET
                                                   clinical_mat$Date_death - clinical_mat$Start_treatment) # else: time to progression

clinical_mat$censoring_death <- ifelse(is.na(clinical_mat$Date_death) == TRUE, 0, 1) # 0: no Death, 1: Death
```

## Calculate time until last PET or relapse

```{r}
clinical_mat$Time_to_progression_or_last_PET <- ifelse(is.na(clinical_mat$Date_progression) == TRUE, # if no progression, then
                                                   clinical_mat$Last_PET - clinical_mat$Start_treatment, # time to last PET
                                                   clinical_mat$Date_progression - clinical_mat$Start_treatment) # else: time to progression

clinical_mat$censoring_progression <- ifelse(is.na(clinical_mat$Date_progression) == TRUE, 0, 1) # 0: no Relapse, 1: Relapse
```

# Combine with RNA/Protein image_mat

## RNA clinical matrix

```{r}
# SpotNr and Description in one mat
full_mat_rna <- spot_mat[,c("SpotNr", "BlockID", "Description", "TissueType", "Location")]

# add respective ImageNumber
image_mat_rna$Description <- image_mat_rna$Metadata_Description
full_mat_rna <- left_join(full_mat_rna, image_mat_rna[,c("ImageNumber", "Description")], by = "Description")

# join with clinical_mat
full_mat_rna <- left_join(full_mat_rna, clinical_mat[,-1], by="BlockID")

# remove Images that were not acquired or removed after initial processing (missing on TMA or too bad quality)
full_mat_rna <- full_mat_rna[is.na(full_mat_rna$ImageNumber) == FALSE,]
```

## Protein clinical matrix

```{r}
# SpotNr and Description in one mat
full_mat_prot <- spot_mat[,c("SpotNr", "BlockID", "Description", "TissueType", "Location")]

image_mat_prot$Description <- image_mat_prot$Metadata_Description

# rename "G1 - split" row in image_mat (this core was acquired in two measurements because the machine stopped due to an error)
image_mat_prot$Description <- ifelse(image_mat_prot$Description == "G1 - split", "G1", image_mat_prot$Description)

# add respective ImageNumber
full_mat_prot <- left_join(full_mat_prot, image_mat_prot[,c("ImageNumber", "Description")], by = "Description")

# join with clinical_mat
full_mat_prot <- left_join(full_mat_prot, clinical_mat[,-1], by = "BlockID")

# remove Images that were not acquired or removed after initial processing (missing on TMA or too bad quality)
full_mat_prot <- full_mat_prot[is.na(full_mat_prot$ImageNumber) == FALSE,]
```

## compare both matrices

```{r}
# check if the BlockID/PatientID of Description is the same in both data sets
rna_sub <- full_mat_rna[,c("Description", "BlockID", "PatientID")]
prot_sub <- full_mat_prot[,c("Description", "BlockID", "PatientID")]

compare <- left_join(prot_sub, rna_sub, by = "Description")
all(compare$BlockID.x == compare$BlockID.y)
unique(compare$PatientID.x == compare$PatientID.y)
```

## remove exact dates

```{r}
# remove exact date in clinical data - Protein
full_mat_prot$Date_death <- format(as.Date(full_mat_prot$Date_death),"01-%b-20%y")
full_mat_prot$Date_surgery <- format(as.Date(full_mat_prot$Date_surgery),"01-%b-20%y")
full_mat_prot$Date_progression <- format(as.Date(full_mat_prot$Date_progression),"01-%b-20%y")
full_mat_prot$Start_treatment <- format(as.Date(full_mat_prot$Start_treatment),"01-%b-20%y")
full_mat_prot$End_treatment <- format(as.Date(full_mat_prot$End_treatment),"01-%b-20%y")
full_mat_prot$Last_PET <- format(as.Date(full_mat_prot$Last_PET),"01-%b-20%y")
full_mat_prot$Last_prev_therapy <- str_replace_all(full_mat_prot$Last_prev_therapy, "-", ".")
full_mat_prot$Last_prev_therapy <- format(as.Date(my(full_mat_prot$Last_prev_therapy)),"01-%b-20%y")

# remove exact date in clinical data - RNA
full_mat_rna$Date_death <- format(as.Date(full_mat_rna$Date_death),"01-%b-20%y")
full_mat_rna$Date_surgery <- format(as.Date(full_mat_rna$Date_surgery),"01-%b-20%y")
full_mat_rna$Date_progression <- format(as.Date(full_mat_rna$Date_progression),"01-%b-20%y")
full_mat_rna$Start_treatment <- format(as.Date(full_mat_rna$Start_treatment),"01-%b-20%y")
full_mat_rna$End_treatment <- format(as.Date(full_mat_rna$End_treatment),"01-%b-20%y")
full_mat_rna$Last_PET <- format(as.Date(full_mat_rna$Last_PET),"01-%b-20%y")
full_mat_rna$Last_prev_therapy <- str_replace_all(full_mat_rna$Last_prev_therapy, "-", ".")
full_mat_rna$Last_prev_therapy <- format(as.Date(my(full_mat_rna$Last_prev_therapy)),"01-%b-20%y")
```

# Save .csv Files

```{r}
write.csv(x = full_mat_prot, file = "data/data_for_analysis/protein/clinical_data_protein.csv", row.names = F)
write.csv(x = full_mat_rna, file = "data/data_for_analysis/rna/clinical_data_RNA.csv", row.names = F)
```

