---
title: "01_RNA_read_data"
author: "toobiwankenobi"
date: "2020-07-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Preparations
This file will load the single-cell data and store it in an SingleCellExperiment data container. In order to successfully run this script, several .csv files have to provided in the data folder of this repository.

## Load Libraries

```{r warning=F, message=F}
library(SingleCellExperiment)
library(dplyr)
library(data.table)
library(ggplot2)
library(data.table)
```

## Load data (DOWNLOAD DATA SET AND STORE FILES IN CORRESPONDING FOLDER WITH CORRECT NAMING)

```{r}
# load cell data
cells <- data.frame(fread(file = "data/rna/cell.csv",stringsAsFactors = FALSE))

# load image metadata as output from CellProfiler
image_mat <- read.csv(file = "data/rna/Image.csv", stringsAsFactors = FALSE)

# load panel information as exported from AirLab and modified to contain ilastik and full columns and nicely written Target names.
panel_mat <- read.csv(file = "data/rna/panel_mat.csv", stringsAsFactors = FALSE )

# load the channel order information. this information is created from the jupyter notebook in the tiffs folder for every single tiff that is created. 
# since it is the same panel for all images we just select one single file. the order or the metals in this file is the actual channels order.
tags <- read.csv( "data/rna/20190731_ZTMA256.1_slide2_TH_s1_p14_r1_a1_ac_full.csv", header = FALSE)

# load clinical data and TMA data to link images with patients
clinical_mat <- read.csv(file = "data/rna/clinical_data_RNA.csv",stringsAsFactors = FALSE)
```

# Pre-processing

## Generate the counts data frame

```{r}
cur_counts <- cells[,grepl("Intensity_MeanIntensityCorrected_FullStackFiltered",colnames(cells))]
```

## Get the scaling factor
the single cell data needs to be multiplied with the scaling factor (16 bit)

```{r}
cur_counts <- cur_counts * image_mat$Scaling_FullStack[1]

# order the channels according to channel number
channelNumber <- as.numeric(sub("^.*_c", "", colnames(cur_counts)))
cur_counts <- cur_counts[,order(channelNumber,decreasing = FALSE)]
```

## Get the information whether a cell is in the tumor mask
any cell that has more than 25 % of its Area in the tumor mask is considered as "TRUE" meaning inside the tumor.

```{r}
tumor_mask <- cells$Intensity_MeanIntensity_tumormask * image_mat$Scaling_FullStack[1]
in_tumor <- tumor_mask > 0.25
```

## Prepare the cell level metadata (colData)

```{r}
cell_meta <- DataFrame(CellNumber = cells$ObjectNumber,
                       ImageNumber = cells$ImageNumber,
                       Center_X = cells$Location_Center_X,
                       Center_Y = cells$Location_Center_Y,
                       Area = cells$AreaShape_Area,
                       MajorAxisLength = cells$AreaShape_MajorAxisLength,
                       MinorAxisLength = cells$AreaShape_MinorAxisLength,
                       NumberOfNeighbors = cells$Neighbors_NumberOfNeighbors_8,
                       in_tumor = in_tumor)

# add a unique cellID to each cell consisting of "dataset"+"ImageNumber"+"ObjectNumber"
cell_meta$cellID <- paste0("RNA_",cell_meta$ImageNumber, "_",cell_meta$CellNumber)
rownames(cell_meta) <- cell_meta$cellID

# order according to ImageNumber
cell_meta <- cell_meta[order(cell_meta$ImageNumber),]
```

## Prepare the row-level metadata (panel/marker information)
here we prepare all the metadata for the rows in the single cell experiment object (rowData)

```{r}
# the channel numbers are the rownumbers in the "tags" file that we create above
tags$channel <- as.numeric(rownames(tags))
colnames(tags) <- c("Metal.Tag","channel")

# include the channel information in the panel metadata (panel_mat)
panel_mat <- merge(panel_mat,tags,by="Metal.Tag")

# now we order the panel metadata by channel. therefore we first modify the column names
panel_mat <- panel_mat[order(panel_mat$channel,decreasing = FALSE),]

# rename CD8a -> CD8 in Targets
panel_mat$Target[26] <- "CD8"

# assign rownames
rownames(panel_mat) <- panel_mat$Target
```

# Create SCE object

## Create the single cell experiment object

```{r}
# create the SCE object
sce <- SingleCellExperiment(assays = list(counts = t(cur_counts)))

# Set marker name as rownames and cellID as colnames
rownames(sce) <- rownames(panel_mat)
colnames(sce) <- rownames(cell_meta)

# add the column and row metadata
colData(sce) <- cell_meta
rowData(sce) <- panel_mat

# we also generate here the generically implemented "logcounts" as asinh transformed counts
assay(sce, "asinh") <- asinh(SingleCellExperiment::counts(sce))
```

## Assign the clinical data to the metadata slot

```{r}
# order according to ImageNumber
clinical_mat <- clinical_mat[order(clinical_mat$ImageNumber),]
metadata(sce) <- as.list(clinical_mat)
```

## Save the SCE object

```{r}
saveRDS(sce,file = "data/sce_RNA.rds")
```
