---
title: "03_RNA_quality_control"
author: "Nils Eling/Daniel Schulz/Tobias Hoch"
date: "2020-07-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Preparations

## Load libraries

```{r message=F}
sapply(list.files("code/helper_functions", full.names = TRUE), source)
library(SingleCellExperiment)
library(dplyr)
library(ggplot2)
library(scater)
library(CATALYST)
library(reshape2)
library(LSD)
library(viridis)
library(ggridges)
library(cowplot)
library(BiocParallel)
library(dittoSeq)
```

## Load the single cell experiment object and the image metadata

```{r}
sce <- readRDS(file = "data/sce_RNA.rds")
```

# Assay

## Add different assays

```{r}
assay(sce, "scaled_counts") <- t(scale(t(assay(sce, "counts"))))
assay(sce, "scaled_asinh") <- t(scale(t(assay(sce, "asinh"))))
```

# General Plots

## Plot Cell Counts for every Image

```{r cell_counts_over_samples, fig.width=15, fig.height=10}
# this function takes all the column metadata from the sce and plots parts thereof
plotCellCounts(sce, colour_by = "Location", split_by = "ImageNumber", imageID = "ImageNumber")
```

## Image 57 has only a few cells and should probably be excluded.
will be flagged below

```{r images with less than 500 cells}
cur_sce <- data.frame(colData(sce))
# show images with less than 500 cells
cur_sce %>%
  group_by(ImageNumber) %>%
  summarise(n=n()) %>%
  filter(n<500)
```

## Flag image in sce object for future exclusion

```{r flag images}
# define vector for each single cell whether to keep (TRUE) or not (FALSE)
includeImage <- colData(sce)$ImageNumber != 57
sce$includeImage <- includeImage
```

## Mean intensity of markers per image

```{r mean marker intensities over images}
# we use a function from Nils. This function makes use of the aggregate function to calculate the mean for each channel over all specified groups
mean_sce <- calculateSummary(sce, split_by = c("ImageNumber", "BlockID", "Location","Mutation","Cancer_Stage", "Status_at_3m","E_I_D","Adjuvant"), exprs_values = "counts")
``` 

## Transform data

```{r transform mean data}
assay(mean_sce, "asinh") <- asinh(assay(mean_sce, "meanCounts"))
assay(mean_sce, "asinh_scaled") <- t(scale(t(asinh(assay(mean_sce, "meanCounts")))))
```

## Plot the mean data

```{r  mean_per_image, fig.height= 10, fig.width=15}
# first we define a vector of markers that we want to plot
plot_targets <- rownames(sce)
plot_targets <- plot_targets[! plot_targets %in% c("DNA1","DNA2","HistoneH3")]

# now we plot the heatmap
plotHeatmap(mean_sce,features = plot_targets  ,exprs_values = "asinh",colour_columns_by = "ImageNumber",color = viridis(100))
```

## Plot the scaled data

```{r mean_per_image_scaled, fig.height=10, fig.width=15}
# now we plot the scaled heatmap
plotHeatmap(mean_sce,features = plot_targets, exprs_values = "asinh_scaled", colour_columns_by = c("ImageNumber"), zlim = c(-3,3),
            color = colorRampPalette(c("dark blue", "white", "dark red"))(100))
```

## Cell level quality control
here we plot the marker intensity distributions for all images. since we have too many images we make groups of 10.

```{r density_plots_all_markers, message=F, fig.height= 10, fig.width=15}
y <- c(rep(1:10,16),rep(11,7))

# add the group information to the sce object
sce$groups <- y[colData(sce)$ImageNumber]

# now we use the function written by Nils
plotDist(sce, plot_type = "ridges", 
         colour_by = "groups", split_by = "rows", 
         exprs_values = "asinh") + 
  theme_minimal(base_size = 15)
 # the distributions look very even across images indicating that we have no major batch effects.

```

## Define markers which had poor staining
By visual inspection, we defined bad markers

```{r define good markers}
rowData(sce)$good_marker <- ! grepl("DNA|Histone|Vimentin|PD1|PDL1|CCR2|CK5",rownames(sce))

# good_marker without RNA channels
rowData(sce)$noRNA_marker <- ! grepl("DNA|Histone|Vimentin|PD1|PDL1|CCR2|CK5|CXCL|CCL|DapB",rownames(sce))
```

## Calculate UMAP

```{r}
set.seed(12345)

# UMAP
start = Sys.time()
sce <- runUMAP(sce, exprs_values = "scaled_counts", 
                     subset_row = rowData(sce)$good_marker)
end = Sys.time()
print(end-start)
```

## Subset SCE for UMAP visualization

```{r}
cur_sce <- sce[, colnames(sce) %in% sample(sce$cellID, round(length(sce$cellID)*0.05))]
cur_sce$ImageNumber <- as.character(cur_sce$ImageNumber)
```

## Visualize features on and UMAP
Next, we will visualize different quality features on these representations.

## UMAP

```{r UMAP_clinical_features, fig.width=35, fig.height=25, message=FALSE}
# Select plots in list
p.list <- list()

# 
p.list$ImageNumber <- dittoDimPlot(cur_sce, var = "ImageNumber", reduction.use = "UMAP", size = 0.5, legend.show = FALSE) 
p.list$Mutation <- dittoDimPlot(cur_sce, var = "Mutation", reduction.use = "UMAP", size = 0.5)
p.list$Cancer_Stage <- dittoDimPlot(cur_sce, var = "Cancer_Stage", reduction.use = "UMAP", size = 0.5)
p.list$relapse <- dittoDimPlot(cur_sce, var = "relapse", reduction.use = "UMAP", size = 0.5)
p.list$Location <- dittoDimPlot(cur_sce, var = "Location", reduction.use = "UMAP", size = 0.5)
p.list$TissueType <- dittoDimPlot(cur_sce, var = "TissueType", reduction.use = "UMAP", size = 0.5)
p.list$MM_location_simplified <- dittoDimPlot(cur_sce, var = "MM_location_simplified", reduction.use = "UMAP", size = 0.5)
p.list$treatment_group_before_surgery <- dittoDimPlot(cur_sce, var = "treatment_group_before_surgery", reduction.use = "UMAP", size = 0.5)


plot_grid(plotlist = p.list, ncol = 4, rel_widths = c(1.5, 1, 1, 1))
```

## Marker Expression on UMAP

```{r asinh_UMAP_marker, fig.width=35, fig.height=25, message=FALSE}
p.list <- list()
for(i in rownames(sce)[rowData(cur_sce)$good_marker]){
  p.list[[i]] <- plotUMAP(cur_sce, colour_by = i, by_exprs_values = "asinh")
}

plot_grid(plotlist = p.list, ncol = 7)
```

## Scaled Expression UMAP

```{r scaled_asinh_UMAP_marker, fig.width=35, fig.height=25, message=FALSE}
p.list <- list()
for(i in rownames(sce)[rowData(cur_sce)$good_marker]){
  p.list[[i]] <- plotUMAP(cur_sce, colour_by = i, by_exprs_values = "scaled_asinh")
}


plot_grid(plotlist = p.list, ncol = 7)
```

# Save data

## Save updated SCE object

```{r save updated sce file}
saveRDS(sce, file = "data/sce_RNA.rds")
```
