---
title: "cytotoxic T cell quantification"
author: "Daniel"
date: "July 2020"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message= FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Load libraries

```{r load packages}
sapply(list.files("code/helper_functions/", full.names = TRUE), source)
library(LSD)
library(SingleCellExperiment)
library(ggplot2)
library(scater)
library(viridis)
library(igraph)
library(CATALYST)
library(reshape2)
library(cowplot)
library(ggridges)
library(pheatmap)
library(tidyverse)
```

## Load Data

```{r load data, echo=FALSE}
sce = readRDS(file = "data/sce_protein.rds")
sce_rna = readRDS(file = "data/sce_RNA.rds")

image_mat <- read.csv("data/protein/Image.csv")
```

# Cytotoxic T cell Scoring

## T Cell Density
We want to calculate the T cell density.
we calculate this per mm2. We use the count for cytotxic T cells as defined under “03_cell_type_definition” script

```{r}
cur_df <- data.frame(celltype = sce$celltype,
                             ImageNumber = sce$ImageNumber)

cellcount <- (t(table(cur_df)))

# here we get the imagesize from the image metadata
im_size <- (image_mat$Height_cellmask * image_mat$Width_cellmask)/1000000


cellcount <- as.data.frame(cellcount)

# we calculate the density for each celltype for 1 mm2
cellcount$density <- cellcount$Freq/im_size[cellcount$ImageNumber]

# there are roughly 60 images with 50 or less cytotoxic T cells
hist(cellcount[cellcount$celltype == "Tcytotoxic",]$density,breaks = 100)
```

## T cell density score per image
images with up to 40 cytotoxic T cells/mm2 are called “absent” images with more than 40 and up to 100 cytotoxic T cells/mm2 are called “low” images with more than 100 and up to 400 cytotoxic T cells/mm2 are called “medium” images with more than 400 cytotoxic T cells/mm2 are called “high”

```{r}
# define a vector with all ImgeNumbers
T_density_scores <-c(1:length(unique(cellcount$ImageNumber)))

T_density_scores  <- rep("unassigned",length(unique(cellcount$ImageNumber)))

T_absent <- which(cellcount[cellcount$celltype == "Tcytotoxic",]$density <= 40)
T_low <- which(cellcount[cellcount$celltype == "Tcytotoxic",]$density > 40 & cellcount[cellcount$celltype == "Tcytotoxic",]$density <= 100)
T_med <- which(cellcount[cellcount$celltype == "Tcytotoxic",]$density > 100 & cellcount[cellcount$celltype == "Tcytotoxic",]$density <= 400)
T_high <- which(cellcount[cellcount$celltype == "Tcytotoxic",]$density > 400)

T_density_scores[T_absent] <- "absent"
T_density_scores[T_low] <- "low"
T_density_scores[T_med] <- "med"
T_density_scores[T_high] <- "high"

# now we add the information to the single cell experiment
sce$Tcell_density_score_image <- T_density_scores[sce$ImageNumber]
```

## Fraction of cytotoxic T cells per ImageNumber

we group images by the fraction of cytotoxic T cells. we calculate the fractions and then make three even groups: low, medium, high by simply splitting the sorted data.

```{r T cell fraction score per ImageNumber}
cur_df <- tibble(celltype = sce$celltype,
                 ImageNumber = sce$ImageNumber)

cellcount <- as_tibble(t(table(cur_df)))

cellcount <- cellcount %>%
  group_by(ImageNumber) %>%
  mutate(total = sum(n))

cellcount <- cellcount %>%
  filter(celltype == "Tcytotoxic")

cellcount <- cellcount %>%
  mutate(Tcytotoxic_frac =n/total)

# here we generate three classes based on ranks
cellcount$rank <- ntile(cellcount$Tcytotoxic_frac,n=3)

# the lowest third of cytotoxic T cell fraction BlockIDs get grouped
cellcount$T_frac_score <- ""
cellcount[which(cellcount$rank == 1),]$T_frac_score <- "low"
cellcount[which(cellcount$rank == 2),]$T_frac_score <- "med"
cellcount[which(cellcount$rank == 3),]$T_frac_score <- "high"

T_frac_score <- cellcount$T_frac_score
names(T_frac_score) <- cellcount$ImageNumber

# add to SCE
sce$T_frac_score_per_ImageNumber <- T_frac_score[sce$ImageNumber]
```

## Fraction of cytotoxic T cells per BlockID

we group images by the fraction of cytotoxic T cells. we calculate the fractions and then make three even groups: low, medium, high by simply splitting the sorted data.

```{r T cell fraction score per BlockID}
cur_df <- tibble(celltype = sce$celltype,
                 BlockID = sce$BlockID)

cellcount <- as_tibble(t(table(cur_df)))

cellcount <- cellcount %>%
  group_by(BlockID) %>%
  mutate(total = sum(n))

cellcount <- cellcount %>%
  filter(celltype == "Tcytotoxic")


cellcount <- cellcount %>%
  mutate(Tcytotoxic_frac =n/total)

# here we generate three classes based on ranks
cellcount$rank <- ntile(cellcount$Tcytotoxic_frac,n=3)

# the lowest third of cytotoxic T cell fraction BlockIDs get grouped
cellcount$T_frac_score <- ""
cellcount[which(cellcount$rank == 1),]$T_frac_score <- "low"
cellcount[which(cellcount$rank == 2),]$T_frac_score <- "med"
cellcount[which(cellcount$rank == 3),]$T_frac_score <- "high"

T_frac_score <- cellcount$T_frac_score
names(T_frac_score) <- cellcount$BlockID

# add to SCE
sce$T_frac_score_per_BlockID <- T_frac_score[sce$BlockID]
```

## Fraction of cytotoxic T cells per PatientID

we group images by the fraction of cytotoxic T cells. we calculate the fractions and then make three even groups: low, medium, high by simply splitting the sorted data.

```{r T cell fraction score per PatientID}
pat_ID <- metadata(sce)$Internal_pat_ID
sce$PatientID <- pat_ID[sce$ImageNumber]

cur_df <- tibble(celltype = sce$celltype,
                 PatientID = sce$PatientID)

cellcount <- as_tibble(t(table(cur_df)))

cellcount <- cellcount %>%
  group_by(PatientID) %>%
  mutate(total = sum(n))

cellcount <- cellcount %>%
  filter(celltype == "Tcytotoxic")


cellcount <- cellcount %>%
  mutate(Tcytotoxic_frac =n/total)

# here we generate three classes based on ranks
cellcount$rank <- ntile(cellcount$Tcytotoxic_frac,n=3)

# the lowest third of cytotoxic T cell fraction BlockIDs get grouped
cellcount$T_frac_score <- ""
cellcount[which(cellcount$rank == 1),]$T_frac_score <- "low"
cellcount[which(cellcount$rank == 2),]$T_frac_score <- "med"
cellcount[which(cellcount$rank == 3),]$T_frac_score <- "high"

T_frac_score <- cellcount$T_frac_score
names(T_frac_score) <- cellcount$PatientID

sce$T_frac_score_per_PatientID <- T_frac_score[sce$PatientID]
```

## E_I_D score compared to T_frac_score_per_ImageNumber

```{r check the cut offs}
cur_df <- data.frame(T_density = sce$T_frac_score_per_ImageNumber,
                     E_I_D = sce$E_I_D)

table(cur_df)
```

__cells from images classified as deserted mostly belong to the absent class defined here. However, in the absent class defined here many cells from images belong to either inflamed or excluded classes.__


## Develope a E_I_D score based on the images in this cohort

we will therefore make use of the tumor mask that we generated using ilastik. Every cell has a tag which assigns it to being within the tumor or not.

We will define as follows:
Deserted: No T cells - same as infiltration = absent -> less than 40 cytotxic T cells/mm2
Excluded: infiltration status == low | med | high. ratio of T cells inside the tumor to outside the tumor < 0.5
Inflamed: infiltration status == low | med | high. ratio of T cells inside the tumor to outside the tumor > 0.5 or if more than 100 Tcells are inside the tumor.

```{r develope E_I_D score based on this data}
# we next calculate the amount of T cells inside and outside the tumor for each image
cur_df <- data.frame(inside_tumor = sce$in_tumor,
                     E_I_D = sce$E_I_D,
                     ImageNumber = sce$ImageNumber,
                     celltype = sce$celltype,
                     Location = sce$Location,
                     Tcell_density_score_image = sce$Tcell_density_score_image)

# reshaping
cur_df <- reshape2::dcast(cur_df,formula = "ImageNumber + E_I_D + inside_tumor + Tcell_density_score_image + Location ~ celltype")
cur_df <- cur_df[,c("ImageNumber","E_I_D","inside_tumor", "Tcell_density_score_image","Tcytotoxic", "Location")]
cur_df <- reshape2::dcast(cur_df,formula = "ImageNumber + E_I_D + Tcell_density_score_image + Location ~ inside_tumor", value.var = "Tcytotoxic")
colnames(cur_df) <- c("ImageNumber" , "E_I_D" ,"Tcell_density_score_image", "Location", "outside_tumor", "inside_tumor")

# convert NAs to 0
cur_df[is.na(cur_df$outside_tumor),]$outside_tumor <- 0

# the ratio still has "Inf" and "NAns" values due to 0 values in the data. we will take care of this below
cur_df$infiltration_ratio <- cur_df$inside_tumor/cur_df$outside_tumor

cur_df$infiltration <- "unassigned"

# for all samples that have abenst Tcell_density_score_image we also enter the status "Deserted"
cur_df[cur_df$Tcell_density_score_image == "absent",]$infiltration <- "Deserted"

# for samples with low | med | high Tcell_density_score_image we calculate the ratio between Tcells inside the tumor mask and outside the tumor mask. If this ratio is < 0.5 we  call the tumor excluded. It >0.5 we call it inflamed.
cur_df[(cur_df$Tcell_density_score_image %in% c("low","med","high") & cur_df$infiltration_ratio < 0.5),]$infiltration <- "Excluded"

cur_df[cur_df$Tcell_density_score_image %in% c("low","med","high") & cur_df$infiltration_ratio > 0.5 | cur_df$inside_tumor > 200,]$infiltration <- "Inflamed"

# control samples
cur_df[cur_df$Location == "CTRL",]$infiltration <- "Control"

# incorporate the information into the single cell experiment
sce$infiltration <- cur_df$infiltration[sce$ImageNumber]
```

## Add Scores to RNA data set

```{r export scores for RNA dataset}
description_data <- data.frame(colData(sce)) %>%
  distinct(Description, .keep_all = TRUE)

col_rna <- data.frame(colData(sce_rna))

# left_join
col_rna <- left_join(col_rna, description_data[,c("Description", "T_frac_score_per_ImageNumber", "T_frac_score_per_BlockID",
                                                  "T_frac_score_per_PatientID", "infiltration")])

# add to sce
sce_rna$T_frac_score_per_ImageNumber <- col_rna$T_frac_score_per_ImageNumber
sce_rna$T_frac_score_per_BlockID <- col_rna$T_frac_score_per_BlockID
sce_rna$T_frac_score_per_PatientID <- col_rna$T_frac_score_per_PatientID
sce_rna$infiltration <- col_rna$infiltration
```

## Write .csv for Levesque group

```{r}
test <- data.frame(colData(sce)) %>%
  select(Description, T_frac_score_per_ImageNumber, infiltration) %>%
  distinct(Description, .keep_all = T) %>%
  write.csv(.,"data/fraction_and_infiltration_scoring.csv")
```


## Save updated SCE

```{r save sce}
saveRDS(sce,file = "data/sce_protein.rds")
saveRDS(sce_rna,file = "data/sce_RNA.rds")
```
