---
title: "TLS identifiaction"
author: "Tobias Hoch"
date: "2020-05-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

This script detects interaction networks of a given celltype (here: B cells) and defines these networks as clusters. Once a cluster is defined, an algorithm screens the neighbourhood of those clusters to identify cells within/surrounding a cluster. These cells are defined as the community of a cluster. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message= FALSE)
```

## Load packages and helper functions

```{r load packages}
sapply(list.files("code/helper_functions", full.names = TRUE), source)
library(SingleCellExperiment)
library(ggplot2)
library(scater)
library(viridis)
library(igraph)
library(CATALYST)
library(reshape2)
library(cowplot)
library(ggridges)
library(tidyverse)
library(viridis)
library(dplyr)
library(cytomapper)
library(concaveman)
library(data.table)
library(sf)
library(ggbeeswarm)
library(RANN)
```

## Load data

```{r load data}
sce_prot = readRDS(file = "data/data_for_analysis/sce_protein.rds")
sce_rna = readRDS(file = "data/data_for_analysis/sce_RNA.rds")
```

## Find patches of B cells and their milieus

A B cell cluster is defined by at least 20 adjacent B cells (Bcell and BnTcell, max distance of 15µm between them). A milieu is defined by all cells within a cluster and the proximity (enlarging distance = 15µm)

```{r Cluster and Community Detection}
sce_prot$bcell_patch <- NULL
sce_prot$bcell_milieu <- NULL
sce_prot$bcell_patch_score <- NULL

start = Sys.time()

# quantiles of cell radius
quantile(sqrt(sce_prot[,sce_prot$celltype %in% c("B cell")]$Area/pi))

# find B cell clusters
sce_prot <- findPatch(sce_prot, sce_prot[,colData(sce_prot)$celltype %in% c("B cell", "BnT cell")]$cellID, 
                    'cellID', 'Center_X', 'Center_Y', 'ImageNumber', 
                    distance = 15, min_clust_size = 10, output_colname = "bcell_patch")

# number of B cell clusters
length(unique(sce_prot$bcell_patch))

# define cells within/surrounding a cluster of B cells
sce_prot <- findMilieu(sce_prot, 
                     'cellID', 'Center_X', 'Center_Y', 'ImageNumber', 'bcell_patch', 
                     distance = 30, output_colname = "bcell_milieu")

# number of chemokine communities
length(unique(sce_prot$bcell_milieu))
end = Sys.time()
print(end-start)
```

## Plot some patches

```{r example_patch, fig.width=10, fig.height=10, dev=c('pdf')}
example <- findPatch(sce_prot[,sce_prot$Description %in% c("D4")], sce_prot[,sce_prot$celltype %in% c("B cell", "BnT cell")]$cellID, 
                    'cellID', 
                    'Center_X', 'Center_Y', 
                    'ImageNumber', 
                    distance = 15, 
                    min_clust_size = 10,
                    output_colname = "example_patch")

example <- findMilieu(example, 
              'cellID', 
              'Center_X', 'Center_Y', 
              'ImageNumber', 
              'example_patch', 
              distance = 30,
              output_colname = "example_milieu",
              plot = TRUE)
```

## Save SCE object

```{r save sce}
saveRDS(sce_prot, file = "data/data_for_analysis/sce_protein.rds")
saveRDS(sce_rna, file = "data/data_for_analysis/sce_RNA.rds")
```

