---
title: "TLS identifiaction"
author: "Tobias Hoch"
date: "2020-05-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

This script detects interaction networks of a given celltype (here: B cells) and defines these networks as clusters. Once a cluster is defined, an algorithm screens the neighbourhood of those clusters to identify cells within/surrounding a cluster. These cells are defined as the community of a cluster. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message= FALSE)
```

## Load packages and helper functions

```{r load packages}
sapply(list.files("code/helper_functions", full.names = TRUE), source)
library(SingleCellExperiment)
library(ggplot2)
library(scater)
library(viridis)
library(igraph)
library(CATALYST)
library(reshape2)
library(cowplot)
library(ggridges)
library(tidyverse)
library(viridis)
library(dplyr)
library(cytomapper)
library(sp)
library(sf)
library(ggbeeswarm)
library(RANN)
```

## Load data

```{r load data}
sce_prot = readRDS(file = "data/sce_protein.rds")
sce_rna = readRDS(file = "data/sce_RNA.rds")

image_prot <- read.csv("data/protein/Image.csv")
image_rna <- read.csv("data/rna/Image.csv")
```

## Find patches of B cells and their milieus

A B cell cluster is defined by at least 20 adjacent B cells (Bcell and BnTcell, max distance of 15µm between them). A milieu is defined by all cells within a cluster and the proximity (enlarging distance = 15µm)

```{r Cluster and Community Detection}
sce_prot$bcell_patch <- NULL
sce_prot$bcell_milieu <- NULL
sce_prot$bcell_patch_score <- NULL

start = Sys.time()
# quantiles of cell radius
quantile(sqrt(sce_prot[,sce_prot$celltype %in% c("Bcell")]$Area/pi))

# find B cell clusters
sce_prot <- findClusters(sce_prot, sce_prot[,colData(sce_prot)$celltype %in% c("Bcell", "BnTcell")]$cellID, 
                    'cellID', 'Center_X', 'Center_Y', 'ImageNumber', 
                    distance = 15, min_clust_size = 20, output_colname = "bcell_patch")

# number of B cell clusters
length(unique(sce_prot$bcell_patch))

# define cells within/surrounding a cluster of B cells
sce_prot <- findCommunity(sce_prot, 
                     'cellID', 'Center_X', 'Center_Y', 'ImageNumber', 'bcell_patch', 
                     distance = 50, output_colname = "bcell_milieu")

# number of chemokine communities
length(unique(sce_prot$bcell_milieu))
end = Sys.time()
print(end-start)
```

# Define grouping for B cell / patch densities

## Cell Densities 

```{r}
# RNA
im_size_rna <- (image_rna$Height_cellmask * image_rna$Width_cellmask)/1000000
im_size_rna <- data.frame(im_size_rna)
im_size_rna$Description <- image_rna$Metadata_Description

# Protein
im_size_prot <- (image_prot$Height_cellmask * image_prot$Width_cellmask)/1000000
im_size_prot <- data.frame(im_size_prot)
im_size_prot$Description <- image_prot$Metadata_Description
```

# B cell / patch grouping

```{r}
# patches per image
data.frame(colData(sce_prot)) %>%
  group_by(Description) %>%
  filter(bcell_patch != 0) %>%
  distinct(bcell_patch, .keep_all = T) %>%
  summarise(n=n()) %>%
  arrange(-n)

# max patch size per image
max_patch <- data.frame(colData(sce_prot)) %>%
  filter(bcell_patch != 0) %>%
  group_by(Description, bcell_patch) %>%
  summarise(n=n()) %>%
  summarise(max_patch_size = max(n)) %>%
  arrange(-max_patch_size)

ggplot(max_patch, aes(x=Description, y=max_patch_size)) + 
  geom_col() + 
  geom_hline(yintercept = 250)
```

## Plot some patches

```{r Fig_3B_clusters_and_communities, fig.width=10, fig.height=10, dev=c('pdf')}
example <- findClusters(sce_prot[,sce_prot$Description == "D4"], sce_prot[,sce_prot$celltype %in% c("Bcell", "BnTcell")]$cellID, 
                    'cellID', 
                    'Center_X', 'Center_Y', 
                    'ImageNumber', 
                    distance = 15, 
                    min_clust_size = 20,
                    output_colname = "example_patch")

example <- findCommunity(example, 
              'cellID', 
              'Center_X', 'Center_Y', 
              'ImageNumber', 
              'example_patch', 
              distance = 50,
              output_colname = "example_milieu",
              plot = TRUE)
```

## Make grouping

1 no B cells (B cell density < 10 Bcells/mm2 and no patches)
2 few B cells / no patches (B cell density > 10 Bcells/mm2 and no patches)
3 small patches (max patch size < 250 B cells)
4 large patches (max patch size > 250 B cells)

```{r}
Bcell <- data.frame(colData(sce_prot)) %>%
  group_by(Description,celltype) %>%
  summarise(n=n()) %>%
  reshape2::dcast(Description ~ celltype, value.var = "n", fill = 0) %>%
  select(Description, Bcell)

Bcell$density <- Bcell$Bcell / im_size_prot[match(Bcell$Description, im_size_prot$Description),]$im_size_prot

ggplot(Bcell) + geom_col(aes(x=Description, y=density))

im_w_patches <- data.frame(colData(sce_prot)) %>%
  group_by(Description) %>%
  filter(bcell_patch != 0) %>%
  distinct(bcell_patch, .keep_all = T) %>%
  summarise(n=n()) %>%
  arrange(-n)
```

```{r}
Bcell$Bcell_patch_score <- ""
Bcell[Bcell$density < 10,]$Bcell_patch_score <- "no Bcells"
Bcell[Bcell$density > 10,]$Bcell_patch_score <- "few Bcells"
Bcell[Bcell$Description %in% max_patch[max_patch$max_patch_size<250,]$Description,]$Bcell_patch_score <- "small patches"
Bcell[Bcell$Description %in% max_patch[max_patch$max_patch_size>=250,]$Description,]$Bcell_patch_score <- "large patches"

Bcell %>%
  group_by(Bcell_patch_score) %>%
  summarise(n=n())

Bcell$Bcell_patch_score <- factor(Bcell$Bcell_patch_score, levels = c("no Bcells", "few Bcells", "small patches", "large patches"))

# add to sce
cur_df <- data.frame(colData(sce_prot))
cur_df <- left_join(cur_df, Bcell[,c("Description", "Bcell_patch_score")])
sce_prot$bcell_patch_score <- cur_df$Bcell_patch_score

# add to rna sce
cur_df <- data.frame(colData(sce_rna))
cur_df <- left_join(cur_df, Bcell[,c("Description", "Bcell_patch_score")])
sce_rna$bcell_patch_score <- cur_df$Bcell_patch_score
```

# Analysis

## B cell densities of the different groups

```{r}
ggplot(Bcell,aes(x=Bcell_patch_score, y = log10(density+1))) + 
  geom_boxplot() + 
  geom_quasirandom() + 
  ylab("B cell density (log10)")
```

## Number of Patients in different groups

```{r}
data.frame(colData(sce_rna)) %>%
  filter(Location != "CTRL") %>%
  distinct(PatientID, .keep_all = T) %>%
  group_by(bcell_patch_score) %>%
  summarise(patients = n())
```

## Save SCE object

```{r save sce}
saveRDS(sce_prot, file = "data/sce_protein.rds")
saveRDS(sce_rna, file = "data/sce_RNA.rds")
```

