---
title: "12_Bcell_Score"
author: "toobiwankenobi"
date: "2021-10-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Load packages and helper functions

```{r load packages}
sapply(list.files("code/helper_functions", full.names = TRUE), source)
library(SingleCellExperiment)
library(ggplot2)
library(scater)
library(viridis)
library(igraph)
library(CATALYST)
library(reshape2)
library(cowplot)
library(ggridges)
library(tidyverse)
library(viridis)
library(dplyr)
library(cytomapper)
library(concaveman)
library(data.table)
library(sf)
library(ggbeeswarm)
library(RANN)
```

## Load data

```{r load data}
sce_prot = readRDS(file = "data/data_for_analysis/sce_protein.rds")
sce_rna = readRDS(file = "data/data_for_analysis/sce_RNA.rds")

image_prot <- read.csv("data/data_for_analysis/protein/Image.csv")

sce_prot$bcell_patch_score <- NULL
sce_rna$bcell_patch_score <- NULL
```

# Define grouping for B cell / patch densities

## Cell Densities 

```{r}
# Protein
im_size_prot <- (image_prot$Height_cellmask * image_prot$Width_cellmask)/1000000
im_size_prot <- data.frame(im_size_prot)
im_size_prot$Description <- image_prot$Metadata_Description
```

# B cell / patch grouping

```{r}
# max patch size per image
max_patch <- data.frame(colData(sce_prot)) %>%
  filter(bcell_patch != 0) %>%
  group_by(Description, bcell_patch) %>%
  summarise(n=n()) %>%
  summarise(max_patch_size = max(n)) %>%
  arrange(-max_patch_size)

# assing patch score
max_patch$bcell_patch_score <- ifelse(max_patch$max_patch_size >= median(max_patch$max_patch_size), "B cell Follicles", "Small B cell Patches")
```

## show median for the groups "B cell follicles" and "Small B cell patches"

```{r}
median(max_patch$max_patch_size)
```


## Make grouping

1 No B cells (lower half of median split in images with no B cell patches)
2 No B cell Patches (upper half of median split in images with no B cell patches)
3 Small B cell Patches (lower half of median split for maximum patch size per image)
4 B cell Follicles (upper half of median split for maximum patch size per image)

```{r}
# images with no patches
noPatch_img <- data.frame(colData(sce_prot)) %>%
  group_by(Description) %>%
  summarise(n=sum(bcell_patch)) %>%
  distinct(Description, .keep_all = T) %>%
  ungroup() %>%
  filter(n==0)

# remove all images with patches
Bcell <- data.frame(colData(sce_prot)) %>%
  filter(Description %in% noPatch_img$Description) %>%
  group_by(Description,celltype) %>%
  summarise(n=n()) %>%
  reshape2::dcast(Description ~ celltype, value.var = "n", fill = 0) %>%
  select(Description, `B cell`)

Bcell$density <- Bcell$`B cell` / im_size_prot[match(Bcell$Description, im_size_prot$Description),]$im_size_prot

# assing patch score
Bcell$bcell_patch_score <- ifelse(Bcell$density >= median(Bcell$density), "No B cell Patches", "No B cells")

# merge both data sets
data <- rbind(Bcell[,c("Description", "bcell_patch_score")], max_patch[,c("Description", "bcell_patch_score")])

# factorize with levels
data$bcell_patch_score <- factor(data$bcell_patch_score, levels = c("No B cells", "No B cell Patches", "Small B cell Patches", "B cell Follicles"))

# group sizes
data %>%
  group_by(bcell_patch_score) %>%
  summarise(n=n())
```

## show median for the groups "No B cells" and "No B cell Patches"

```{r}
median(Bcell$density)
```


## add to sce object

```{r}
cur_rna <- data.frame(colData(sce_rna))[,c("Description", "ImageNumber")]
cur_prot <- data.frame(colData(sce_prot))[,c("Description", "ImageNumber")]

cur_rna <- left_join(cur_rna, data)
cur_prot <- left_join(cur_prot, data)

sce_rna$bcell_patch_score <- cur_rna$bcell_patch_score
sce_prot$bcell_patch_score <- cur_prot$bcell_patch_score
```

## Number of Patients in different groups

```{r}
data.frame(colData(sce_rna)) %>%
  filter(Location != "CTRL") %>%
  distinct(PatientID, .keep_all = T) %>%
  group_by(bcell_patch_score) %>%
  summarise(patients = n())
```

## Save SCE object

```{r save sce}
saveRDS(sce_prot, file = "data/data_for_analysis/sce_protein.rds")
saveRDS(sce_rna, file = "data/data_for_analysis/sce_RNA.rds")
```
