---
title: "13_Dysfunction Stain"
author: ""
date: "2021-01-19"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction
Here, we load the data from the dysfunction stain. Images with a T cell dysfunction score were re-acquired with an extended T cell marker panel. An SCE object will be created and saved at the end and serves for further analyses (e.g. Fig 4D, Fig S10B)

# Preparation

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Libraries

```{r}
library(SingleCellExperiment)
library(imcRtools)
library(BiocParallel)
library(BiocNeighbors)
library(dittoSeq)
library(scater)
library(scales)
library(cowplot)
library(Hmisc)
library(tidyverse)
library(ggpubr)
library(scran)
set.seed(12345)
```

## Load Data

```{r}
sce_1 <- read_steinbock(path = "data/full_data/exhaustion_stain/",
                        return_as = "sce", graphs_folder = NULL, regionprops_folder = NULL, 
                        image_file = NULL)

sce_2 <- read_steinbock(path = "data/full_data/revision_stain/",
                        return_as = "sce", graphs_folder = NULL, regionprops_folder = NULL,
                        image_file = NULL)

sce_RNA <- readRDS(file = "data/data_for_analysis/sce_RNA.rds")
```

## add metadata to sce_1

```{r}
assay(sce_1,"asinh") <- asinh(counts(sce_1))
colnames(sce_1) <- paste0(sce_1$sample_id,"_",sce_1$ObjectNumber)

# add information from dysfunction score and Description
imnames <- c("E5","H9","F9","L8","N10","F2","D8","E4","J4","G11","L4","O9")
names(imnames) <- unique(sce_1$sample_id)

sce_1$Description <- imnames[sce_1$sample_id]

# Dysfunction Score
highDys <- unique(sce_RNA[,sce_RNA$dysfunction_score %in% "High Dysfunction"]$Description)
lowDys <- unique(sce_RNA[,sce_RNA$dysfunction_score %in% "Low Dysfunction"]$Description)

sce_1$dysfunction_score <- NA
sce_1[,sce_1$Description %in% highDys]$dysfunction_score <- "High Dysfunction"
sce_1[,sce_1$Description %in% lowDys]$dysfunction_score <- "Low Dysfunction"
```

## add metadata to sce_2

```{r}
assay(sce_2,"asinh") <- asinh(counts(sce_2))
colnames(sce_2) <- paste0(sce_2$sample_id,"_",sce_2$ObjectNumber)

# add information from dysfunction score and Description
imnames <- c("E2","D4","D3","G9","B6","O2","M10","B3","N8","A4","I11","K2","I5","M9")
names(imnames) <- unique(sce_2$sample_id)[c(1,7:14,2:6)]

sce_2$Description <- imnames[sce_2$sample_id]

sce_2$dysfunction_score <- NA
sce_2[,sce_2$Description %in% highDys]$dysfunction_score <- "High Dysfunction"
sce_2[,sce_2$Description %in% lowDys]$dysfunction_score <- "Low Dysfunction"
```

## merge sce objects
we will also remove images E2. this images was the first image of a measurement after downtime of the machine and had substantially higher intensities that all other images.

```{r}
sce <- cbind(sce_1,sce_2)
rm(sce_1,sce_2)

sce <- sce[,sce$Description != "E2"]

# remove images with no dysfunction score
sce <- sce[,sce$dysfunction_score %in% c("High Dysfunction", "Low Dysfunction")]
```

# Analysis

## define cut offs for the definition of CD8 T cells per sample
here we check if there are any global intensity biases for CD8

```{r, fig.height=12, fig.width=8}
dittoRidgePlot(sce,var="CD8a", group.by= "Description",assay = "asinh")
```
__samples E2, F2 L8, D4 and B3 should be inspected further since they have higher CD8 background. However, we will more generally chose a conservative cut off for T cell definition__

```{r}
for (i  in unique(sce$Description)) {
  p <-dittoScatterPlot(sce[,which(sce$Description == i )],x.var = "CD8a",y.var = "CXCL13",assay.x = "asinh",assay.y = "asinh", main=i,)
  plot(p)
}
```

CD8 cut-off at asinh = 2

```{r, fig.height=5, fig.width=5}
for (i  in unique(sce$Description)) {
  p <-dittoScatterPlot(sce[,which(sce$Description == i )],x.var = "CD3",y.var = "CD8a",assay.x = "asinh",assay.y = "asinh", main=i,)
  plot(p)
}
```

general cut off for CD3: 1

## Define Cell Types manually

```{r}
sce$celltype <- "other"
sce$CXCL13 <- "negative"

sce[,which(t(assay(sce,"asinh"))[,"CD8a"] > 2 & t(assay(sce,"asinh"))[,"CD3"] > 1) ]$celltype <- "CD8_Tcell"
sce[,which(t(assay(sce,"asinh"))[,"CXCL13"] > 2 ) ]$CXCL13 <- "positive"

sce[,which(sce$celltype == "CD8_Tcell" & sce$CXCL13 == "positive")]$celltype <- "CD8_CXCL13+_Tcell"
sce[,which(sce$celltype == "other" & sce$CXCL13 == "positive")]$celltype <- "other_CXCL13+"
```

## Save SCE object

```{r save sce}
saveRDS(sce, file = "data/data_for_analysis/sce_dysfunction.rds")
```
