---
title: "Figure 4"
author: "Tobias Hoch and Daniel Schulz"
date: "2020-08-04"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction
This script generates plots for Figure 4.

# Preparations

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message= FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Load libraries

First, we will load the libraries needed for this part of the analysis.

```{r, message=FALSE}
sapply(list.files("code/helper_functions", full.names = TRUE), source)
library(SingleCellExperiment)
library(reshape2)
library(tidyverse)
library(dplyr)
library(data.table) 
library(ggplot2)
library(cba)
library(ComplexHeatmap)
library(colorRamps)
library(circlize)
library(RColorBrewer)
library(ggpubr)
library(ggbeeswarm)
library(gridExtra)
library(tidyr)
library(ggpmisc)
library(circlize)
library(coxme)
library(dittoSeq)
library(scater)
library(cowplot)
library(survminer)
library(ggalluvial)
library(cytomapper)
library(ggrepel)
library(rstatix)
```

## Read the data

```{r}
sce_rna = readRDS(file = "data/data_for_analysis/sce_RNA.rds")
sce_prot = readRDS(file = "data/data_for_analysis/sce_protein.rds")

# meta data
dat_relation = fread(file = "data/data_for_analysis/protein/Object relationships.csv",stringsAsFactors = FALSE)
dat_relation_rna = fread(file = "data/data_for_analysis/RNA/Object relationships.csv",stringsAsFactors = FALSE)

# image
image_mat_prot <- read.csv("data/data_for_analysis/protein/Image.csv")

# surv_dat
dat_survival_prot <- fread(file = "data/data_for_analysis/protein/clinical_data_protein.csv")
```

## Prepare Relation Data Protein

```{r}
# prepare data and add cellID
dat_relation$cellID_first <- paste("protein", paste(dat_relation$`First Image Number`, dat_relation$`First Object Number`, sep = "_"), sep = "_")
dat_relation$cellID_second <- paste("protein", paste(dat_relation$`Second Image Number`, dat_relation$`Second Object Number`, sep = "_"), sep = "_")

# add celltype status to first and second label
celltype_first <- data.frame(colData(sce_prot))[,c("cellID", "celltype", "celltype_clustered")]
colnames(celltype_first) <- c("cellID_first", "celltype_first", "celltype_clust_first")
celltype_second <- data.frame(colData(sce_prot))[,c("cellID", "celltype", "celltype_clustered")]
colnames(celltype_second) <- c("cellID_second", "celltype_second", "celltype_clust_second")

dat_relation <- left_join(dat_relation, celltype_first, by = "cellID_first")
dat_relation <- left_join(dat_relation, celltype_second, by = "cellID_second")

colnames(dat_relation)[5] <- "FirstImageNumber"
```

## Prepare Relation Data RNA

```{r}
# prepare data and add cellID
dat_relation_rna$cellID_first <- paste("RNA", paste(dat_relation_rna$`First Image Number`, dat_relation_rna$`First Object Number`, sep = "_"), sep = "_")
dat_relation_rna$cellID_second <- paste("RNA", paste(dat_relation_rna$`Second Image Number`, dat_relation_rna$`Second Object Number`, sep = "_"), sep = "_")

# add celltype status to first and second label
celltype_first <- data.frame(colData(sce_rna))[,c("cellID", "celltype_rf", "celltype_clustered")]
colnames(celltype_first) <- c("cellID_first", "celltype_first", "celltype_clust_first")
celltype_second <- data.frame(colData(sce_rna))[,c("cellID", "celltype_rf", "celltype_clustered")]
colnames(celltype_second) <- c("cellID_second", "celltype_second", "celltype_clust_second")

dat_relation_rna <- left_join(dat_relation_rna, celltype_first, by = "cellID_first")
dat_relation_rna <- left_join(dat_relation_rna, celltype_second, by = "cellID_second")

colnames(dat_relation_rna)[5] <- "FirstImageNumber"
```

# Figure 4A

## Tumor Marker Profile for different Tcell Scoring Groups

```{r Fig_4A_tumor_profile_tcell, fig.height=6, fig.width=9, dev=c('pdf')}
tumor_marker_protein <- c("pS6", "H3K27me3", "HLADR", "PDL1", "IDO1")
tumor_marker_rna <- c("B2M")

# rna data 
dat_rna <- data.frame(t(assay(sce_rna[tumor_marker_rna, sce_rna$celltype == "Tumor"], "asinh")))
dat_rna$cellID <- rownames(dat_rna)
dat_rna <- left_join(dat_rna, data.frame(colData(sce_rna))[,c("cellID", "Tcell_density_score_image", "Description", "MM_location", "Location")])

# filter
dat_rna <- dat_rna %>%
  filter(Location != "CTRL")

# mean per image
dat_rna <- dat_rna %>%
  select(-cellID) %>%
  group_by(Description, Tcell_density_score_image) %>%
  summarise_if(is.numeric, mean, na.rm = TRUE)

# melt
dat_rna <- dat_rna %>%
  reshape2::melt(id.vars = c("Description", "Tcell_density_score_image"), variable.name = "channel", value.name = "asinh")

# protein data
dat_prot <- data.frame(t(assay(sce_prot[tumor_marker_protein,, sce_prot$celltype == "Tumor"], "asinh")))
dat_prot$cellID <- rownames(dat_prot)
dat_prot <- left_join(dat_prot, data.frame(colData(sce_prot))[,c("cellID", "Tcell_density_score_image", "Description", "MM_location", "Location")])

# filter
dat_prot <- dat_prot %>%
  filter(Location != "CTRL")

# mean per image
dat_prot <- dat_prot %>%
  select(-cellID) %>%
  group_by(Description, Tcell_density_score_image) %>%
  summarise_if(is.numeric, mean, na.rm = TRUE)

# melt
dat_prot <- dat_prot %>%
  reshape2::melt(id.vars = c("Description", "Tcell_density_score_image"), variable.name = "channel", value.name = "asinh")

# join both data sets
comb <- rbind(dat_prot, dat_rna)


# adjusted wilcox.test for groups
group_comparison <- list(c("absent", "high"), c("med", "high"))

stat.test <- comb %>%
  group_by(channel) %>%
  wilcox_test(data = ., asinh ~ Tcell_density_score_image) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj",cutpoints = c(0, 1e-04, 0.001, 0.01, 0.1, 1)) %>%
  add_xy_position(x = "Tcell_density_score_image", dodge = 0.8, comparisons = group_comparison) %>%
  filter(is.na(y.position) == FALSE)

# plot 
p <- ggplot(comb, aes(x=Tcell_density_score_image, y=asinh, 
                      group=Tcell_density_score_image)) + 
  geom_boxplot(alpha=0.2, lwd=1, aes(group=Tcell_density_score_image, fill = Tcell_density_score_image)) +
  facet_wrap(~channel, scales = "free", ncol=3) + 
  stat_pvalue_manual(stat.test, label = "p.adj.signif", size = 7) + 
  geom_quasirandom(alpha=0.6, size=2, aes(group=Tcell_density_score_image, col = Tcell_density_score_image)) +
  scale_color_discrete(guide = FALSE) +
  theme_bw() +
  theme(text = element_text(size=18),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  xlab("") + 
  ylab("Mean Count per Image (asinh)") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  guides(fill=guide_legend(title="T cell Score", override.aes = c(lwd=0.5, alpha=1)))

leg <- get_legend(p)

grid.arrange(p + theme(legend.position = "none"))
```

```{r Fig_4A_tumor_profile_tcell_legend, fig.height=2, fig.width=2, dev=c('pdf')}
grid.arrange(leg)
```

# Figure 4B

## Load masks

```{r}
all_mask <- loadImages(x = "data/full_data/rna/cpout/",
                       pattern = "ilastik_s2_Probabilities_equalized_cellmask.tiff")
```

## add the ImageNumber to masks

```{r match image numbers}
# we load the metadata for the images.
image_mat_rna <- as.data.frame(read.csv(file = "data/data_for_analysis/rna/Image.csv",stringsAsFactors = FALSE))

# we extract only the FileNames of the masks as they are in the all_masks object
cur_df <- data.frame(cellmask = image_mat_rna$FileName_cellmask,
                     ImageNumber = image_mat_rna$ImageNumber,
                     Description = image_mat_rna$Metadata_Description)

# we set the rownames of the extracted data to be equal to the names of all_masks
rownames(cur_df) <- gsub(pattern = ".tiff",replacement = "",image_mat_rna$FileName_cellmask)

# we add the extracted information via mcols in the order of the all_masks object
mcols(all_mask) <- cur_df[names(all_mask),]
```

## scale the masks

```{r scale masks}
all_mask <- scaleImages(all_mask,2^16-1)
```

## Plot two example Images

```{r Fig_4B_example_images_B2Mexpression, dev=("pdf")}
# subset masks
mask_sub <- all_mask[mcols(all_mask)$Description %in% c("L11", "N3")]
sce_rna_sub <- sce_rna[,sce_rna$Description %in% c("L11","N3")]

plotCells(mask = mask_sub, 
          object = sce_rna_sub,
          img_id = "Description", cell_id = "CellNumber",
          colour_by = c("CD3","CD8", "Mart1", "SOX10", "B2M"),
          colour = list(CD3 = c("black", "green"),
                        CD8 = c("black", "green"),
                        Mart1 = c("black", "blue"),
                        SOX10 = c("black", "blue"),
                        B2M = c("black", "red")),
          display = "single",
          exprs_values = "scaled_asinh",
          scale = TRUE)
```

# Figure 4C

## Correlation with mean B2M expression and T cell density

```{r Fig_4C_B2M_correlation, fig.height=4, fig.width=10.7, dev=c('pdf')}
tumor_dat <- data.frame(t(assay(sce_rna["B2M", sce_rna$celltype == "Tumor" & sce_rna$Location != "CTRL"], "asinh")))
tumor_dat$Description <- sce_rna[, sce_rna$celltype == "Tumor" & sce_rna$Location != "CTRL"]$Description

tumor_dat <- tumor_dat %>%
  group_by(Description) %>%
  summarise(mean_B2M = mean(B2M))

cur_df <- data.frame(colData(sce_rna)) %>%
  filter(Location != "CTRL") %>%
  group_by(Description, BlockID, celltype) %>%
  summarise(n=n()) %>%
  mutate(fraction = n/sum(n)) %>%
  ungroup() %>%
  complete(Description, celltype, fill = list(fraction = 0)) %>%
  filter(celltype == "CD8+ T cell")

cur_df_chemokine <- data.frame(colData(sce_rna)) %>%
  filter(Location != "CTRL") %>%
  group_by(Description, chemokine) %>%
  summarise(n=n()) %>%
  reshape2::dcast(Description ~ chemokine, value.var = "n", fill = 0) %>%
  mutate(fraction_positive = `TRUE` / (`FALSE` + `TRUE`))
  
tumor_dat <- left_join(tumor_dat, cur_df)
tumor_dat_chemokine <- left_join(tumor_dat, cur_df_chemokine)

# remove bad images and controls
tumor_dat <- tumor_dat
tumor_dat_chemokine <- tumor_dat_chemokine

# boxplot
a <- ggplot(tumor_dat, aes(y=mean_B2M, x=log10(fraction))) + 
  geom_point(size=3) +
  geom_smooth(method = "lm", formula = y ~ poly(x,2)) +
  stat_regline_equation(formula = y ~ poly(x,2), 
                aes(label =  ..rr.label..), 
                size=10) +
  ylab("Mean B2M Expression (asinh)") +
  xlab("Cytotoxic T cell Fraction (log10)") +
  theme_bw() + 
  theme(text = element_text(size=18))

b <- ggplot(tumor_dat_chemokine, aes(y=mean_B2M, x=log10(fraction_positive))) + 
  geom_point(size=3) +
  geom_smooth(method = "lm", formula = y~poly(x,2)) +
  stat_poly_eq(formula = y ~ poly(x,2), 
                aes(label =  ..rr.label..), 
                parse = TRUE, size=10) +
  ylab("Mean B2M Expression (asinh)") +
  xlab("Chemokine-Expressing Cell Fraction (log10)") +
  theme_bw() + 
  theme(text = element_text(size=18))

grid.arrange(a,b, nrow=1)
```

# Figure 4D

## Interaction between Dysfunctional CD8 and Tumor Cells

```{r Fig_4D_corr_interaction_and_dysfunction, dev=("pdf"), fig.width=8.4, fig.height=3.5}
# number of interactions CD8+ T cell and tumor per image
dat_relation_sub <- dat_relation_rna[dat_relation_rna$celltype_first %in% c("exhausted Tcytotoxic", "Tcytotoxic") & 
                                   dat_relation_rna$celltype_second == "Tumor"]

# count number of interactions
count <- dat_relation_sub %>%
  group_by(FirstImageNumber) %>%
  summarise(n=n())
names(count)[1] <- "ImageNumber"

# fraction of exhausted cd8 per image
dysfunction <- data.frame(colData(sce_rna)) %>%
  mutate(celltype2 = paste(celltype, CXCL13, sep = "_")) %>%
  group_by(ImageNumber, celltype2) %>%
  summarise(n=n()) %>%
  reshape2::dcast(ImageNumber ~ celltype2, value.var = "n", fill = 0) %>%
  reshape2::melt(id.vars = c("ImageNumber"), variable.name = "celltype2", value.name = "n") %>%
  group_by(ImageNumber) %>%
  mutate(fraction = n / sum(n)) %>%
  filter(celltype2 == "CD8+ T cell_1") %>%
  ungroup() %>%
  select(ImageNumber, fraction)

# correlation plot
cur_dat <- left_join(count, dysfunction)

ggplot(cur_dat, aes(x=log10(fraction), y=log10(n))) + 
  geom_point(size=3) + 
  geom_smooth(method="lm") +
  stat_cor(method = "pearson",
           aes(label = paste0("atop(", ..r.label..,  ",", ..rr.label.. ,")")),
           size = 8, cor.coef.name = "R", label.sep="\n", label.y = 1, label.x = -2) + 
  theme_bw() +
  theme(text = element_text(size=14)) +
  xlab("Fraction of Dysfunctional Cytotoxic T Cells (log10)") +
  ylab("Number of CD8/Tumor Interactions\n(log10)")
```

# Figure 4E

## Boxplots with interaction counts per Image 

```{r Fig_4E_interaction_cytotoxic_tumor_boxplots, fig.width=8.8, fig.height=3.7, dev=("pdf")}
# add dysfunction score to dat_relation
ex_score <- data.frame(colData(sce_prot)) %>%
  distinct(ImageNumber, .keep_all = T) %>%
  select(ImageNumber, dysfunction_score, MM_location)

ex_score$FirstImageNumber <- ex_score$ImageNumber

dat_relation <- left_join(dat_relation, ex_score[,c("FirstImageNumber", "dysfunction_score", "MM_location")])

sum <- dat_relation %>%
  filter(celltype_first == "CD8+ T cell" & celltype_second == "Tumor" & !is.na(dysfunction_score)) %>%
  group_by(FirstImageNumber, MM_location, dysfunction_score, celltype_first, celltype_clust_second) %>%
  summarise(n=n()) %>%
  reshape2::dcast(FirstImageNumber + MM_location + dysfunction_score + celltype_first ~ celltype_clust_second, value.var = "n", fill=0) %>%
  reshape2::melt(id.vars = c("FirstImageNumber", "MM_location", "dysfunction_score", "celltype_first"), variable.name = "celltype", value.name = "n")

#sum <- sum[sum$MM_location == "skin_subcutaneous",]

# calculate fractions for every image (makes it more comparable)
sum2 <- sum %>%
  group_by(FirstImageNumber) %>%
  mutate(fraction = n/sum(n)) %>%
  ungroup()

stat.test <- sum2 %>%
  group_by(celltype) %>%
  wilcox_test(data = ., fraction ~ dysfunction_score) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj",cutpoints = c(0, 1e-04, 0.001, 0.01, 0.1, 1)) %>%
  add_x_position(x = "celltype", dodge = 0.8)

ggplot(sum2, aes(x=celltype, y=fraction)) +
  geom_boxplot(alpha=.2, lwd=1, aes(fill = dysfunction_score)) +
  geom_quasirandom(alpha=.6, dodge.width=.75, size=2, aes(group = dysfunction_score, col=dysfunction_score)) +
  stat_pvalue_manual(stat.test, label = "p.adj.signif", size = 7, y.position = 0.9) + 
  scale_color_discrete(guide = FALSE) +
  theme_bw() +
  theme(text = element_text(size = 16),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  guides(fill=guide_legend(title="Dysfunction Score", override.aes = aes(lwd=0.5))) +
  xlab("") + 
  ylab("Fraction of Interactions") +
  ylim(0,1)
```

# Figure 4F

## Tumor Marker Profile for different Scoring Groups per Image

```{r Fig_4F_tumor_profile_per_image, fig.height=4.8, fig.width=7.6, dev=c('pdf')}
tumor_marker_protein <- c("S100", "MiTF")
tumor_marker_rna <- c("Mart1", "pRB")

# rna data 
dat_rna <- data.frame(t(assay(sce_rna[tumor_marker_rna, sce_rna$celltype == "Tumor"], "asinh")))
dat_rna$cellID <- rownames(dat_rna)
dat_rna <- left_join(dat_rna, data.frame(colData(sce_rna))[,c("cellID", "dysfunction_score", "Description", "MM_location")])

# filter
dat_rna <- dat_rna %>%
  filter(dysfunction_score %in% c("high dysfunction", "low dysfunction"))

# mean per image
dat_rna <- dat_rna %>%
  select(-cellID) %>%
  group_by(Description, dysfunction_score) %>%
  summarise_if(is.numeric, mean, na.rm = TRUE)

# melt
dat_rna <- dat_rna %>%
  reshape2::melt(id.vars = c("Description", "dysfunction_score"), variable.name = "channel", value.name = "asinh")

# protein data
dat_prot <- data.frame(t(assay(sce_prot[tumor_marker_protein,, sce_prot$celltype == "Tumor"], "asinh")))
dat_prot$cellID <- rownames(dat_prot)
dat_prot <- left_join(dat_prot, data.frame(colData(sce_prot))[,c("cellID", "dysfunction_score", "Description", "MM_location")])

# filter
dat_prot <- dat_prot %>%
  filter(dysfunction_score %in% c("high dysfunction", "low dysfunction"))

# mean per image
dat_prot <- dat_prot %>%
  select(-cellID) %>%
  group_by(Description, dysfunction_score) %>%
  summarise_if(is.numeric, mean, na.rm = TRUE)

# melt
dat_prot <- dat_prot %>%
  reshape2::melt(id.vars = c("Description", "dysfunction_score"), variable.name = "channel", value.name = "asinh")

# join both data sets
comb <- rbind(dat_prot, dat_rna)

stat.test <- comb %>%
  group_by(channel) %>%
  wilcox_test(data = ., asinh ~ dysfunction_score) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj",cutpoints = c(0, 1e-04, 0.001, 0.01, 0.1, 1)) %>%
  add_xy_position(x = "celltype", dodge = 0.8)

# plot 
ggplot(comb, aes(x=dysfunction_score, y=asinh)) + 
  geom_boxplot(alpha=0.2, lwd=1, aes(fill=dysfunction_score)) +
  geom_quasirandom(alpha=0.6, size=3, aes(col=dysfunction_score)) +
  scale_color_discrete(guide = FALSE) +
  theme_bw() +
  theme(text = element_text(size=16),
        legend.position = "none") +
  facet_wrap(~channel, scales = "free") + 
  stat_pvalue_manual(stat.test, label = "p.adj.signif", size = 7) + 
  ylab("Mean Expression (asinh)") + 
  xlab("") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.2)))
```

# Figure 4G

## S100 and T Cell Dysfunction

```{r Fig_4G_S100_dysfunction, dev=("pdf"), fig.width=7.7, fig.height=3.5}
# fraction of exhausted cd8 per image
dysfunction <- data.frame(colData(sce_rna)) %>%
  mutate(celltype2 = paste(celltype, CXCL13, sep = "_")) %>%
  group_by(ImageNumber, celltype2) %>%
  summarise(n=n()) %>%
  reshape2::dcast(ImageNumber ~ celltype2, value.var = "n", fill = 0) %>%
  reshape2::melt(id.vars = c("ImageNumber"), variable.name = "celltype2", value.name = "n") %>%
  group_by(ImageNumber) %>%
  mutate(fraction = n / sum(n)) %>%
  filter(celltype2 == "CD8+ T cell_1") %>%
  ungroup() %>%
  select(ImageNumber, fraction)

# rna data 
dat_rna <- data.frame(t(assay(sce_rna["S100", sce_rna$celltype == "Tumor"], "asinh")))
dat_rna$cellID <- rownames(dat_rna)
dat_rna <- left_join(dat_rna, data.frame(colData(sce_rna))[,c("cellID", "ImageNumber")])

# mean per image
dat_rna <- dat_rna %>%
  select(-cellID) %>%
  group_by(ImageNumber) %>%
  summarise_if(is.numeric, mean, na.rm = TRUE)

# melt
dat_rna <- dat_rna %>%
  reshape2::melt(id.vars = c("ImageNumber"), variable.name = "channel", value.name = "asinh")

# correlation plot
cur_dat <- left_join(dysfunction, dat_rna)

# high density images
cur_dat <- cur_dat[cur_dat$ImageNumber %in% unique(sce_rna[,colData(sce_rna)$dysfunction_score %in% c("high dysfunction", "low dysfunction")]$ImageNumber),]

ggplot(cur_dat, aes(x=asinh, y=log10(fraction))) + 
  geom_point(size=3) + 
  geom_smooth(method="lm") +
  stat_cor(method = "pearson",
           aes(label = paste(..r.label.., ..rr.label.., sep = "~`,`~")),
           size = 8, cor.coef.name = "R", label.sep="\n", label.y.npc = "top", label.x.npc = "left") + 
  theme_bw() +
  theme(text = element_text(size=15)) +
  xlab("Mean S100 (asinh)") +
  ylab("Fraction of Dysfunctional T cells\n(log10)")
```

# Figure 4H

## Response / Survival Analysis

```{r}
# select blocks in ICI subgroup that are treatment-naive and non-adjuvant
blocks <- unique(sce_prot[,sce_prot$treatment_status_before_surgery == "naive" & 
                            sce_prot$treatment_group_after_surgery == "ICI" &
                            sce_prot$Adjuvant == "n"]$BlockID)

# subset survival data (only data points that contain info for 3m response)
dat_survival_prot_sub <- dat_survival_prot[dat_survival_prot$BlockID %in% blocks & 
                                             is.na(dat_survival_prot$Status_at_3m) == FALSE, ]

# remove LN margin samples
dat_survival_prot_sub <- dat_survival_prot_sub %>%
  mutate(mmLocationPunch = paste(MM_location, Location, sep = "_")) %>%
  filter(mmLocationPunch != "LN_M")
```

## Celltypes Clustered in Responder vs. Non-Responder

```{r, Fig_4H_responder_nonresponder_tumor, fig.height=4, fig.width=8.8, dev=("pdf")}
im_size <- as.data.frame(cbind(image_mat_prot$Metadata_Description, (image_mat_prot$Height_cellmask * image_mat_prot$Width_cellmask)/1000000))
names(im_size) <- c("Description", "mm2")
im_size$mm2 <- as.numeric(im_size$mm2)
im_size[im_size$Description %in% c("G1", "G1 - split"), ]$mm2 <- mean(im_size[im_size$Description %in% c("G1", "G1 - split"), ]$mm2)
im_size <- im_size[im_size != "G1 - split",]

cur_dat <- data.frame(colData(sce_prot[,sce_prot$BlockID %in% unique(dat_survival_prot_sub$BlockID)])) %>%
  mutate(mmLocationPunch = paste(MM_location, Location, sep = "_")) %>%
  #filter(mmLocationPunch != "LN_M") %>% # remove LN margin images 
  #filter(celltype != "Tumor") %>%
  group_by(PatientID,BlockID, Description, bcell_patch_score, celltype_clustered) %>%
  summarise(n=n()) %>% 
  reshape2::dcast(PatientID + BlockID + Description + bcell_patch_score ~ celltype_clustered, value.var = "n", fill=0) %>%
  reshape2::melt(id.vars = c("PatientID","BlockID", "Description", "bcell_patch_score"), 
                 variable.name = "celltype", value.name = "n") %>%
  group_by(Description) %>%
  mutate(fraction = n/sum(n)) %>%
  mutate(total_cells = sum(n)) %>%
  ungroup() %>%
  filter(celltype %in% c("Tumor_1", "Tumor_2", "Tumor_3", "Tumor_4", "Tumor_5", "Tumor_6", "Tumor_7", "Tumor_8", "Tumor_9"))

cur_dat <- left_join(cur_dat, dat_survival_prot_sub[,c("BlockID", "Status_at_3m", "Primary_melanoma_type")]) %>%
  distinct(Description, celltype, .keep_all = T)

cur_dat <- left_join(cur_dat, im_size[im_size$Description %in% unique(cur_dat$Description),])
cur_dat$n_per_mm2 <- cur_dat$n / cur_dat$mm2

stat.test <- cur_dat %>%
  group_by(celltype) %>%
  mutate(y_log10 = log10(n_per_mm2+1)) %>%
  wilcox_test(data = ., y_log10 ~ Status_at_3m) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj",cutpoints = c(0, 1e-04, 0.001, 0.01, 0.1, 1)) %>%
  add_x_position(x = "celltype", dodge = 0.8)

#stat.test$p.adj.format <- ifelse(stat.test$p.adj < 0.05, paste("p = ", round(stat.test$p.adj,2), sep = ""), "n.s.")

ggplot(cur_dat, aes(x=celltype, y=log10(n_per_mm2+1))) + 
  geom_boxplot(alpha=.5, outlier.shape = NA, aes(fill=Status_at_3m)) +
  geom_quasirandom(size=1, dodge.width = 0.75, aes(group = Status_at_3m, col=Status_at_3m)) +
  stat_pvalue_manual(stat.test, x = "celltype", label = "p.adj.signif", size = 7, y.position = 4) + 
  theme_bw() +
  theme(text = element_text(size = 16),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  ylab("Cells per mm2 (log10)") +
  xlab("") +
  scale_color_manual(guide = FALSE, values=c("blue", "orange")) +
  scale_fill_manual(values=c("blue", "orange")) + 
  guides(fill=guide_legend(title="Response 3m", override.aes = c(lwd=0.5, size=1))) +
  scale_y_continuous(expand = c(0.1, 0))

# number of patients
dat_survival_prot_sub %>%
  distinct(PatientID, .keep_all = T) %>%
  group_by(Status_at_3m) %>%
  summarise(n=n())

# number of images
dat_survival_prot_sub %>%
  distinct(Description, .keep_all = T) %>%
  group_by(Status_at_3m) %>%
  summarise(n=n())
```
