---
title: "Figure 1"
author: "Tobias Hoch and Daniel Schulz"
date: "2020-08-04"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction
This script generates plots for Figure 1. 

# Preparation

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message= FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Load libraries

```{r load_libraries, message=FALSE, warning=FALSE}
library(SingleCellExperiment)
library(data.table)
library(scater)
library(ggplot2)
library(dittoSeq)
library(cba)
library(ComplexHeatmap)
library(corrplot)
library(dplyr)
library(reshape2)
library(tidyverse)
library(rms)
library(cytomapper)
library(ggrepel)
library(circlize)
library(ggbeeswarm)
library(dendextend)
```

## Load data

```{r load_data}
sce_rna <- readRDS(file = "data/data_for_analysis/sce_RNA.rds")
sce_prot <- readRDS(file = "data/data_for_analysis/sce_protein.rds")
```

# Figure 1C

## Heatmap showing marker expression of celltypes

```{r Fig_1C_Scaled_Heatmap_RNA, fig.width=10, fig.height=11.2, dev=c('pdf')}
# add marker expression to cells
marker_expression <- data.frame(t(assay(sce_rna[rowData(sce_rna)$good_marker,], "asinh")))
marker_expression$cellID <- rownames(marker_expression)
dat <- data.frame(colData(sce_rna))[,c("cellID", "celltype")]
dat <- left_join(dat, marker_expression)
dat$cellID <- NULL

# aggregate the groups
dat_aggr <- dat %>%
  group_by(celltype) %>%
  summarise_all(funs(mean))

# number of cells per group
dat_sum <- dat %>%
  group_by(celltype) %>%
  summarise(n=n())

dat_sum <- data.frame(t(dat_sum))

# scale and center expression
dat_aggr[,-c(1)] <- scale(dat_aggr[,-c(1)])

# create matrix
m <- as.matrix(t(dat_aggr[,-c(1)]))
colnames(m) <- dat_aggr$celltype

# top annotation with number of cells
# ha <- HeatmapAnnotation("Number of Cells" = anno_barplot(ifelse(as.numeric(dat_sum[2,])>100000, 101000, as.numeric(dat_sum[2,])),
#                                                          height = unit(3,"cm"),
#                                                          ylim = range(0,100000),
#                                                          gp = gpar(fill = ifelse(as.numeric(dat_sum[2,]) > 100000, "white", "black"),
#                                                                    col="white"),
#                                                          axis_param = list(gp = gpar(fontsize=14))),
#                         "Numbers" = anno_text(round(as.numeric(dat_sum[2,])), 
#                                               which = "column", 
#                                               rot = 0, 
#                                               height = unit(0.5,"cm"),
#                                               just = "center", 
#                                               location = 0.5,
#                                               gp = gpar(fontsize=10,col = ifelse(as.numeric(dat_sum[2,]) > 100000, "red", "black"))),
#                         annotation_name_gp = gpar(fontsize = 16))

ha <- HeatmapAnnotation("Numbers" = anno_text(format(round(as.numeric(dat_sum[2,])), nsmall = 0, big.mark = "'"), 
                                                   which = "column", 
                                                   rot = 0, 
                                                   just = "center",
                                              height = unit(1,"cm"),
                                                   location = 0.5,
                                                   gp = gpar(fontsize=12,col = "black", border = "black")))

# row_split for markers
rowData(sce_rna)$heatmap_relevance <- ""
rowData(sce_rna[rowData(sce_rna)$good_marker,])$heatmap_relevance <- "Lineage"
rowData(sce_rna[grepl("CXCL|CCL|DapB", rownames(sce_rna)),])$heatmap_relevance <- "Chemokine"
rowData(sce_rna[grepl("B2M|GLUT1|CD134|Lag3|CD163|cleavedPARP|pRB", rownames(sce_rna)),])$heatmap_relevance <- "Other"

# plot heatmap
h <- Heatmap(m, name = "Scaled Expression",
             row_split = rowData(sce_rna[rowData(sce_rna)$good_marker,])$heatmap_relevance,
             cluster_columns = TRUE,
             show_column_dend = FALSE,
             column_names_gp = gpar(fontsize=18),
             column_names_rot = 45,
             column_names_centered = FALSE,
             show_column_names = TRUE,
             top_annotation = ha,
             row_names_gp = gpar(fontsize = 16),
             row_title_gp = gpar(fontsize = 23),
             col = colorRamp2(c(-2, 0, 2), c("blue", "white", "red")),
             heatmap_legend_param = list(at = c(-2:2), legend_width = unit(6,"cm"), 
                                         direction="horizontal", title_gp = gpar(fontsize=16),
                                         labels_gp = gpar(fontsize=12), title_position = "topcenter"),
             column_names_side = "top",
             height = unit(20, "cm"),
             width = unit(17,"cm"))

draw(h, heatmap_legend_side = "bottom")
```

## Heatmap showing marker expression of celltypes

```{r Fig_1C_Scaled_Heatmap_Protein, fig.width=10, fig.height=11.4, dev=c('pdf')}
# add marker expression to cells
marker_expression <- data.frame(t(assay(sce_prot[rowData(sce_prot)$good_marker,], "asinh")))
marker_expression$cellID <- rownames(marker_expression)
dat <- data.frame(colData(sce_prot))[,c("cellID", "celltype")]
dat <- left_join(dat, marker_expression)
dat$cellID <- NULL

# aggregate the groups
dat_aggr <- dat %>%
  group_by(celltype) %>%
  summarise_all(funs(mean))

# number of cells per group
dat_sum <- dat %>%
  group_by(celltype) %>%
  summarise(n=n())

dat_sum <- data.frame(t(dat_sum))

# scale and center expression
dat_aggr[,-c(1)] <- scale(dat_aggr[,-c(1)])

# create matrix
m <- as.matrix(t(dat_aggr[,-c(1)]))
colnames(m) <- dat_aggr$celltype

# top annotation with number of cells
# ha <- HeatmapAnnotation("Number of Cells" = anno_barplot(ifelse(as.numeric(dat_sum[2,])>70000, 70000, as.numeric(dat_sum[2,])),
#                                                          height = unit(3,"cm"),
#                                                          ylim = range(0,70000),
#                                                          gp = gpar(fill = ifelse(as.numeric(dat_sum[2,]) > 70000, "white", "black"),
#                                                                    col = "white"),
#                                                          axis_param = list(gp = gpar(fontsize=14))),
#                         "Numbers" = anno_text(round(as.numeric(dat_sum[2,])), 
#                                               which = "column", 
#                                               rot = 0, 
#                                               just = "center",
#                                               height = unit(0.5,"cm"),
#                                               location = 0.5,
#                                               gp = gpar(fontsize=10,col = ifelse(as.numeric(dat_sum[2,]) > 70000, "red", "black"))),
#                         annotation_name_gp = gpar(fontsize = 16))

ha <- HeatmapAnnotation("Numbers" = anno_text(format(round(as.numeric(dat_sum[2,])), nsmall = 0, big.mark = "'"), 
                                                   which = "column", 
                                                   rot = 0, 
                                                   just = "center",
                                              height = unit(1,"cm"),
                                                   location = 0.5,
                                                   gp = gpar(fontsize=12,col = "black", border = "black")))

# row_split for markers
rowData(sce_prot)$heatmap_relevance <- ""
rowData(sce_prot[rowData(sce_prot)$good_marker,])$heatmap_relevance <- "Lineage"
rowData(sce_prot[grepl("PDL1|CD11b|CD206|PARP|CXCR2|CD11c|pS6|GrzB|IDO1|CD45RA|H3K27me3|TCF7|CD45RO|PD1|pERK|ICOS|Ki67", rownames(sce_prot)),])$heatmap_relevance <- "Other"

# plot heatmap
h <- Heatmap(m, name = "Scaled Expression",
             row_split = rowData(sce_prot[rowData(sce_prot)$good_marker,])$heatmap_relevance,
             cluster_columns = TRUE,
             show_column_dend = FALSE,
             column_names_gp = gpar(fontsize=18),
             column_names_rot = 45,
             column_names_centered = FALSE,
             show_column_names = TRUE,
             row_names_gp = gpar(fontsize = 16),
             row_title_gp = gpar(fontsize = 23),
             top_annotation = ha,
             col = colorRamp2(c(-2, 0, 2), c("blue", "white", "red")),
             heatmap_legend_param = list(at = c(-2:2), legend_width = unit(6,"cm"), 
                                         direction="horizontal", title_gp = gpar(fontsize=16),title_position = "topcenter",
                                         labels_gp = gpar(fontsize=12)),
             column_names_side = "top",
             height = unit(20, "cm"),
             width = unit(17,"cm"))

draw(h, heatmap_legend_side = "bottom")
```

# Figure 1D

## Cell Type Correlation between the two data sets

```{r Fig_1D_Celltype_Correlation, dev=c('pdf'), fig.height=8, fig.width=9}
cur_rna <- data.frame(colData(sce_rna))
cur_prot <- data.frame(colData(sce_prot))

rna_sum <- cur_rna %>%
  group_by(Description, celltype) %>%
  summarise(n = n()) %>%
  mutate(fraction = n/sum(n)) %>%
  reshape2::dcast(Description ~ celltype, value.var = "fraction", fill = 0)

prot_sum <- cur_prot %>%
  group_by(Description, celltype) %>%
  summarise(n = n()) %>%
  mutate(fraction = n/sum(n)) %>%
  reshape2::dcast(Description ~ celltype, value.var = "fraction", fill = 0)

# Correlation Plot - shared celltypes
shared <- c("CD8+ T cell", "Macrophage", "Neutrophil", "Tumor")
order_prot <- c("B cell", "BnT cell", "pDC", "Stroma", "unknown", "FOXP3+ T cell", "CD4+ T cell", shared)
order_rna <- c("CD38", "HLA-DR", "Stroma", "Vasculature", "unknown", "CD8- T cell", shared)

corrplot(cor(rna_sum[,order_rna], prot_sum[,order_prot], method = "pearson"), 
         addCoef.col = "white",
         method = "circle",
         tl.col="black",
         tl.cex = 1.5)

# Mean Correlation between shared celltypes
mean(c(cor(rna_sum$Macrophage, prot_sum$Macrophage, method = "pearson"), 
       cor(rna_sum$Neutrophil, prot_sum$Neutrophil, method = "pearson"), 
       cor(rna_sum$`CD8+ T cell`, prot_sum$`CD8+ T cell`, method = "pearson"), 
       cor(rna_sum$Tumor, prot_sum$Tumor, method = "pearson")))
```

# Figure 1E

## Load masks

```{r}
all_mask <- loadImages(x = "data/full_data/protein/cpout/",
                       pattern = "ilastik_s2_Probabilities_equalized_cellmask.tiff")
```

## add the ImageNumber to masks

```{r match image numbers}
# we load the metadata for the images.
image_mat <- as.data.frame(read.csv(file = "data/data_for_analysis/protein/Image.csv",stringsAsFactors = FALSE))

# we extract only the FileNames of the masks as they are in the all_masks object
cur_df <- data.frame(cellmask = image_mat$FileName_cellmask,
                     ImageNumber = image_mat$ImageNumber,
                     Description = image_mat$Metadata_Description)

# we set the rownames of the extracted data to be equal to the names of all_masks
rownames(cur_df) <- gsub(pattern = ".tiff",replacement = "",image_mat$FileName_cellmask)

# we add the extracted information via mcols in the order of the all_masks object
mcols(all_mask) <- cur_df[names(all_mask),]
```

## scale the masks

```{r scale masks}
all_mask <- scaleImages(all_mask,2^16-1)
```

## Plot two example Images

```{r Fig_1E_example_images_immune_cells, dev=("pdf")}
# subset masks
mask_sub <- all_mask[mcols(all_mask)$Description %in% c("H9", "L7", "H2")]
sce_prot_sub <- sce_prot[,sce_prot$Description %in% c("H9", "L7", "H2")]

# rename all cells that are not tumor and not tcytotoxic
#sce_prot_sub$celltype <- ifelse(sce_prot_sub$celltype %in% c("Tumor", "CD8+ T cell"), sce_prot_sub$celltype, "other")

col_list <- list()
col_list$`Cell Type` <- metadata(sce_prot)$colour_vectors$celltype

sce_prot_sub$`Cell Type` <- sce_prot_sub$celltype

plotCells(mask = mask_sub, 
          object = sce_prot_sub,
          cell_id = "CellNumber", img_id = "Description", 
          colour_by = "Cell Type",
          colour = col_list,
          display = "single")
```

# Figure 1F

## Barplot containing Cell Types per Image

```{r Fig_1F_Protein_Barplot, fig.width=13, fig.height=8, dev=c('pdf')}
# Create table with celltype fractions
cur_df <- data.frame(celltype = sce_prot$celltype,
                     Description = sce_prot$Description,
                     Location = sce_prot$Location)

# remove control samples 
cur_df <- cur_df %>%
  filter(Location != "CTRL") %>%
  group_by(Description, celltype) %>%
  summarise(n=n()) %>% 
  group_by(Description) %>%
  mutate(fraction = n / sum(n)) %>%
  reshape2::dcast(Description ~ celltype, value.var = "fraction", fill=0)

matrixrownames <- cur_df$Description 

# now we create a matrix from the data and cluster the data based on the cell fractions
hm_dat = as.matrix(cur_df[,-1])
rownames(hm_dat) <- as.character(matrixrownames)

# calculate distance and then cluster images based on cluster fraction
dd <- dist((hm_dat), method = "euclidean")
hc <- hclust(dd, method = "ward.D2")
row_sorted <- hc$labels

# now we generate the clinical metadata and order it in the same way as the celltype data
patient_meta <- metadata(sce_prot)
patient_meta <- data.frame(patient_meta[c("Description", "MM_location_simplified", "Location")])
patient_meta <- patient_meta %>%
  filter(Location != "CTRL")

mrownames <- patient_meta$Description
patient_meta <- as.matrix(patient_meta)
rownames(patient_meta) <- mrownames
patient_meta <- data.frame(patient_meta[row_sorted,])

# generate the barplot. this is generated as the annotation for the heatmap of the Patient_ID that is generated below.
hm_dat <- hm_dat[row_sorted,]

# bring cell types in order (column order)
col_order <- c("Tumor", "Macrophage", "Neutrophil", "CD8+ T cell", "CD4+ T cell", "FOXP3+ T cell", "B cell", "BnT cell", "pDC", "Stroma", "unknown")
hm_dat <- hm_dat[, col_order]

col_vector <- metadata(sce_prot)$colour_vector$celltype[colnames(hm_dat)]
col_vector_location <- structure(c("blue", "red", "green"), 
                                 names = c("LN", "other", "skin"))

# rename punch locations
patient_meta[patient_meta$Location == "M", ]$Location <- "margin"
patient_meta[patient_meta$Location == "C", ]$Location <- "core"

col_vector_punch <- structure(c("black", "grey"), 
                                 names = c("core", "margin"))

# create annotation with cell type proportions
ha <- rowAnnotation(`Punch Location` = patient_meta[,c("Location")],
                    `Cell Type Proportion` =
                      anno_barplot(hm_dat, 
                                   gp=gpar(fill=col_vector),
                                   bar_width = 1,
                                   height = unit(30,"cm"),
                                   width = unit(15,"cm"),
                                   show_row_names = FALSE),
                        col = list(`Punch Location` = col_vector_punch),
                    show_legend = FALSE)


dend <- as.dendrogram(hc)
dend <- color_branches(dend, k = 4, col = c("gray50", "blue", "green", "red"), groupLabels = TRUE) # `color_branches()` returns a dendrogram object

# heatmap consisting of the patient_IDs. one color per patient
h1 = Heatmap(patient_meta[,c("MM_location_simplified")], 
             col = col_vector_location, 
             width = unit(0.5, "cm"), 
             cluster_rows = dend,
             row_dend_width = unit(3, "cm"),
             height = unit(30, "cm"),
             show_heatmap_legend = FALSE, 
             heatmap_legend_param = list(title = "Met Location"),
             row_names_gp = gpar(cex=0.5),
             show_row_names = TRUE,
             right_annotation =  ha,
             column_labels = "Met Location")

# plot the data
ht = grid.grabExpr(draw(h1))
grid.newpage()
pushViewport(viewport(angle = 270))
grid.draw(ht)
```

## Legend for Barplot

```{r Fig_1F_Protein_Barplot_Legend, fig.width=2, fig.height=4, dev=c('pdf')}
lgd1 <- Legend(labels = names(col_vector), 
             title = "Cell Type", 
             legend_gp = gpar(fill = col_vector),
             ncol = 1)

lgd2 <- Legend(labels = names(col_vector_location), 
             legend_gp = gpar(fill = unname(col_vector_location)),
             title = "Met Location",
             ncol = 1)

lgd3 <- Legend(labels = names(col_vector_punch), 
             legend_gp = gpar(fill = unname(col_vector_punch)),
             title = "Punch Location",
             ncol = 1)

pd = packLegend(lgd1, lgd2, lgd3, direction = "vertical", 
    column_gap = unit(0.5, "cm"))
draw(pd)
```
