---
title: "Figure 5"
author: "Tobias Hoch and Daniel Schulz"
date: "2021-01-30"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction
This script generates plots for Figure 5.

# Preparations

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message= FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Load libraries

First, we will load the libraries needed for this part of the analysis.

```{r, message=FALSE}
sapply(list.files("code/helper_functions", full.names = TRUE), source)
library(SingleCellExperiment)
library(reshape2)
library(tidyverse)
library(dplyr)
library(data.table) 
library(ggplot2)
library(cba)
library(ComplexHeatmap)
library(colorRamps)
library(circlize)
library(RColorBrewer)
library(ggpubr)
library(ggbeeswarm)
library(gridExtra)
library(tidyr)
library(ggpmisc)
library(circlize)
library(coxme)
library(dittoSeq)
library(scater)
library(cowplot)
library(survminer)
library(ggalluvial)
library(cytomapper)
library(corrplot)
library(ggridges)
library(rstatix)
library(sf)
library(concaveman)
library(RANN)
```

## Read the data

```{r}
sce_rna = readRDS(file = "data/data_for_analysis/sce_RNA.rds")
sce_prot = readRDS(file = "data/data_for_analysis/sce_protein.rds")

# meta data
dat_relation = fread(file = "data/data_for_analysis/protein/Object relationships.csv",stringsAsFactors = FALSE)
dat_relation_rna = fread(file = "data/data_for_analysis/RNA/Object relationships.csv",stringsAsFactors = FALSE)

# image
image_mat_rna <- read.csv("data/data_for_analysis/rna/Image.csv")

# surv_dat
dat_survival_prot <- fread(file = "data/data_for_analysis/protein/clinical_data_protein.csv")
```

## Prepare Relation Data Protein

```{r}
# prepare data and add cellID
dat_relation$cellID_first <- paste("protein", paste(dat_relation$`First Image Number`, dat_relation$`First Object Number`, sep = "_"), sep = "_")
dat_relation$cellID_second <- paste("protein", paste(dat_relation$`Second Image Number`, dat_relation$`Second Object Number`, sep = "_"), sep = "_")

# add celltype status to first and second label
celltype_first <- data.frame(colData(sce_prot))[,c("cellID", "celltype", "celltype_clustered")]
colnames(celltype_first) <- c("cellID_first", "celltype_first", "celltype_clust_first")
celltype_second <- data.frame(colData(sce_prot))[,c("cellID", "celltype", "celltype_clustered")]
colnames(celltype_second) <- c("cellID_second", "celltype_second", "celltype_clust_second")

dat_relation <- left_join(dat_relation, celltype_first, by = "cellID_first")
dat_relation <- left_join(dat_relation, celltype_second, by = "cellID_second")

colnames(dat_relation)[5] <- "FirstImageNumber"
```

## Prepare Relation Data RNA

```{r}
# prepare data and add cellID
dat_relation_rna$cellID_first <- paste("RNA", paste(dat_relation_rna$`First Image Number`, dat_relation_rna$`First Object Number`, sep = "_"), sep = "_")
dat_relation_rna$cellID_second <- paste("RNA", paste(dat_relation_rna$`Second Image Number`, dat_relation_rna$`Second Object Number`, sep = "_"), sep = "_")

# add celltype status to first and second label
celltype_first <- data.frame(colData(sce_rna))[,c("cellID", "celltype_rf", "celltype_clustered")]
colnames(celltype_first) <- c("cellID_first", "celltype_first", "celltype_clust_first")
celltype_second <- data.frame(colData(sce_rna))[,c("cellID", "celltype_rf", "celltype_clustered")]
colnames(celltype_second) <- c("cellID_second", "celltype_second", "celltype_clust_second")

dat_relation_rna <- left_join(dat_relation_rna, celltype_first, by = "cellID_first")
dat_relation_rna <- left_join(dat_relation_rna, celltype_second, by = "cellID_second")

colnames(dat_relation_rna)[5] <- "FirstImageNumber"
```

# Figure 5A

## Generate adjacency matrix for all images

```{r Fig_5A_interaction_chord_groups, fig.height=7, fig.width=5, dev=c('pdf')}
# subset sce for inflamed/exhausted in high samples
sce_protein_sub <- sce_prot[, sce_prot$dysfunction_density %in% c("high - high dysfunction", "high - low dysfunction")]

# sample 9 images each
images <- data.frame(colData(sce_protein_sub)) %>%
  distinct(ImageNumber, .keep_all = T) %>%
  group_by(dysfunction_density) %>%
  #filter(ImageNumber %in% sample(ImageNumber, 9)) %>%
  select(ImageNumber, dysfunction_density)

return <- list()

for (i in c("high - high dysfunction", "high - low dysfunction")){
  # title name
  title_name <- i

  # count interactions in 20 sample images
  cur_dat_relation <- data.frame(dat_relation) %>%
    filter(FirstImageNumber %in% images[images$dysfunction_density == i,]$ImageNumber) %>%
    select("celltype_first" ,"celltype_second") %>%
    dplyr::count(celltype_first,celltype_second) %>%
    data.frame()
  
  # remove tumor-tumor interactions
  cur_dat_relation <- cur_dat_relation[cur_dat_relation$celltype_first != "Tumor",]
  cur_dat_relation <- cur_dat_relation[cur_dat_relation$celltype_first != "unknown",]
  cur_dat_relation <- cur_dat_relation[cur_dat_relation$celltype_second != "unknown",]
  
    # count interactions
  cur_dat_relation_subcluster <- data.frame(dat_relation) %>%
    filter(FirstImageNumber %in% images[images$dysfunction_density == i,]$ImageNumber) %>%
    select("celltype_first" , "celltype_clust_second") %>%
    dplyr::count(celltype_first, celltype_clust_second) %>%
    data.frame()
  
  # remove tumor-tumor interactions
  cur_dat_relation_subcluster <- cur_dat_relation_subcluster[cur_dat_relation_subcluster$celltype_first == "Tcytotoxic",]
  cur_dat_relation_subcluster <- cur_dat_relation_subcluster[cur_dat_relation_subcluster$celltype_clust_second %in% 
                                                               unique(sce_prot[,sce_prot$celltype == "Tumor"]$celltype_clustered),]
  # return subcluster adj df
  return[[i]] <- cur_dat_relation_subcluster
  
  # make coord diagram
  chordDiagramFromDataFrame(cur_dat_relation,
                            grid.col = metadata(sce_prot)$colour_vectors$celltype,
                            reduce = 0.05, 
                            transparency = ifelse(cur_dat_relation[[1]] %in% c("Tcytotoxic"),0,0.6),
                            annotationTrack = c("grid"))
  
  # add percentages
  circos.track(track.index = 1, panel.fun = function(x, y) {
    xlim = get.cell.meta.data("xlim")
    ylim = get.cell.meta.data("ylim")
    sector.name = get.cell.meta.data("sector.index")
    xplot = get.cell.meta.data("xplot")
    
    circos.lines(xlim, c(mean(ylim), mean(ylim)), lty = 3) # dotted line
    by = ifelse(abs(xplot[2] - xplot[1]) > 30, 0.2, 0.5)
    for(p in seq(by, 1, by = by)) {
        circos.text(p*(xlim[2] - xlim[1]) + xlim[1], mean(ylim) + 0.1, 
            paste0(p*100, "%"), cex = 0.5, adj = c(0.5, 0), niceFacing = TRUE)
    }
}, bg.border = NA)
  
}
```

```{r Fig_5A_legend, dev=c('pdf'), fig.height=5, fig.width=1}
# create legend for tumor subclusters
#lgd1 = Legend(labels = names(col_vec[grep("Tumor", names(col_vec))]), title = "Tumor\nSubclusters", legend_gp = gpar(fill = unname(col_vec[grep("Tumor", names(col_vec))])))
lgd2 = Legend(labels = names(metadata(sce_prot)$colour_vector$celltype), title = "Cell Type", legend_gp = gpar(fill = unname(metadata(sce_prot)$colour_vector$celltype)))
draw(packLegend(lgd2, 
                #lgd1, 
                gap = unit(2, "cm")))
```

# Figure 5B

## Celltypes

```{r Fig_5B_celltype_fractions_dysfunction, dev=("pdf"), fig.width=10, fig.height=5}
celltypes <- data.frame(colData(sce_prot)) %>%
  filter(celltype != "Tumor") %>%
  group_by(ImageNumber, bcell_patch_score, dysfunction_score, celltype) %>%
  summarise(n=n()) %>%
  mutate(fraction = n/sum(n)) %>%
  filter(is.na(dysfunction_score) == FALSE)

celltypes$bcell_patch_score <- as.character(celltypes$bcell_patch_score)
celltypes$bcell_patch_score <- factor(celltypes$bcell_patch_score, levels = c("no B cells", "no B cell patches", "small B cell patches", "TLS"))

stat.test <- celltypes %>%
  group_by(celltype) %>%
  wilcox_test(data = ., fraction ~ dysfunction_score) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj") %>%
  add_x_position(x = "dysfunction_score", dodge = 0.8)


ggplot(celltypes, aes(x=celltype, y=fraction)) +
  geom_boxplot(alpha=.5, lwd = 1, outlier.shape = NA, aes(fill=dysfunction_score)) +
  #geom_jitter(alpha=1, position = position_jitterdodge(dodge.width = 0.75,jitter.width = 0.1), size = 4, 
              #aes(x = celltype, y = fraction, #col=ifelse(celltype %in% c("Bcell", "BnTcell"), bcell_patch_score, " "),
                  #group = dysfunction_score)) + 
  geom_quasirandom(dodge.width=0.75, alpha=1, size=1) + 
  stat_pvalue_manual(stat.test, x = "celltype", label = "p.adj.signif", size = 7, y.position = -0.05) + 
  theme_bw() +
  theme(text = element_text(size = 16),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) + 
  guides(fill=guide_legend(title="Dysfunction Score", override.aes = c(lwd=0.5, alpha=1)), col=guide_legend(title="Bcell Score")) +
  xlab("") + 
  ylab("Cell Type Fractions") +
  scale_color_manual(values = c("black", "lightblue", "darkblue", "red", "red4"),
                    labels = c(" ", "no Bcells", "few Bcells", "small patches", "large patches"),
                    guide = TRUE)

# check images above median in low dysfunction group for B cells (all have large patches)
celltypes %>%
  filter(dysfunction_score == "low dysfunction") %>%
  filter(celltype == "Bcell") %>%
  group_by(dysfunction_score) %>%
  mutate(median_frac = median(fraction)) %>%
  filter(fraction > median_frac)
```

## Composition of Dysfunction Groups (Met Location and Punch Location)

```{r alluvial_dysfunction_score, fig.width=10, fig.height=5, dev=("pdf"), eval=F}
c <- data.frame(colData(sce_rna)) %>%
  distinct(Description, .keep_all = T) %>%
  filter(dysfunction_score %in% c("low dysfunction", "high dysfunction")) %>%
  dplyr::count(MM_location, Location, dysfunction_score) %>%
  ggplot(aes(axis1 = MM_location, axis2 = Location, axis3 = dysfunction_score,y=n))+
  geom_alluvium(aes(fill=as.factor(dysfunction_score)))+
  geom_stratum()+
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  theme_minimal()+
  scale_x_discrete(limits = c("Met Location", "Punch Location", "Dysfunction Score"), expand = c(.2, .05)) +
  theme(text = element_text(size=20)) +
  guides(fill=guide_legend(title="Dysfunction Score")) +
  ylab("Number of Samples")
plot(c)
```

# Figure 5C

## B cell patch grouping for dysfunction groups

```{r Fig_5C_bcell_groups_fraction, dev=("pdf"), fig.width=10, fig.height=3}
a <- data.frame(colData(sce_rna)) %>%
  distinct(Description, .keep_all = T) %>%
  group_by(bcell_patch_score, dysfunction_score) %>%
  summarise(n=n()) %>%
  filter(is.na(dysfunction_score) == F) %>%
  group_by(dysfunction_score) %>%
  mutate(fraction = n / sum(n)) %>%
  ungroup()

ggplot(a) + 
  geom_col(aes(y=dysfunction_score, x=fraction, fill=bcell_patch_score)) +
  theme_minimal() + 
  theme(text = element_text(size = 22)) +
  xlab("Fraction of Samples") +
  ylab("") +
  guides(fill=guide_legend(title="Bcell Score"))
```

# Figure 5D

## Analysis

```{r Fig_5D_Bcell_patches_CXCL13, fig.height=6, fig.width=9, dev=c('pdf')}
# im_size
im_size <- as.data.frame(cbind(image_mat_rna$Metadata_Description, (image_mat_rna$Height_cellmask * image_mat_rna$Width_cellmask)/1000000))
names(im_size) <- c("Description", "mm2")
im_size$mm2 <- as.numeric(im_size$mm2)

cxcl13 <- data.frame(colData(sce_rna)) %>%
  mutate(mmLocationPunch = paste(MM_location, Location, sep = "_")) %>%
  filter(mmLocationPunch != "LN_M") %>% # remove LN margin samples 
  filter(CXCL13 == 1) %>%
  group_by(Description, MM_location, celltype) %>%
  summarise(n=n()) %>%
  reshape2::dcast(Description + MM_location ~ celltype, value.var = "n", fill = 0)

all_images <- data.frame(colData(sce_rna)) %>%
  mutate(mmLocationPunch = paste(MM_location, Location, sep = "_")) %>%
  filter(mmLocationPunch != "LN_M") %>%
  distinct(Description, .keep_all = T) %>%
  select(Description)

# left_join to have all images
all_images <- left_join(all_images, cxcl13)

# replace NA
all_images[is.na(all_images)] <- 0

all_images <- all_images %>%
  reshape2::melt(id.vars=c("Description", "MM_location"), variable.name = "celltype", value.name = "n")

# add mm2 and calculate density
all_images <- left_join(all_images, im_size)
all_images$density <- all_images$n / all_images$mm2

#
Bcell <- data.frame(colData(sce_rna)) %>%
  distinct(Description, .keep_all = T) %>%
  group_by(Description) %>%
  select(Description, bcell_patch_score)

# add grouping
all_images <- left_join(all_images, Bcell[,c("Description","bcell_patch_score")]) %>%
  filter(celltype %in% c("Tcytotoxic", "Tcell", "HLA-DR")) 

#all_images$interaction <- interaction(all_images$celltype, all_images$bcell_patch_score)

ggplot(all_images,aes(x=bcell_patch_score, y = log10(as.numeric(density+1)), fill=celltype)) + 
  geom_boxplot(alpha=1, lwd=1, outlier.shape = NA) + 
  #scale_x_discrete(labels = rep(unique(all_images$bcell_patch_score), each = 3)) +
  geom_quasirandom(dodge.width=0.75, alpha=1, size=2) +
  #geom_line(aes(group = Description), alpha = ifelse(all_images$celltype != "HLA-DR", 0.6, 0), colour = "black") +
  #geom_point(aes(group= interaction), position = position_dodge(0.75)) +
  ylab("CXCL13 Expressing Cells per mm2\n(log10 + 1)") +
  xlab("") + 
  theme_bw() +
  theme(text = element_text(size=18),
        axis.text.x = element_text(angle=45, vjust=1, hjust=1)) +
  scale_fill_manual(values = unname(metadata(sce_rna)$colour_vectors$celltype[c("Tcytotoxic", "Tcell", "HLA-DR")]),
                    breaks = names(metadata(sce_rna)$colour_vectors$celltype[c("Tcytotoxic", "Tcell", "HLA-DR")])) +
  #scale_color_manual(values = unname(metadata(sce_rna)$colour_vectors$celltype[c("Tcytotoxic", "Tcell", "HLA-DR")]),
                    #breaks = names(metadata(sce_rna)$colour_vectors$celltype[c("Tcytotoxic", "Tcell", "HLA-DR")]),
                    #guide = FALSE) +
  guides(fill=guide_legend(title="Cell Type", override.aes = c(lwd=0.5)))
```

# Figure 5E

## Example of CXCL10 Cluster and corresponding Community

```{r Fig_5E_clusters_and_communities, fig.width=6, fig.height=6, dev=c('pdf')}
example <- findPatch(sce_prot[,sce_prot$Description == "C9"], sce_prot[,colData(sce_prot)$celltype %in% c("Bcell", "BnTcell")]$cellID, 
                    'cellID', 
                    'Center_X', 'Center_Y', 
                    'Description', 
                    distance = 15, 
                    min_clust_size = 20,
                    output_colname = "example_patch")

example <- findMilieu(example, 
              'cellID', 
              'Center_X', 'Center_Y', 
              'Description', 
              'example_patch', 
              distance = 50,
              output_colname = "example_milieu",
              plot = TRUE)
```

# Figure 5F

## Fraction of Tcf7 Tcells cells in bcell patch groups

```{r Fig_5F_cd8_tcf7_pd1_bcells, dev=("pdf"), fig.width=9, fig.height=4}
celltypes <- data.frame(colData(sce_prot)) %>%
  mutate(mmLocationPunch = paste(MM_location, Location, sep = "_")) %>%
  filter(mmLocationPunch != "LN_M") %>% # remove LN margin images 
  #filter(celltype != "Tumor") %>%
  mutate(TCF7_PD1 = paste(PD1, TCF7, sep = "_")) %>%
  group_by(PatientID,BlockID, Description, bcell_patch_score, celltype, TCF7_PD1) %>%
  summarise(n=n()) %>% 
  reshape2::dcast(PatientID + BlockID + Description + bcell_patch_score + celltype ~ TCF7_PD1, value.var = "n", fill=0) %>%
  reshape2::melt(id.vars = c("PatientID","BlockID", "Description", "bcell_patch_score", "celltype"), 
                 variable.name = "TCF7_PD1", value.name = "n") %>%
  group_by(Description, celltype) %>%
  mutate(fraction = n/sum(n)) %>%
  mutate(total_cells = sum(n)) %>%
  ungroup() %>%
  filter(celltype %in% c("Tcytotoxic", "Thelper")) 

celltypes$bcell_patch_score <- as.character(celltypes$bcell_patch_score)
celltypes$bcell_patch_score <- factor(celltypes$bcell_patch_score, levels = c("no B cells", "no B cell patches", "small B cell patches", "TLS"))

stat.test <- celltypes %>%
  group_by(celltype, TCF7_PD1) %>%
  kruskal_test(data = ., fraction ~ bcell_patch_score) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj") %>%
  mutate(group1 = celltype, group2 = TCF7_PD1) %>%
  add_xy_position()

ggplot(celltypes, aes(x=TCF7_PD1, y=fraction)) +
  geom_boxplot(alpha=1, lwd = 0.5, outlier.size = 0.5, aes(fill=bcell_patch_score)) +
  stat_pvalue_manual(x = "group2", stat.test, y.position = -0.1, size = 6) +
  theme_bw() +
  theme(text = element_text(size = 14),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) + 
  guides(fill=guide_legend(title="Bcell Score", override.aes = c(lwd=0.5))) +
  xlab("") + 
  ylab("Fraction of Population") +
  facet_wrap(~celltype, scales = "free") +
  ylim(-0.1,1.1)
```

# Figure 5G

## Percentage of TCF7+PD1+CD8+ cells that are part of a Bcell milieu

```{r Fig_5G_milieu_presence_tcells, fig.height=5, fig.width=8, dev=("pdf")}
# what fraction of each celltype is part of a milieu?
cur_dat <- data.frame(colData(sce_prot)) %>%
  filter(Location != "CTRL") %>%
  filter(celltype %in% c("Tcytotoxic", "Thelper")) %>%
  mutate(status = paste(TCF7, PD1, sep = "_")) %>%
  #filter(status == "TCF7+_PD1+") %>%
  mutate(milieu_binary = ifelse(bcell_milieu > 0, 1, 0)) %>%
  group_by(Description, milieu_binary, celltype, status) %>%
  summarise(n=n()) %>%
  group_by(Description, celltype, status) %>%
  mutate(fraction_in_milieu = n / sum(n)) %>%
  filter(milieu_binary == 1)

# what fraction of the total cell area is made up by cells that are part of a milieu
milieu_area <- data.frame(colData(sce_prot)) %>%
  filter(Location != "CTRL") %>%
  mutate(milieu_binary = ifelse(bcell_milieu > 0, 1, 0)) %>%
  group_by(Description, milieu_binary) %>%
  summarise(area = sum(Area)) %>% 
  group_by(Description) %>%
  mutate(fraction_of_area = area / sum(area))  %>%
  filter(row_number() == n())

sum <- left_join(cur_dat, milieu_area, by="Description")

# calculate metric - milieu_fraction divided by milieu-area-fraction (this normalizes the first metric)
sum$metric <- sum$fraction_in_milieu / sum$fraction_of_area

# one-sample t test
stat.test <- sum %>%
  group_by(celltype, status) %>%
  mutate(log10_metric = log10(metric)) %>%
  t_test(log10_metric ~ 1, mu = 0, alternative = "greater", detailed = TRUE) %>%
  adjust_pvalue() %>%
  add_significance("p.adj")

ggplot(sum, aes(x=sum$status, y=log10(metric))) +
  geom_boxplot(outlier.shape = NA) + 
  geom_quasirandom(size=0.75) +
  geom_hline(yintercept = 0) +
  facet_wrap(~celltype) +
  theme_bw() + 
  theme(text=element_text(size=14),
        axis.text.x = element_text(angle=45, vjust=1, hjust=1)) +
  stat_pvalue_manual(
    stat.test, x = "status", y.position = 1.5,
    label = "p.adj.signif",
    position = position_dodge(0.8), 
    size=7) + 
  ylab("Area-Normalized Log10 Fold-Change") +
  xlab("") #+
  #coord_cartesian(ylim=c(-1,1.75))
```
