---
title: "Supplementary Figure 12"
author: "Tobias Hoch and Daniel Schulz"
date: "2021-11-18"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction
This script generates plots for Supplementary Figure 12.

# Preparations

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message= FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Load libraries

```{r}
sapply(list.files("code/helper_functions", full.names = TRUE), source)
library(ggridges)
library(SingleCellExperiment)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggbeeswarm)
library(ggrastr)
library(Hmisc)
library(data.table)
library(ggpubr)
library(corrplot)
library(gridExtra)
library(scater)
library(dittoSeq)
library(ComplexHeatmap)
library(colorRamps)
library(cowplot)
library(stringr)
library(circlize)
```

## Load data

```{r load_data}
sce_rna <- readRDS(file = "data/data_for_analysis/sce_RNA.rds")
sce_prot <- readRDS(file = "data/data_for_analysis/sce_protein.rds")

# remove LN margin samples and control samples
sce_rna$PunchLocation <- paste(sce_rna$MM_location, sce_rna$Location, sep = "_")
sce_prot$PunchLocation <- paste(sce_prot$MM_location, sce_prot$Location, sep = "_")
sce_rna <- sce_rna[,sce_rna$PunchLocation != "LN_M" & sce_rna$Location != "CTRL"]
sce_prot <- sce_prot[,sce_prot$PunchLocation != "LN_M" & sce_prot$Location != "CTRL"]

# image
image_mat_prot <- read.csv("data/data_for_analysis/protein/Image.csv")
image_mat_rna <- read.csv("data/data_for_analysis/RNA/Image.csv")
```

## Prepare image size data

```{r}
im_size_prot <- as.data.frame(cbind(image_mat_prot$Metadata_Description, (image_mat_prot$Height_cellmask * image_mat_prot$Width_cellmask)/1000000))
names(im_size_prot) <- c("Description", "mm2_prot")
im_size_prot$mm2_prot <- as.numeric(im_size_prot$mm2_prot)
im_size_prot[im_size_prot$Description %in% c("G1", "G1 - split"), ]$mm2_prot <- mean(im_size_prot[im_size_prot$Description %in% c("G1", "G1 - split"), ]$mm2_prot)
im_size_prot <- im_size_prot[im_size_prot != "G1 - split",]
im_size_prot <- im_size_prot[1:166,]
```

## Prepare image size data

```{r}
im_size_rna <- as.data.frame(cbind(image_mat_rna$Metadata_Description, (image_mat_rna$Height_cellmask * image_mat_rna$Width_cellmask)/1000000))
names(im_size_rna) <- c("Description", "mm2_rna")
im_size_rna$mm2_rna <- as.numeric(im_size_rna$mm2_rna)
im_size_rna <- im_size_rna[1:166,]
```

# Supp Figure 12A/12B

## Diffusion Map for CXCL13 Producing Cells

```{r Fig_12A12B_DiffusionMap_producing_cells, fig.height=7, fig.width=11, dev=c('pdf')}
# loop through all patches 
for(i in c("cxcl13only_clust")){
  # subset sce object to only contain community cells
  sce_sub <- sce_rna[,colData(sce_rna)[,i] > 0]

  assay(sce_sub, "scaled_asinh") <- t(scale(t(assay(sce_sub, "asinh"))))
  
  # create UAMP
  set.seed(12345)
  sce_sub <- runDiffusionMap(sce_sub, 
                             exprs_values = "asinh", 
                             subset_row = rowData(sce_sub)$good_marker,
                             ncomponents = 2)
  
  # add patch size to sce
  cur_df <- data.frame(colData(sce_sub))

  clust_size <- cur_df %>%
    group_by(cur_df[,i]) %>%
    summarise(clust_size = n())
   
  names(clust_size)[1] <- i 
  
  cur_df <- left_join(cur_df, clust_size)
  sce_sub$clust_size = as.numeric(log10(cur_df$clust_size))
  
   # col by clust size
  a <- dittoDimPlot(sce_sub,
                    reduction.use = "DiffusionMap",
                    var = "clust_size", 
                    size = 1,
                    legend.show = TRUE,
                    opacity = 1,
                    max.color = "red", min.color = "blue",
                    main = NULL,
                    legend.title = "Patch Size (log10)") +
    xlim(quantile(reducedDim(sce_sub, "DiffusionMap")[,1], 0.05)[[1]],
         quantile(reducedDim(sce_sub, "DiffusionMap")[,1], 0.95)[[1]]) +
    ylim(quantile(reducedDim(sce_sub, "DiffusionMap")[,2], 0.05)[[1]],
         quantile(reducedDim(sce_sub, "DiffusionMap")[,2], 0.95)[[1]]) +
    theme_bw() +
    theme(text = element_text(size=18))

  # col by celltype
  b <- dittoDimPlot(sce_sub, 
             reduction.use = "DiffusionMap", 
             var = "celltype", 
             opacity = 1,
             color.panel = metadata(sce_sub)$colour_vector$celltype,
             size = 1,
             legend.show = TRUE,
                    main = NULL,
             legend.title = "Cell Type") +
    theme_bw() +
    xlim(quantile(reducedDim(sce_sub, "DiffusionMap")[,1], 0.05)[[1]],
         quantile(reducedDim(sce_sub, "DiffusionMap")[,1], 0.95)[[1]]) +
    ylim(quantile(reducedDim(sce_sub, "DiffusionMap")[,2], 0.05)[[1]],
         quantile(reducedDim(sce_sub, "DiffusionMap")[,2], 0.95)[[1]]) +
    theme(text = element_text(size=18)) + 
    guides(colour = guide_legend(override.aes = list(alpha = 1, size=3))) 
  
  leg_a <- cowplot::get_legend(a)
  leg_b <- cowplot::get_legend(b)

}

sce_sub <- sce_rna[,colData(sce_rna)[,"cxcl13only_clust"] > 0]

# add patch size to sce
cur_df <- data.frame(colData(sce_sub))

clust_size <- cur_df %>%
  group_by(cxcl13only_clust) %>%
  summarise(clust_size = n())
   
names(clust_size)[1] <- "cxcl13only_clust"
  
cur_df <- left_join(cur_df, clust_size)

sce_sub$clust_size = as.numeric(log10(cur_df$clust_size))

# add clust size correlation plot 
clust_size <- data.frame(colData(sce_sub)) %>%
  distinct(Description, cxcl13only_clust, .keep_all = T) %>%
  group_by(Description) %>%
  summarise(maxClustSize = max(clust_size))

Bcell_patch <- data.frame(colData(sce_prot)) %>%
  group_by(Description, bcell_patch) %>%
  summarise(n=n()) %>%
  mutate(n = ifelse(bcell_patch == 0, 0, n)) %>%
  mutate(maxPatchSize = log10(max(n+1))) %>%
  distinct(Description, .keep_all = T) %>% 
  select(Description, maxPatchSize)

Bcell_patch <- left_join(Bcell_patch, clust_size)
Bcell_patch[is.na(Bcell_patch$maxClustSize), ]$maxClustSize <- 0

Bcell <- data.frame(colData(sce_prot)) %>%
  group_by(Description, celltype) %>%
  summarise(n=n()) %>%
  mutate(fraction = n / sum(n)) %>%
  reshape2::dcast(Description ~ celltype, value.var = "fraction", fill = 0) %>%
  select(Description, `B cell`)

Bcell <- left_join(Bcell, clust_size)
Bcell[is.na(Bcell$maxClustSize), ]$maxClustSize <- 0

# only when not 0 - CHANGE?
c <- ggplot(Bcell_patch[rowSums(Bcell_patch[,-1]) > 0,], aes(x = maxClustSize, y = maxPatchSize, label=Description)) + 
  geom_point() +
  geom_smooth(method="lm") + 
  stat_cor(method = "pearson",
           aes(label = paste0("atop(", ..r.label..,  ",", ..p.label.. ,")")),
           size = 6, cor.coef.name = "R", label.sep="\n", label.y = 0.5, label.x = 2) + 
  ylab("Max Size of B cell\nPatches (log10)") +
  xlab("Max Size of CXCL13 Patches (log10)") +
  theme_bw() + 
  theme(text = element_text(size=18))

d <- ggplot(Bcell, aes(x = maxClustSize, y = `B cell`, label=Description)) + 
  geom_point() +
  geom_smooth(method="lm") + 
  stat_cor(method = "pearson",
           aes(label = paste0("atop(", ..r.label..,  ",", ..p.label.. ,")")),
           size = 6, cor.coef.name = "R", label.sep="\n", label.y = 0.05, label.x = 2) + 
  ylab("B Cell Fraction") +
  xlab("Max Size of CXCL13 Patches (log10)") +
  theme_bw() + 
  theme(text = element_text(size=18))

plot_grid(grid.arrange(a + theme(legend.position = "none"), 
                       b + theme(legend.position = "none"),
                       ncol = 2),
          grid.arrange(d,c, ncol = 2),
          ncol = 1, 
          rel_heights = c(0.65,0.35))
```

## Legend for Plot

```{r Supp_Fig_12A12B_Legend, dev=c('pdf'), fig.height=3, fig.width=3}
grid.arrange(leg_a)
grid.arrange(leg_b)
```

# Supp Figure 12C

## Prepare data 

```{r}
sce_rna$CD8pos_CXCL13 <- 0
sce_rna[,sce_rna$celltype %in% c("CD8+ T cell") & sce_rna$CXCL13 == 1]$CD8pos_CXCL13 <- 1

sce_rna$CD8neg_CXCL13 <- 0
sce_rna[,sce_rna$celltype %in% c("CD8- T cell") & sce_rna$CXCL13 == 1]$CD8neg_CXCL13 <- 1
```

## Density CD8+CXCL13+ and density B cells per Patch score

```{r}
perc_cd8_cxcl13 <- as.data.frame(colData(sce_rna)) %>%
  group_by(Description, CD8pos_CXCL13) %>%
  summarise(n_cd8pos=n()) %>%
  ungroup() %>%
  complete(Description, CD8pos_CXCL13, fill=list(n_cd8pos=0)) %>%
  filter(CD8pos_CXCL13 == 1)

perc_cd20 <- as.data.frame(colData(sce_prot)) %>%
  group_by(Description, celltype) %>%
  summarise(n_cd20=n()) %>%
  ungroup() %>%
  complete(Description, celltype, fill=list(n_cd20=0)) %>%
  filter(celltype == "B cell")

info <- as.data.frame(colData(sce_prot)) %>%
  distinct(Description, .keep_all = T) %>%
  select(Description, bcell_patch_score)

perc_cd20 <- left_join(info, perc_cd20)
data <- left_join(perc_cd20, perc_cd8_cxcl13)

data_cd8pos <- left_join(data, im_size_prot)
data_cd8pos <- left_join(data_cd8pos, im_size_rna)
data_cd8pos$n_cd20 <- data_cd8pos$n_cd20 / data_cd8pos$mm2_prot
data_cd8pos$n_cd8pos <- data_cd8pos$n_cd8pos / data_cd8pos$mm2_rna
```

## Density CD8-CXCL13+ and density B cells per Patch score

```{r}
perc_cd4_cxcl13 <- as.data.frame(colData(sce_rna)) %>%
  group_by(Description, CD8neg_CXCL13) %>%
  summarise(n_cd8neg=n()) %>%
  ungroup() %>%
  complete(Description, CD8neg_CXCL13, fill=list(n_cd8neg=0)) %>%
  filter(CD8neg_CXCL13 == 1)

perc_cd20 <- as.data.frame(colData(sce_prot)) %>%
  group_by(Description, celltype) %>%
  summarise(n_cd20=n()) %>%
  ungroup() %>%
  complete(Description, celltype, fill=list(n_cd20=0)) %>%
  filter(celltype == "B cell")

info <- as.data.frame(colData(sce_prot)) %>%
  distinct(Description, .keep_all = T) %>%
  select(Description, bcell_patch_score)

perc_cd20 <- left_join(info, perc_cd20)
data <- left_join(perc_cd20, perc_cd4_cxcl13)

data_cd8neg <- left_join(data, im_size_prot)
data_cd8neg <- left_join(data_cd8neg, im_size_rna)
data_cd8neg$n_cd20 <- data_cd8neg$n_cd20 / data_cd8neg$mm2_prot
data_cd8neg$n_cd8neg <- data_cd8neg$n_cd8neg / data_cd8neg$mm2_rna
```

## Plot Density CXCL13+ T cells and density B cells per Patch score

```{r Supp_Figure_12C_CXCL13TcellDensity, dev=("pdf"), fig.height=3.2, fig.width=9.4}
data_complete <- left_join(data_cd8neg, data_cd8pos[,c("Description","n_cd8pos")])

# remove images with follicles
data_complete <- data_complete[data_complete$bcell_patch_score != "B cell Follicles",]

a <- ggplot(data_complete, aes(x=n_cd8neg, y=n_cd20)) +
  geom_point() +
  geom_smooth(method="lm") +
  stat_cor(method = "pearson",
           aes(label = paste(..r.label.., ..p.label.., sep = "~`,`~")),
           size = 4, cor.coef.name = "R", label.sep="\n", label.y.npc = "top", label.x.npc = "left") +
  xlab("CXCL13+CD8- T cell / mm2") +
  ylab("B cells / mm2") +
  theme_bw() +
  theme(text=element_text(size=15))

b <- ggplot(data_complete, aes(x=n_cd8pos, y=n_cd20)) +
  geom_point() +
  geom_smooth(method="lm") +
  stat_cor(method = "pearson",
           aes(label = paste(..r.label.., ..p.label.., sep = "~`,`~")),
           size = 4, cor.coef.name = "R", label.sep="\n", label.y.npc = "top", label.x.npc = "left") +
  xlab("CXCL13+CD8+ T cell / mm2") +
  ylab("B cells / mm2") +
  theme_bw() +
  theme(text=element_text(size=15))

grid.arrange(a,b, ncol=2)
```

# Supp Figure 12D

## Chemokine Correlation with Tcf7+CD8+

```{r Supp_Fig_12D_chemokines_tcfpd1_correlation, fig.height=6.2, fig.width=10.1, dev=("pdf")}
targets <- metadata(sce_rna)$chemokines_morethan600_withcontrol

# top abundant chemokines
cur_rna <- data.frame(colData(sce_rna)) %>%
  filter(Location != "CTRL")

# protein data
cur_prot <- data.frame(colData(sce_prot)) %>%
  filter(Location != "CTRL")

# sum
rna_sum <- cur_rna %>%
  group_by(Description) %>%
  mutate(total_cells=n()) %>%
  ungroup() %>%
  group_by(Description, total_cells, expressor) %>%
  summarise(n=n()) %>%
  mutate(fraction=n/total_cells) %>%
  reshape2::dcast(Description ~ expressor, value.var = "fraction", fill = 0) 

# only keep highly abundant chemokines
rna_sum <- rna_sum[,c("Description", targets)]

prot_sum <- cur_prot %>%
  mutate(celltype2 = ifelse(celltype %in% c("CD4+ T cell", "CD8+ T cell"), 
                            paste(paste(celltype, TCF7, sep="_"), PD1, sep = "_"), celltype)) %>%
  group_by(Description, celltype2) %>%
  summarise(n = n()) %>%
  group_by(Description) %>%
  mutate(fraction = n/sum(n)) %>%
  #filter(!celltype2 %in% c("CD4+ T cell_TCF7+_PD1+", "CD4+ T cell_TCF7-_PD1+", "CD8+ T cell_TCF7+_PD1+")) %>%
  reshape2::dcast(Description ~ celltype2, value.var = "fraction", fill = 0) %>%
  select(Description,contains(c("CD8+ T cell", "CD4+ T cell")))

# equal images
all(rna_sum$Description == prot_sum$Description)

# correlation
cor <- cor(rna_sum[,-1], prot_sum[,-1], method = "pearson")

ha <- t(str_split_fixed(colnames(cor), "_", n=3))

dat_sum <- cur_prot %>%
  filter(celltype %in% c("CD4+ T cell", "CD8+ T cell")) %>%
  mutate(celltype2 = ifelse(celltype %in% c("CD4+ T cell", "CD8+ T cell"), 
                            paste(paste(celltype, TCF7, sep="_"), PD1, sep = "_"), celltype)) %>%
  group_by(celltype2) %>%
  summarise(n = n())


ha1 <- HeatmapAnnotation("TCF7_PD1" = anno_text(paste(ha[2,], ha[3,], sep = " ")),
                         "Cell Type" = ha[1,],
                         "Number of Cells" = anno_barplot(dat_sum$n,
                                                         height = unit(2,"cm"),
                                                         axis_param = list(gp = gpar(fontsize=14))),
                        "Numbers" = anno_text(dat_sum$n, 
                                              which = "column", 
                                              rot = 0,
                                              height = unit(0.5,"cm"),
                                              just = "center", 
                                              location = 0.5),
                         col = list("Cell Type" = metadata(sce_prot)$colour_vectors$celltype[c("CD4+ T cell", "CD8+ T cell")]),
                         show_legend = FALSE)

h <- Heatmap(cor,
        name = "Pearson\nCorrelation",
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        show_column_names = FALSE,
        show_row_names = TRUE,
        cell_fun = function(j, i, x, y, width, height, fill) {
          grid.text(sprintf("%.2f", cor[i, j]), x, y, gp = gpar(fontsize = 15, col = "black"))
          },
        col = colorRamp2(c(-1, 0, 1), c("red", "white", "blue")),
        row_title = "Expressor",
        row_names_side = "left",
        top_annotation = ha1,
        width = unit(18, "cm"),
        height = unit(10, "cm"),
        show_heatmap_legend = FALSE)

# draw heatmap
draw(h)
```

## Legend heatmap

```{r Supp_Fig_12D_legend_heatmap, dev="pdf", fig.height=3, fig.width=2}
lgd1 = color_mapping_legend(h@matrix_color_mapping, plot = FALSE, legend_direction = "vertical", legend_width=unit(3,"cm"), at = c(-1:1))
lgd2 = color_mapping_legend(ha1@anno_list$`Cell Type`@color_mapping, plot = FALSE, legend_direction = "vertical", nrow = 4)

lgd_list = packLegend(lgd1,lgd2,direction = "vertical", gap = unit(1,"cm"))
draw(lgd_list)
```
