---
title: "Supplementary Figure 1"
author: "Tobias Hoch and Daniel Schulz"
date: "2020-08-04"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction
This script generates plots for Supplementary Figure 1.

# Preparations

## Load libraries

```{r}
sapply(list.files("code/helper_functions", full.names = TRUE), source)
library(SingleCellExperiment)
library(data.table)
library(ggplot2)
library(dplyr)
library(stringr)
library(ggpubr)
library(ComplexHeatmap)
library(circlize)
library(gridExtra)
library(ggbeeswarm)
library(ggrepel)
library(ggpmisc)
library(ggrastr)
```

## Load and process the data

```{r}
# load data
cells1 = fread("data/data_for_analysis/12plex_validation/overexpression/20190305/cell.csv", header = T,sep=",")
cells2 = fread("data/data_for_analysis/12plex_validation/overexpression/20190306/cell.csv", header = T,sep=",")
meta1 = fread("data/data_for_analysis/12plex_validation/overexpression/20190305/Image.csv",header = T,sep=",")
meta2 = fread("data/data_for_analysis/12plex_validation/overexpression/20190306/Image.csv",header = T,sep=",")
panel = fread( "data/data_for_analysis/12plex_validation/overexpression/20190305/20190305_A431_overexpression.csv")

# extract replicate and stain info
meta1 = meta1[,.(ImageNumber,FileName_CellImage, Group_Number, Metadata_Target)]
meta1[,':=' (replicate = getInfoFromString(FileName_CellImage,"_",1)), by=ImageNumber]
meta2 = meta2[,.(ImageNumber,FileName_CellImage, Group_Number, Metadata_Target)]
meta2[,':=' (replicate = getInfoFromString(FileName_CellImage,"_",1)), by=ImageNumber]

# sort Metal Number in same order than in the Image.csv file
panel$`Metal Number` = str_extract(panel$`Metal Tag`, "[0-9]+")
panel = panel[order(panel$`Metal Number`),]
panel$channel = c(1:nrow(panel))

cells_long1 = melt.data.table(cells1,id.vars = c("ImageNumber", "ObjectNumber"),variable.factor = F)
cells_long2 = melt.data.table(cells2,id.vars = c("ImageNumber", "ObjectNumber"),variable.factor = F)

# create unique ID with Image and ObjectNumber
cells_long1[,id := paste(ImageNumber,ObjectNumber,sep ="_")]
cells_long2[,id := paste(ImageNumber,ObjectNumber,sep ="_")]

# multiply value by 2E16 since it divided by this number in CellProfiler
cells_long1$value = cells_long1$value * 65535
cells_long2$value = cells_long2$value * 65535

# calculate counts_asinh
cells_long1[,':=' (channel = as.integer(getInfoFromFileList(variable,sep="_",strPos = 4,censorStr = "c")),
                 counts_asinh = asinh(value/1)),
                 by=id]
cells_long2[,':=' (channel = as.integer(getInfoFromFileList(variable,sep="_",strPos = 4,censorStr = "c")),
                 counts_asinh = asinh(value/1)),
                 by=id]

# take only FullStack entries and not FullStackFiltered
cells_long1[,measurement:=getInfoFromString(variable,"_",3),by=variable]
cells_long1[,signal_type:=getInfoFromString(variable,"_",2),by=variable]
cells_long2[,measurement:=getInfoFromString(variable,"_",3),by=variable]
cells_long2[,signal_type:=getInfoFromString(variable,"_",2),by=variable]

unique(cells_long1$signal_type)
unique(cells_long2$signal_type)
```

## Merge meta data with cells data

```{r}
cells1 = merge(cells_long1, meta1,by="ImageNumber")
cells2 = merge(cells_long2, meta2,by="ImageNumber")

cells = rbind(cells1, cells2)
```

## Exclude DNA, Histone and panCytokeratin data and then get the info of PPIB staining

```{r}
# exclude some channels, make sure to exclude the right ones!
cells_panel = merge(cells,panel[,.(channel,Target, `Metal Tag`, `Metal Number`)],by="channel")
cells_panel = cells_panel[Target %in% c("T1","T2","T3","T4","T5","T6","T7","T8","T9","T10","T11","T12"),]
cells_panel = cells_panel[measurement != "FullStack",]
cells_panel = cells_panel[signal_type != "MeanIntensityCorrectedLS",]
unique(cells_panel$signal_type)

cells_panel_LScorrected = cells_panel[cells_panel$signal_type == "MeanIntensityCorrectedLS",]
cells_panel_uncorrected = cells_panel[cells_panel$signal_type == "MeanIntensity",]
cells_panel_corrected = cells_panel[cells_panel$signal_type == "MeanIntensityCorrected",]
```

# Supp Figure 1E

## Unspecific Amplifier Binding 

```{r Supp_Fig_1E_heatmap_unspecific_binding, dev=c('pdf')}
results = matrix(nrow=12, ncol = 12)
rownames(results) = unique(cells_panel$Target)
colnames(results) = unique(cells_panel$Target)
results = as.data.frame(results)

cells_panel_corrected = as.data.frame(cells_panel_corrected)

# calculate percentage of unspecific binding
for(i in unique(cells_panel$Target)){
  mean_target = mean(cells_panel_corrected[cells_panel_corrected["Metadata_Target"] == i & cells_panel_corrected["Target"] == i, "value"])
  for(j in unique(cells_panel$Target)){
    row_index = which(rownames(results) %in% i)
    col_index = which(colnames(results) %in% j)
    results[row_index, col_index] = mean(cells_panel_corrected[cells_panel_corrected["Metadata_Target"] == j & cells_panel_corrected["Target"] == i, "value"]) / mean_target * 100
  }
}

# set crosstalk to 0 from one channel to same channel
results[results == 100] <- 0

a = Heatmap(as.matrix(results),
        col = colorRamp2(c(0,1,3), c("white","blue", "red")),
        row_order = order(as.numeric(gsub("T", "", unique(cells_panel_corrected$Target)))),
        column_order =  order(as.numeric(gsub("T", "", unique(cells_panel_corrected$Target)))),
        heatmap_legend_param = list(title = "% cross-talk", size =15),
        column_names_side = "top",
        column_title = "Stained Channel",
        row_title = "Other Channels",
        row_names_side = "left",
        column_names_rot = 0,
        column_names_gp = gpar(fontsize = 15),
        row_names_gp = gpar(fontsize = 15),
        column_names_centered = T,
        cell_fun = function(j, i, x, y, width, height, fill) {
        grid.text(sprintf("%.1f", as.matrix(results)[i, j]), x, y, gp = gpar(fontsize = 10))
})
draw(a)
```

# Supp Figure 1D

## Boxplot with uncorrected and corrected signal

```{r Supp_Fig_1D_corrected signal, dev=("pdf"), fig.width=8, fig.height=5}
cells_panel2 <- cells_panel
cells_panel2[cells_panel2$signal_type == "MeanIntensity", ]$signal_type = "uncorrected"
cells_panel2[cells_panel2$signal_type == "MeanIntensityCorrected", ]$signal_type = "spill-over corrected"

ggplot(cells_panel2[cells_panel2$Metadata_Target == "T9",], aes(x=`Metal Tag`, y=counts_asinh, fill=signal_type)) +
  geom_boxplot(outlier.shape = NA, lwd=0.5) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        text = element_text(size=18)) +
  guides(fill=guide_legend(title="Signal")) +
  ylab("Mean Count (asinh)") +
  xlab("") +
  facet_wrap(~Metadata_Target,labeller = as_labeller(c(T9 = "mRNA probe for Channel 9 (Er168)")))
```

## Remove previous data 

```{r}
rm(list = ls())

# Reload helper functions
sapply(list.files("code/helper_functions", full.names = TRUE), source)
```

## Load the data

```{r}
cells1 = fread("data/data_for_analysis/12plex_validation/HeLa/20190208/cell.csv", header = T,sep=",")
cells3 = fread("data/data_for_analysis/12plex_validation/HeLa/20190215/cell.csv", header = T,sep=",")
cells4 = fread("data/data_for_analysis/12plex_validation/HeLa/20190222/cell.csv", header = T,sep=",")

meta1 = fread("data/data_for_analysis/12plex_validation/HeLa/20190208/Image.csv",header = T,sep=",")
meta3 = fread("data/data_for_analysis/12plex_validation/HeLa/20190215/Image.csv",header = T,sep=",")
meta4 = fread("data/data_for_analysis/12plex_validation/HeLa/20190222/Image.csv",header = T,sep=",")

meta1 = meta1[,.(ImageNumber,FileName_CellImage)]
meta3 = meta3[,.(ImageNumber,FileName_CellImage)]
meta4 = meta4[,.(ImageNumber,FileName_CellImage)]

meta1[,':=' (replicate = getInfoFromString(FileName_CellImage,"_",1),stain = getInfoFromString(FileName_CellImage,"_",6), 
             secondary_stain = getInfoFromString(FileName_CellImage,"_",7)),by=ImageNumber]
meta3[,':=' (replicate = getInfoFromString(FileName_CellImage,"_",1),stain = getInfoFromString(FileName_CellImage,"_",6), 
             secondary_stain = getInfoFromString(FileName_CellImage,"_",7)),by=ImageNumber]
meta4[,':=' (replicate = getInfoFromString(FileName_CellImage,"_",1),stain = getInfoFromString(FileName_CellImage,"_",7), 
             secondary_stain = getInfoFromString(FileName_CellImage,"_",8), 
             another_stain = getInfoFromString(FileName_CellImage,"_",6)),by=ImageNumber]

# rename positive stain 
meta1[meta1$stain=="positive" & meta1$secondary_stain == "noAb",]$stain = "positive without Ab"
meta1[meta1$stain=="positive" & meta1$secondary_stain == "withAb",]$stain = "positive with Ab"
#meta2[meta2$stain=="positive" & meta2$secondary_stain == "noAb",]$stain = "positive without Ab"
#meta2[meta2$stain=="positive" & meta2$secondary_stain == "withAb",]$stain = "positive with Ab"
meta3[meta3$stain=="positive" & meta3$secondary_stain == "noAb",]$stain = "positive without Ab"
meta3[meta3$stain=="positive" & meta3$secondary_stain == "withAb",]$stain = "positive with Ab"
meta3[meta3$stain=="negative" & meta3$secondary_stain == "s0",]$stain = "negative 1st"
meta3[meta3$stain=="negative" & meta3$secondary_stain == "new",]$stain = "negative"
meta4[meta4$stain=="noAb",]$stain = "positive without Ab"
meta4[meta4$stain=="withAb",]$stain = "positive with Ab"
meta4[meta4$another_stain=="negative",]$stain = "negative"
meta4[meta4$another_stain=="T12",]$stain = "T12"

# change stain name
meta3[meta3$stain=="T7new",]$stain = "T7"
meta3[meta3$stain=="T5new",]$stain = "T5"
meta3[meta3$stain=="T11new",]$stain = "T11"
meta3[meta3$replicate =="20190221",]$replicate = "20190215"

panel = fread( "data/data_for_analysis/12plex_validation/HeLa/20190212_HeLa_12plex_validation.csv")

# sort Metal Number in same order than in the Image.csv file!!
panel$`Metal Number` = str_extract(panel$`Metal Tag`, "[0-9]+")
panel = panel[order(panel$`Metal Number`),]
panel$channel = c(1:nrow(panel))

# melt table
cells1_long = melt.data.table(cells1,id.vars = c("ImageNumber", "ObjectNumber"),variable.factor = F)
cells3_long = melt.data.table(cells3,id.vars = c("ImageNumber", "ObjectNumber"),variable.factor = F)
cells4_long = melt.data.table(cells4,id.vars = c("ImageNumber", "ObjectNumber"),variable.factor = F)

# multiply value by 2E16 since it divided by this number in CellProfiler
cells1_long$value = cells1_long$value * 65535
cells3_long$value = cells3_long$value * 65535
cells4_long$value = cells4_long$value * 65535

cells1_long[,id := paste(ImageNumber,ObjectNumber,sep ="_")]
cells1_long[,':=' (channel = as.integer(getInfoFromFileList(variable,sep="_",strPos = 4,censorStr = "c")),
                 counts_asinh = asinh(value/1)),
                 by=id]

cells3_long[,id := paste(ImageNumber,ObjectNumber,sep ="_")]
cells3_long[,':=' (channel = as.integer(getInfoFromFileList(variable,sep="_",strPos = 4,censorStr = "c")),
                 counts_asinh = asinh(value/1)),
                 by=id]

cells4_long[,id := paste(ImageNumber,ObjectNumber,sep ="_")]
cells4_long[,':=' (channel = as.integer(getInfoFromFileList(variable,sep="_",strPos = 4,censorStr = "c")),
                 counts_asinh = asinh(value/1)),
                 by=id]

# take only FullStack entries and not FullStackFiltered (not possible with RNA measurement, only with Ab's)
cells1_long[,measurement:=getInfoFromString(variable,"_",3),by=variable]
cells3_long[,measurement:=getInfoFromString(variable,"_",3),by=variable]
cells4_long[,measurement:=getInfoFromString(variable,"_",3),by=variable]

cells1_long = cells1_long[measurement=="FullStack",]
cells3_long = cells3_long[measurement=="FullStack",]
cells4_long = cells4_long[measurement=="FullStack",]
```

## Merge meta dat with cells data and then then merge all the files to have one file with all three replicates

```{r}
cells1_long = merge(cells1_long,meta1,by="ImageNumber")
cells3_long = merge(cells3_long,meta3,by="ImageNumber")
cells4_long = merge(cells4_long,meta4,by="ImageNumber")

# select only 1 negative measuremenet in 3th measurement
cells3_long = cells3_long[cells3_long$stain != "negative 1st", ]

# remove additional columns which are not needed
cells1_long = cells1_long[,!("secondary_stain")]
cells3_long = cells3_long[,!("secondary_stain")]
cells4_long = cells4_long[,!("secondary_stain")]
cells4_long = cells4_long[,!("another_stain")]


cells = rbind(cells1_long, cells3_long)
cells = rbind(cells, cells4_long)
```

## Merge cells and panel data and exclude non-relevant channels

```{r}
# exclude certain channels, make sure to exclude the right ones!
cells_panel <- merge(cells,panel[,.(channel,Target)],by="channel")
cells_panel <- cells_panel[Target %in% c("T1","T2","T3","T4","T5","T6","T7","T8","T9","T10","T11","T12"),]
```

## Load data frolm 8plex validation and make it comparable

```{r}
# import data from 8plex validation
cells_panel_8plex = readRDS(file = "data/data_for_analysis/12plex_validation/HeLa/2018_data_8plex_validation_DS.rds")

# make both data.frames same so they can be merged
cells_panel = cells_panel[, !("measurement")]
cells_panel_8plex = cells_panel_8plex[, !("AreaShape_Area")]

# rename the stains in the 8plex data
unique(cells_panel_8plex$stain)
cells_panel_8plex = cells_panel_8plex[!(stain %in% ("oldMix")), ]
cells_panel_8plex[stain == "C1", ]$stain = "T1"
cells_panel_8plex[stain == "C2", ]$stain = "T4"
cells_panel_8plex[stain == "C3", ]$stain = "T3"
cells_panel_8plex[stain == "C4", ]$stain = "T8"
cells_panel_8plex[stain == "C5", ]$stain = "T12"
cells_panel_8plex[stain == "C6", ]$stain = "T2"
cells_panel_8plex[stain == "C7", ]$stain = "T9"
cells_panel_8plex[stain == "C8", ]$stain = "T6"
cells_panel_8plex[stain == "all", ]$stain = "positive with Ab"
cells_panel_8plex[stain == "neg", ]$stain = "negative"
cells_panel_8plex[Target == "C1", ]$Target = "T1"
cells_panel_8plex[Target == "C2", ]$Target = "T4"
cells_panel_8plex[Target == "C3", ]$Target = "T3"
cells_panel_8plex[Target == "C4", ]$Target = "T8"
cells_panel_8plex[Target == "C5", ]$Target = "T12"
cells_panel_8plex[Target == "C6", ]$Target = "T2"
cells_panel_8plex[Target == "C7", ]$Target = "T9"
cells_panel_8plex[Target == "C8", ]$Target = "T6"

# check
unique(cells_panel_8plex$stain)
unique(cells_panel_8plex$Target)
all(colnames(cells_panel) == colnames(cells_panel_8plex))

# merge both files
cells_panel_12plex = rbind(cells_panel, cells_panel_8plex)
```

# Supp Figure 1A

## Throw out the DNA, histone and panCytokeratin data and then get the info of PPIB staining

```{r Supp_Fig_1A_merged_comparison_per_channel, fig.width=10, fig.height=8, dev=c('pdf')}
# subset
cells_panel_12plex_sub = cells_panel_12plex[stain == Target,]

# relevel factor
cells_panel_12plex_sub$Target <- factor(cells_panel_12plex_sub$Target, levels = c("T1","T2","T3","T4","T5","T6","T7","T8","T9","T10","T11","T12"))

# check out difference in signal intensity for PPIB in the different channels
ggplot(data = cells_panel_12plex_sub[Target %in%c("T1","T2","T3","T4","T5","T6","T7","T8","T9","T10","T11","T12"),], aes(x=as.factor(Target),y=counts_asinh)) +
  geom_boxplot(fill="deepskyblue1")+
  theme_minimal()+
  theme(text = element_text(size=25), axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Mean Intensity Count per Cell [asinh]") + 
  xlab("Target Channels") + 
  geom_hline(yintercept = mean(cells_panel_12plex_sub$counts_asinh), linetype = 2) +
  scale_x_discrete(breaks=c("T1", "T2", "T3","T4","T5","T6","T7","T8", "T9", "T10","T11", "T12"),
                      labels= c("T1 PPIB","T2 PPIB","T3 PPIB", "T4 PPIB", "T5 PPIB","T6 PPIB","T7 PPIB","T8 PPIB","T9 PPIB","T10 PPIB", "T11 PPIB", "T12 PPIB")) 
```

## Signal range

```{r}
cells_panel_12plex_sub %>%
  group_by(stain) %>%
  summarize(mean_per_channel = mean(counts_asinh)) %>%
  mutate(sd = sd(mean_per_channel), mean = mean(mean_per_channel), min = min(mean_per_channel), max = max(mean_per_channel))
```

# Supp Figure 1C

## Influence of Antibody Incubation

```{r Supp_Fig_1C_overnight_Ab_incubation, fig.width=10, fig.height=8, dev=c('pdf')}
dd = cells_panel[stain == "positive without Ab" | stain == "positive with Ab",]

# relevel factor
dd$Target <- factor(dd$Target, levels = c("T1","T2","T3","T4","T5","T6","T7","T8","T9","T10","T11","T12"))
dd$stain <- ifelse(dd$stain == "positive with Ab", "+ Antibodies", "- Antibodies")

# plot
ggplot(data=dd, aes(x=Target, y=counts_asinh, fill = stain)) + 
  geom_boxplot() +
  scale_x_discrete(breaks=c("T1", "T2", "T3","T4","T5","T6","T7","T8", "T9", "T11","T10", "T12"),
                      labels= c("T1 POLR2A","T2 PPIB","T3 UBC", "T4 HPRT1", "T5 TUB","T6 RPL28","T7 RPL5","T8 B2M","T9 ACTB","T10 LDHA", "T11 RPLP0", "T12 GAPDH")) +
  theme_minimal()+
  theme(text = element_text(size=25), axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ylab("Mean Intensity Count per Cell [asinh]") +
  xlab("Target Channels") + 
  scale_fill_discrete(name = "Protocol Type") 
```

# Supp Figure 1B

## Compare Positive Controls with RNAseq data

```{r Supp_Fig_1B_correlation_with_RNAseq, fig.width=8, fig.height=7, dev=c('pdf')}
# add gene name
cells_panel_12plex$gene <- ""
cells_panel_12plex[cells_panel_12plex$Target == "T1",]$gene <- "POLR2A"
cells_panel_12plex[cells_panel_12plex$Target == "T2",]$gene <- "PPIB"
cells_panel_12plex[cells_panel_12plex$Target == "T3",]$gene <- "UBC"
cells_panel_12plex[cells_panel_12plex$Target == "T4",]$gene <- "HPRT1"
cells_panel_12plex[cells_panel_12plex$Target == "T5",]$gene <- "TUB"
cells_panel_12plex[cells_panel_12plex$Target == "T6",]$gene <- "RPL28"
cells_panel_12plex[cells_panel_12plex$Target == "T7",]$gene <- "RPL5"
cells_panel_12plex[cells_panel_12plex$Target == "T8",]$gene <- "B2M"
cells_panel_12plex[cells_panel_12plex$Target == "T9",]$gene <- "ACTB"
cells_panel_12plex[cells_panel_12plex$Target == "T10",]$gene <- "LDHA"
cells_panel_12plex[cells_panel_12plex$Target == "T11",]$gene <- "RPLP0"
cells_panel_12plex[cells_panel_12plex$Target == "T12",]$gene <- "GAPDH"

median_per_gene <- cells_panel_12plex %>%
  filter(stain == "positive with Ab") %>%
  filter(replicate %in% c("20190208", "20190215", "20190222")) %>%
  group_by(gene) %>%
  summarise(expression_RNAScope = median(counts_asinh))

# data from human protein atlas
rna_seq <- data.frame(median_per_gene$gene)
colnames(rna_seq) <- "gene"

# Normalized Expression Data from Human Protein Atlas 
rna_seq$expression_RNAseq <- c(145.3, 89.2, 217.1, 21.3, 74.9, 1.2, 54.7, 87.9, 83.6, 72.4, 23.9, 62.9)

# merge
sum <- left_join(median_per_gene, rna_seq)

# plot 
ggplot(sum, aes(x=log2(expression_RNAseq), y=expression_RNAScope, label=gene)) + 
  geom_label_repel(size=7, fill="deepskyblue1") + 
  geom_smooth(method = "lm") +
  xlab("Normalized Expression RNA-Seq [log2]") +
  ylab("Median Expression RNAScope [asinh]") + 
  stat_cor(method = "pearson",
           aes(label = paste0("atop(", ..r.label..,  ",", ..rr.label.. ,")")),
           size = 10, cor.coef.name = "R", label.sep="\n", label.y.npc = "top") + 
  theme_minimal() + 
  theme(text = element_text(size=20)) 
```

## Correlation

```{r}
# correlation of the two technologies
cor(sum$expression_RNAScope, sum$expression_RNAseq, method = c("spearman"))
```

# Supp Figure 1F

## Load Data

```{r}
sce_rna <- readRDS("data/data_for_analysis/sce_RNA.rds")
```

## Detection of chemokine expressing cells

for the detection of chemokine expressing cells we make use of the fact that we also measured a negative control (DapB).

```{r}
# get the names of the chemokine channels without the negative control channel
chemokine_channels = rownames(sce_rna[which(grepl("T\\d+_",rownames(sce_rna)) & ! grepl("DapB",rownames(sce_rna))),])
chemokine_channels_sub <- c("T2_CCL22")

# run function to define chemokine expressing cells 
output_list <- compute_difference(sce_rna, 
                          cellID = "cellID", 
                          assay_name = "asinh", 
                          threshold = 0.01, 
                          mRNA_channels = chemokine_channels_sub, 
                          negative_control = "T6_DapB", 
                          return_calc_metrics = TRUE)
```

## Plot Results from Chemokine Detection

```{r Supp_Fig_1F_chemokine_detection,fig.height=7, fig.width=17, dev=c("pdf"), dpi=25}
# check difference between DapB and signal (histogram)
plot_list = list()
for(i in chemokine_channels_sub){
  
  # subset whole data set for visualization purposes
  diff_chemo <- output_list[[i]]
  diff_chemo_sub <- diff_chemo[sample(nrow(diff_chemo), nrow(diff_chemo)*0.5), ]

  a <- ggplot(data = diff_chemo_sub, aes(x=scaled_diff)) + 
  geom_histogram(binwidth = 0.05, aes(fill = 
                                       ifelse(padj <= 0.01 & scaled_diff > 0, 'p<0.01', 'n.s.'))) + 
  xlab(paste(paste("Scaled Difference ", i, sep = " "), " - DapB", sep = "")) + 
    xlim(-5,7) +
    theme_minimal() + 
    theme(text = element_text(size=20),
          legend.position = "none") + 
    scale_fill_manual(values = c("black", "deepskyblue1"))
  
  # significant cells defined by subtraction
  b <- ggplot(data=diff_chemo_sub, aes(x=mean_negative_control, y=mean_chemokine)) + 
    geom_point_rast(alpha=0.2, aes(col = 
                                ifelse(padj <= 0.01 & scaled_diff > 0, 'p<0.01', 'n.s.'))) + 
    scale_color_manual(values = c("black", "deepskyblue1"), 
                       name = "Legend") +
    guides(color = guide_legend(override.aes = list(alpha=1, size=3))) +
    xlim(0,5.5) + ylim(0,5.5) +
    ylab(paste("Mean expression of", i, sep=" ")) +
    xlab("Mean DapB mRNA expression") +
    theme_minimal() + 
    theme(text = element_text(size=20)) 

  
  grid.arrange(a,b, nrow = 1, ncol=2)
}
```
