---
title: "Supplementary Figure 5"
author: "Tobias Hoch and Daniel Schulz"
date: "2021-01-30"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction
This script generates plots for Figure 3.

# Preparations

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message= FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Load libraries

First, we will load the libraries needed for this part of the analysis.

```{r load-libraries, message=FALSE}
sapply(list.files("code/helper_functions", full.names = TRUE), source)
library(cytomapper)
library(SingleCellExperiment)
library(reshape2)
library(tidyverse)
library(dplyr)
library(data.table) 
library(ggplot2)
library(colorRamps)
library(RColorBrewer)
library(gridExtra)
library(ggpmisc)
library(ComplexHeatmap)
library(scater)
library(dittoSeq)
library(ggbeeswarm)
library(corrplot)
library(ggpubr)
library(cowplot)
library(circlize)
library(ggrepel)
library(rstatix)
```

## Read the data

```{r load data}
sce_rna = readRDS(file = "data/data_for_analysis/sce_RNA.rds")
sce_prot = readRDS(file = "data/data_for_analysis/sce_protein.rds")

# image
image_mat_prot <- read.csv("data/data_for_analysis/protein/Image.csv")

# surv_dat
dat_survival_prot <- fread(file = "data/data_for_analysis/protein/clinical_data_protein.csv")
```

## Prepare image size data

```{r}
im_size <- as.data.frame(cbind(image_mat_prot$Metadata_Description, (image_mat_prot$Height_cellmask * image_mat_prot$Width_cellmask)/1000000))
names(im_size) <- c("Description", "mm2")
im_size$mm2 <- as.numeric(im_size$mm2)
im_size[im_size$Description %in% c("G1", "G1 - split"), ]$mm2 <- mean(im_size[im_size$Description %in% c("G1", "G1 - split"), ]$mm2)
im_size <- im_size[im_size != "G1 - split",]
```

# Supp Figure 5A/5B

## Diffusion Map for CXCL13 Producing Cells

```{r Fig_5A5B_DiffusionMap_producing_cells, fig.height=7, fig.width=11, dev=c('pdf')}
# loop through all patches 
for(i in c("cxcl13only_clust")){
  # subset sce object to only contain community cells
  sce_sub <- sce_rna[,colData(sce_rna)[,i] > 0]

  assay(sce_sub, "scaled_asinh") <- t(scale(t(assay(sce_sub, "asinh"))))
  
  # create UAMP
  set.seed(12345)
  sce_sub <- runDiffusionMap(sce_sub, 
                             exprs_values = "asinh", 
                             subset_row = rowData(sce_sub)$good_marker,
                             ncomponents = 2)
  
  # add patch size to sce
  cur_df <- data.frame(colData(sce_sub))

  clust_size <- cur_df %>%
    group_by(cur_df[,i]) %>%
    summarise(clust_size = n())
   
  names(clust_size)[1] <- i 
  
  cur_df <- left_join(cur_df, clust_size)
  sce_sub$clust_size = as.numeric(log10(cur_df$clust_size))
  
   # col by clust size
  a <- dittoDimPlot(sce_sub,
                    reduction.use = "DiffusionMap",
                    var = "clust_size", 
                    size = 1,
                    legend.show = TRUE,
                    opacity = 1,
                    max.color = "red", min.color = "blue",
                    main = NULL,
                    legend.title = "Patch Size (log10)") +
    xlim(quantile(reducedDim(sce_sub, "DiffusionMap")[,1], 0.01)[[1]],
         quantile(reducedDim(sce_sub, "DiffusionMap")[,1], 0.99)[[1]]) +
    ylim(quantile(reducedDim(sce_sub, "DiffusionMap")[,2], 0.01)[[1]],
         quantile(reducedDim(sce_sub, "DiffusionMap")[,2], 0.99)[[1]]) +
    theme_bw() +
    theme(text = element_text(size=18))

  # col by celltype
  b <- dittoDimPlot(sce_sub, 
             reduction.use = "DiffusionMap", 
             var = "celltype", 
             opacity = 1,
             color.panel = metadata(sce_sub)$colour_vector$celltype,
             size = 1,
             legend.show = TRUE,
                    main = NULL,
             legend.title = "Cell Type") +
    theme_bw() +
    xlim(quantile(reducedDim(sce_sub, "DiffusionMap")[,1], 0.01)[[1]],
         quantile(reducedDim(sce_sub, "DiffusionMap")[,1], 0.99)[[1]]) +
    ylim(quantile(reducedDim(sce_sub, "DiffusionMap")[,2], 0.01)[[1]],
         quantile(reducedDim(sce_sub, "DiffusionMap")[,2], 0.99)[[1]]) +
    theme(text = element_text(size=18)) + 
    guides(colour = guide_legend(override.aes = list(alpha = 1, size=3))) 
  
  leg_a <- cowplot::get_legend(a)
  leg_b <- cowplot::get_legend(b)

}

sce_sub <- sce_rna[,colData(sce_rna)[,"cxcl13only_clust"] > 0]

# add patch size to sce
cur_df <- data.frame(colData(sce_sub))

clust_size <- cur_df %>%
  group_by(cxcl13only_clust) %>%
  summarise(clust_size = n())
   
names(clust_size)[1] <- "cxcl13only_clust"
  
cur_df <- left_join(cur_df, clust_size)

sce_sub$clust_size = as.numeric(log10(cur_df$clust_size))

# add clust size correlation plot 
clust_size <- data.frame(colData(sce_sub)) %>%
  distinct(Description, cxcl13only_clust, .keep_all = T) %>%
  group_by(Description) %>%
  summarise(maxClustSize = max(clust_size))

Bcell_patch <- data.frame(colData(sce_prot)) %>%
  group_by(Description, bcell_patch) %>%
  summarise(n=n()) %>%
  mutate(n = ifelse(bcell_patch == 0, 0, n)) %>%
  mutate(maxPatchSize = log10(max(n+1))) %>%
  distinct(Description, .keep_all = T) %>% 
  select(Description, maxPatchSize)

Bcell_patch <- left_join(Bcell_patch, clust_size)
Bcell_patch[is.na(Bcell_patch$maxClustSize), ]$maxClustSize <- 0

Bcell <- data.frame(colData(sce_prot)) %>%
  group_by(Description, celltype) %>%
  summarise(n=n()) %>%
  mutate(fraction = n / sum(n)) %>%
  reshape2::dcast(Description ~ celltype, value.var = "fraction", fill = 0) %>%
  select(Description, Bcell)

Bcell <- left_join(Bcell, clust_size)
Bcell[is.na(Bcell$maxClustSize), ]$maxClustSize <- 0

# only when not 0 - CHANGE?
c <- ggplot(Bcell_patch[rowSums(Bcell_patch[,-1]) > 0,], aes(x = maxClustSize, y = maxPatchSize, label=Description)) + 
  geom_point() +
  geom_smooth(method="lm") + 
  stat_poly_eq(formula = y ~ x, 
                aes(label =  ..rr.label..), 
                parse = TRUE, size=10) +
  ylab("Max Size of Bcell\nPatches (log10)") +
  xlab("Max Size of CXCL13 Patches (log10)") +
  theme_bw() + 
  theme(text = element_text(size=18))

d <- ggplot(Bcell, aes(x = maxClustSize, y =Bcell, label=Description)) + 
  geom_point() +
  geom_smooth(method="lm") + 
  stat_poly_eq(formula = y ~ x, 
                aes(label =  ..rr.label..), 
                parse = TRUE, size=10) +
  ylab("B Cell Fraction") +
  xlab("Max Size of CXCL13 Patches (log10)") +
  theme_bw() + 
  theme(text = element_text(size=18))

# add clust size correlation plot 
clust_size <- data.frame(colData(sce_sub)) %>%
  distinct(Description, cxcl13only_clust, .keep_all = T) %>%
  group_by(Description) %>%
  summarise(maxClustSize = max(clust_size))

Bcell <- data.frame(colData(sce_prot)) %>%
  group_by(Description, celltype) %>%
  summarise(n=n()) %>%
  mutate(fraction = n / sum(n)) %>%
  reshape2::dcast(Description ~ celltype, value.var = "fraction", fill = 0) %>%
  select(Description, Bcell)

Bcell <- left_join(Bcell, clust_size)
Bcell[is.na(Bcell$maxClustSize), ]$maxClustSize <- 0

plot_grid(grid.arrange(a + theme(legend.position = "none"), 
                       b + theme(legend.position = "none"),
                       ncol = 2),
          grid.arrange(d,c, ncol = 2),
          ncol = 1, 
          rel_heights = c(0.65,0.35))
```

## Legend for Plot

```{r Supp_Fig_5A5B_Legend, dev=c('pdf'), fig.height=3, fig.width=3}
grid.arrange(leg_a)
grid.arrange(leg_b)
```

# Supp Figure 5C

## Tcf7+ correlation with Tcf7+PD1+ in Tcells

```{r, Supp_Fig_5C_tcf7_pd1_tcell_corr, fig.width=10, fig.height=4, dev=("pdf")}
cur_dat <- data.frame(colData(sce_prot)) %>%
  mutate(mmLocationPunch = paste(MM_location, Location, sep = "_")) %>%
  filter(mmLocationPunch != "LN_M") %>% # remove LN margin images 
  filter(Location != "CTRL") %>%
  mutate(TCF7_PD1 = paste(TCF7, PD1, sep = "_")) %>%
  group_by(Description, bcell_patch_score, celltype, TCF7_PD1) %>%
  summarise(n=n()) %>% 
  reshape2::dcast(Description + bcell_patch_score + celltype ~ TCF7_PD1, value.var = "n", fill=0) %>%
  reshape2::melt(id.vars = c("Description", "bcell_patch_score", "celltype"), 
                 variable.name = "TCF7_PD1", value.name = "n") %>%
  group_by(Description, celltype) %>%
  mutate(fraction = n/sum(n)) %>%
  mutate(total_cells = sum(n)) %>%
  ungroup() %>%
  filter(celltype %in% c("Tcytotoxic", "Thelper")) 

cur_dat <- left_join(cur_dat, im_size)

cur_dat$density <- cur_dat$n / cur_dat$mm2

cur_dat <- cur_dat %>%
  filter(TCF7_PD1 %in% c("TCF7+_PD1+", "TCF7+_PD1-")) %>%
  reshape2::dcast(Description + bcell_patch_score + celltype ~ TCF7_PD1, value.var = "density", fill=0)

ggplot(cur_dat, aes(x=log10(`TCF7+_PD1+`+1), y=log10(`TCF7+_PD1-`+1))) + 
  geom_point(size=3) + 
  geom_smooth(method="lm") +
  stat_poly_eq(formula = y ~ x, 
                aes(label =  ..rr.label..), 
                parse = TRUE, size=10, label.y = 3) +
  theme_bw() +
  theme(text = element_text(size=16)) +
  xlab("TCF7+PD1+ Cells per mm2 (log10+1)") +
  ylab("TCF7+PD1- Cells per mm2 (log10+1)") +
  facet_wrap(~celltype, scales = "free")
```

# Supp Figure 5D

## Response / Survival Analysis

```{r}
# select blocks in ICI subgroup that are treatment-naive and non-adjuvant
blocks <- unique(sce_prot[,sce_prot$treatment_status_before_surgery == "naive" & 
                            sce_prot$treatment_group_after_surgery == "ICI" &
                            sce_prot$Adjuvant == "n"]$BlockID)

# subset survival data (only data points that contain info for 3m response)
dat_survival_prot_sub <- dat_survival_prot[dat_survival_prot$BlockID %in% blocks & 
                                             is.na(dat_survival_prot$Status_at_3m) == FALSE, ]

# remove LN margin samples
dat_survival_prot_sub <- dat_survival_prot_sub %>%
  mutate(mmLocationPunch = paste(MM_location, Location, sep = "_")) %>%
  filter(mmLocationPunch != "LN_M")

# group sizes
dat_survival_prot_sub %>%
  distinct(Internal_pat_ID, .keep_all = T) %>%
  group_by(Status_at_3m) %>%
  summarise(n=n())
```

## TCF7/PD1 fraction in responder vs. non responder

```{r, Supp_Fig_5D_responder_nonresponder_tcf7pd1, fig.height=8, fig.width=10, dev=("pdf")}
cur_dat <- data.frame(colData(sce_prot[,sce_prot$BlockID %in% unique(dat_survival_prot_sub$BlockID)])) %>%
  mutate(mmLocationPunch = paste(MM_location, Location, sep = "_")) %>%
  filter(mmLocationPunch != "LN_M") %>% # remove LN margin images 
  #filter(celltype != "Tumor") %>%
  mutate(TCF7_PD1 = paste(TCF7, PD1, sep = "_")) %>%
  group_by(PatientID,BlockID, Description, bcell_patch_score, celltype, TCF7_PD1) %>%
  summarise(n=n()) %>% 
  reshape2::dcast(PatientID + BlockID + Description + bcell_patch_score + celltype ~ TCF7_PD1, value.var = "n", fill=0) %>%
  reshape2::melt(id.vars = c("PatientID","BlockID", "Description", "bcell_patch_score", "celltype"), 
                 variable.name = "TCF7_PD1", value.name = "n") %>%
  group_by(Description, celltype) %>%
  mutate(fraction = n/sum(n)) %>%
  mutate(total_cells = sum(n)) %>%
  ungroup() %>%
  filter(celltype %in% c("Tcytotoxic", "Thelper")) 

cur_dat <- left_join(cur_dat, dat_survival_prot_sub[,c("BlockID", "Status_at_3m", "Primary_melanoma_type")]) %>%
  distinct(Description, celltype, TCF7_PD1, .keep_all = T)

cur_dat <- left_join(cur_dat, im_size[im_size$Description %in% unique(cur_dat$Description),])
cur_dat$n_per_mm2 <- cur_dat$n / cur_dat$mm2
cur_dat$celltype2 <- paste(cur_dat$celltype, cur_dat$TCF7_PD1, sep = ", ")

stat.test <- cur_dat %>%
  mutate(log10_n_per_mm2 = log10(n_per_mm2 + 1)) %>%
  group_by(celltype2) %>%
  wilcox_test(data = ., n_per_mm2 ~ Status_at_3m) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj")

info <- cur_dat[,c("celltype2", "TCF7_PD1")] %>%
  distinct(celltype2, .keep_all = T)

# filter relevant tests
stat.test <- left_join(stat.test, info) %>%
  filter(TCF7_PD1 %in% c("TCF7+_PD1+", "TCF7+_PD1-"))

stat.test$p_and_p.adj <- paste(paste("p = ", round(stat.test$p,2), sep = ""), paste("p.adj = ", round(stat.test$p.adj,2), sep = ""), sep = "\n")

ggplot(cur_dat[cur_dat$TCF7_PD1 %in% c("TCF7+_PD1+", "TCF7+_PD1-"),], aes(x=Status_at_3m, y=log10(n_per_mm2+1), label=PatientID)) + 
  geom_boxplot(alpha=.4, outlier.shape = NA, aes(fill=Status_at_3m)) +
  geom_beeswarm(size=3) +
  geom_label_repel(aes(label=PatientID), max.overlaps = 12, size = 5, alpha=.5) + 
  #stat_compare_means(label.x.npc = "left", label.y = -0.9, size=6, label = "p.format") +
  stat_pvalue_manual(stat.test, label = "p_and_p.adj", size=5, label.x.npc = "left", y.position = -0.9, bracket.size = 0) +
  theme_bw() +
  theme(text = element_text(size=18)) +
  ylab("Cells per mm2 (log10)") +
  xlab("") +
  guides(fill=guide_legend(title="Response 3m", override.aes = c(lwd=0.5, size=1))) +
  facet_wrap(~celltype2, scales = "free")
```

# Supp Figure 5E

## Chemokine Correlation with Tcf7+CD8+

```{r Supp_Fig_5E_chemokines_tcfpd1_correlation, fig.height=5.3, fig.width=9.8, dev=("pdf")}
targets <- metadata(sce_rna)$chemokines_morethan600_withcontrol

# top abundant chemokines
cur_rna <- data.frame(colData(sce_rna)) %>%
  filter(Location != "CTRL")

# protein data
cur_prot <- data.frame(colData(sce_prot)) %>%
  filter(Location != "CTRL")

# sum
rna_sum <- cur_rna %>%
  group_by(Description) %>%
  mutate(total_cells=n()) %>%
  ungroup() %>%
  group_by(Description, total_cells, expressor) %>%
  summarise(n=n()) %>%
  mutate(fraction=n/total_cells) %>%
  reshape2::dcast(Description ~ expressor, value.var = "fraction", fill = 0) 

# only keep highly abundant chemokines
rna_sum <- rna_sum[,c("Description", targets)]

prot_sum <- cur_prot %>%
  mutate(celltype2 = ifelse(celltype %in% c("Thelper", "Tcytotoxic"), 
                            paste(paste(celltype, TCF7, sep="_"), PD1, sep = "_"), celltype)) %>%
  group_by(Description, celltype2) %>%
  summarise(n = n()) %>%
  group_by(Description) %>%
  mutate(fraction = n/sum(n)) %>%
  filter(!celltype2 %in% c("Thelper_TCF7+_PD1+", "Thelper_TCF7-_PD1+", "Tcytotoxic_TCF7+_PD1+")) %>%
  reshape2::dcast(Description ~ celltype2, value.var = "fraction", fill = 0) %>%
  select(Description,contains(c("Tcytotoxic", "Thelper")))

# equal images
all(rna_sum$Description == prot_sum$Description)

# correlation
cor <- cor(rna_sum[,-1], prot_sum[,-1], method = "pearson")

ha <- t(str_split_fixed(colnames(cor), "_", n=3))

ha1 <- HeatmapAnnotation("TCF7_PD1" = anno_text(paste(ha[2,], ha[3,], sep = " ")),
                         "Cell Type" = ha[1,],
                         col = list("Cell Type" = metadata(sce_prot)$colour_vectors$celltype[c("Thelper", "Tcytotoxic")]),
                         show_legend = FALSE)

h <- Heatmap(cor,
        name = "Pearson\nCorrelation",
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        show_column_names = FALSE,
        show_row_names = TRUE,
        cell_fun = function(j, i, x, y, width, height, fill) {
          grid.text(sprintf("%.2f", cor[i, j]), x, y, gp = gpar(fontsize = 15, col = "black"))
          },
        col = colorRamp2(c(-1, 0, 1), c("red", "white", "blue")),
        row_title = "Expressor",
        row_names_side = "left",
        top_annotation = ha1,
        width = unit(18, "cm"),
        height = unit(10, "cm"),
        show_heatmap_legend = FALSE)

# draw heatmap
draw(h)
```

## Legend heatmap

```{r Supp_Fig_5E_legend_heatmap, dev="pdf", fig.height=3, fig.width=2}
lgd1 = color_mapping_legend(h@matrix_color_mapping, plot = FALSE, legend_direction = "vertical", legend_width=unit(3,"cm"), at = c(-1:1))
lgd2 = color_mapping_legend(ha1@anno_list$`Cell Type`@color_mapping, plot = FALSE, legend_direction = "vertical", nrow = 4)

lgd_list = packLegend(lgd1,lgd2,direction = "vertical", gap = unit(1,"cm"))
draw(lgd_list)
```

