---
title: "Supplementary Figure 5"
author: "Tobias Hoch and Daniel Schulz"
date: "2020-11-18"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction
This script generates plots for Supplementary Figure 5.

# Preparations

## Load libraries

```{r, message=FALSE}
sapply(list.files("code/helper_functions", full.names = TRUE), source)
library(SingleCellExperiment)
library(reshape2)
library(tidyverse)
library(dplyr)
library(data.table) 
library(ggplot2)
library(ComplexHeatmap)
library(rms)
library(ggrepel)
library(ggbeeswarm)
library(circlize)
library(ggpubr)
library(ggridges)
library(gridExtra)
library(rstatix)
library(cowplot)
library(ggrastr)
```

## Read the data

```{r}
# clinical data
dat <- read_csv("data/data_for_analysis/protein/clinical_data_protein.csv")

# Protein data

sce_prot <- readRDS("data/data_for_analysis/sce_protein.rds")
sce_prot <- sce_prot[,sce_prot$Location != "CTRL"]

# take panel_meta because there are the channel names from the shiny output (and only there, not in the SCE)
panel_meta_prot <- read.csv(file = "data/data_for_analysis/protein/melanoma_1.06_protein.csv", 
                       sep= ",",  stringsAsFactors = FALSE )


# RNA data

sce_rna <- readRDS("data/data_for_analysis/sce_rna.rds")
sce_rna <- sce_rna[,sce_rna$Location != "CTRL"]

# take panel_meta because there are the channel names from the shiny output (and only there, not in the SCE)
panel_meta_rna <- read.csv(file = "data/data_for_analysis/rna/panel_mat.csv", stringsAsFactors = FALSE )
```

# Supp Figure 5A

## Clinical features of the cohort 

Note: as the cohort is very diverse, we are using the BlockID as the minimal unit since clinical parameters are described per BlockID. However, sometimes we do have patients of which we have multiple FFPE blocks (BlockIDs). Nonetheless, clinical parameters are not given per patient but per patient FFPE block and are therefore considered the minimial unit.

## Number of Samples

```{r Supp_Fig_5A_number_of_samples, fig.height=6, fig.width=17, dev=c('pdf')}
dat[dat$BlockID %in% unique(sce_prot[,sce_prot$Location == "CTRL"]$BlockID),]$MM_location <- "Control"

# remove control samples
dat <- dat[dat$BlockID %in% unique(sce_prot[,sce_prot$Location != "CTRL"]$BlockID),]

p1 <- unique(dat[,c("BlockID","MM_location")]) %>%
  ggplot()+
  geom_bar(aes(y=MM_location),stat ="count") +
  xlab("Biopsy Blocks per Location") +
  ylab("Metastasis Location") +
  theme_bw()+
  theme(text = element_text(size=16))
  
p2 <- dat %>%
  ggplot()+
  geom_bar(aes(x=BlockID, fill=(MM_location)),stat="count")+
  theme_bw()+
  theme(axis.text.x = element_blank(),
        axis.ticks.x=element_blank(),
        text = element_text(size=16)) + 
  ylab("Number of Samples") +
  xlab("Biopsy Blocks") +
  scale_y_continuous(limits = c(0,4), expand = c(0, 0)) +
  guides(fill=guide_legend(title="Metasis Location"))

plot_grid(p1,p2,rel_widths = c(1.25,3))  
```

# Supp Figure 5B

## Select one random file per celltype and load file

```{r}
# load all subsetted sce object from hierarchichal gating and combine the
label.files <- list.files("data/data_for_analysis/protein/celltype_classifier", full.names = TRUE)
file_names <- data.frame(path = label.files)
file_names$fileName <- sub(".*/", "", label.files)
file_names$celltype <- sub("\\_.*", "", file_names$fileName)

# select one random file per celltype
file_names <- file_names %>%
  group_by(celltype) %>%
  sample_n(1)

# Read in SCE objects
cur_sces <- lapply(file_names$path, readRDS)
```

## Print gating scheme - Protein

```{r Supp_Fig_5B_gating_scheme_protein, fig.height=8, fig.width=6, dev=("pdf")}
for(element in cur_sces){
  # current labelled sce object
  sce_label <- element
  cur_celltype <- unique(sce_label$cytomapper_CellLabel)
  # select all gates from metadata
  gates <- metadata(sce_label)[grepl("cytomapper_gate_", names(metadata(sce_label)))]
  cur_img <- gates[[1]]$img_id
  # subset sce to current image 
  cur_sce <- sce_prot[,sce_prot$ImageNumber == cur_img]
  plots <- list()
  
  for(i in names(gates)) {
    plots[[i]] <- local({
      # select gates and imageID
      cur_gate <- as.data.frame(gates[[i]]$gate)
      cur_img <- gates[[i]]$img_id
      
      # names of channels (complicated because shiny has different channel names than SCE)
      x_gate <- rownames(cur_gate)[1]
      y_gate <- rownames(cur_gate)[2]
      x_metal <- panel_meta_prot[panel_meta_prot$clean_target == x_gate, "Metal.Tag"]
      y_metal <- panel_meta_prot[panel_meta_prot$clean_target == y_gate, "Metal.Tag"]
      x_original <- rownames(rowData(cur_sce)[rowData(cur_sce)[,"Metal.Tag"] == x_metal,])
      y_original <- rownames(rowData(cur_sce)[rowData(cur_sce)[,"Metal.Tag"] == y_metal,])
      
      # select current channels
      cur_counts <- as.data.frame(t(assay(cur_sce, "asinh")[c(x_original, y_original),]))
      
      # check if cells are in gate
      cur_counts$in_gate <- ifelse(cur_counts[,1] >= cur_gate[1,1] &
                                     cur_counts[,1] <= cur_gate[1,2] &
                                     cur_counts[,2] >= cur_gate[2,1] &
                                     cur_counts[,2] <= cur_gate[2,2], TRUE, FALSE)
      
      # cellIDs of cells in gate (for next gate)
      cur_cells <- rownames(cur_counts[cur_counts$in_gate == TRUE,])
      
      # update cur_sce
      cur_sce <<- cur_sce[,cur_sce$cellID %in% cur_cells] ####### !!!!
      
      # plot
      p <- ggplot() +
        geom_point_rast(data=cur_counts, aes(x=cur_counts[,1], y=cur_counts[,2], 
                                        color=ifelse(cur_counts[,"in_gate"] == TRUE, "red", "black")),
                   alpha=ifelse(cur_counts[,"in_gate"] == TRUE, 0.5,0.1)) +
        geom_rect(data = cur_gate, aes(xmin=cur_gate[1,1], xmax=cur_gate[1,2], ymin=cur_gate[2,1],
                                       ymax=cur_gate[2,2]),
                  color="black", alpha=0.2) +
        xlab(x_gate) +
        ylab(y_gate) +
        scale_color_identity() +
        theme_bw() +
        theme(legend.position = "none") +
        coord_cartesian(xlim = c(0,6.5), ylim = c(0,6.5))
      p
    })
  }
  
  n <- length(plots)
  nCol <- floor(sqrt(n))
  do.call("grid.arrange", c(plots, ncol=3,nrow=4, top=cur_celltype))
}
```

# Supp Figure 5C

## Select one random file per celltype and load file

```{r}
# load all subsetted sce object from hierarchichal gating and combine the
label.files <- list.files("data/data_for_analysis/rna/celltype_classifier", full.names = TRUE)
file_names <- data.frame(path = label.files)
file_names$fileName <- sub(".*/", "", label.files)
file_names$celltype <- sub("\\_.*", "", file_names$fileName)

# select one random file per celltype
file_names <- file_names %>%
  group_by(celltype) %>%
  sample_n(1)

# Read in SCE objects
cur_sces <- lapply(file_names$path, readRDS)
```

## Print gating scheme

```{r Supp_Fig_5C_gating_scheme_rna, fig.height=8, fig.width=6, dev=("pdf")}
for(element in cur_sces){
  # current labelled sce object
  sce_label <- element
  cur_celltype <- unique(sce_label$cytomapper_CellLabel)
  # select all gates from metadata
  gates <- metadata(sce_label)[grepl("cytomapper_gate_", names(metadata(sce_label)))]
  cur_img <- gates[[1]]$img_id
  # subset sce to current image 
  cur_sce <- sce_rna[,sce_rna$ImageNumber == cur_img]
  plots <- list()
  
  for(i in names(gates)) {
    plots[[i]] <- local({
      # select gates and imageID
      cur_gate <- as.data.frame(gates[[i]]$gate)
      cur_img <- gates[[i]]$img_id
      
      # names of channels (complicated because shiny has different channel names than SCE)
      x_gate <- rownames(cur_gate)[1]
      x_metal <- panel_meta_rna[panel_meta_rna$clean_target == x_gate, "Metal.Tag"]
      x_original <- rownames(rowData(cur_sce)[rowData(cur_sce)[,"Metal.Tag"] == x_metal,])
      
      if(nrow(gates[[i]]$gate) == 2) {
        y_gate <- rownames(cur_gate)[2]
        y_metal <- panel_meta_rna[panel_meta_rna$clean_target == y_gate, "Metal.Tag"]
        y_original <- rownames(rowData(cur_sce)[rowData(cur_sce)[,"Metal.Tag"] == y_metal,])
        
        # select current channels
        cur_counts <- as.data.frame(t(assay(cur_sce, "asinh")[c(x_original, y_original),]))
        
        # check if cells are in gate
        cur_counts$in_gate <- ifelse(cur_counts[,1] >= cur_gate[1,1] &
                                       cur_counts[,1] <= cur_gate[1,2] &
                                       cur_counts[,2] >= cur_gate[2,1] &
                                       cur_counts[,2] <= cur_gate[2,2], TRUE, FALSE)
        
        # cellIDs of cells in gate (for next gate)
        cur_cells <- rownames(cur_counts[cur_counts$in_gate == TRUE,])
        
        # update cur_sce
        cur_sce <<- cur_sce[,cur_sce$cellID %in% cur_cells]
        
        # plot
        p <- ggplot() +
          geom_point_rast(data=cur_counts, aes(x=cur_counts[,1], y=cur_counts[,2], 
                                               color=ifelse(cur_counts[,"in_gate"] == TRUE, "red", "black")),
                          alpha=ifelse(cur_counts[,"in_gate"] == TRUE, 0.5,0.1)) +
          geom_rect(data = cur_gate, aes(xmin=cur_gate[1,1], xmax=cur_gate[1,2], ymin=cur_gate[2,1],
                                         ymax=cur_gate[2,2]),
                    color="black", alpha=0.2) +
          xlab(x_gate) +
          ylab(y_gate) +
          scale_color_identity() +
          theme_bw() +
          theme(legend.position = "none") +
          coord_cartesian(xlim = c(0,8), ylim = c(0,8))
        p
      }
      else {
        # select current channels
        cur_counts <- as.data.frame(assay(cur_sce, "asinh")[c(x_original),])
        
        # check if cells are in gate
        cur_counts$in_gate <- ifelse(cur_counts[,1] >= cur_gate[1,1] &
                                       cur_counts[,1] <= cur_gate[1,2], TRUE, FALSE)
        
        # cellIDs of cells in gate (for next gate)
        cur_cells <- rownames(cur_counts[cur_counts$in_gate == TRUE,])
        
        # update cur_sce
        cur_sce <<- cur_sce[,cur_sce$cellID %in% cur_cells]
        
        # plot
        p <- ggplot() +
          geom_jitter_rast(data=cur_counts, aes(x=rownames(cur_gate), y=cur_counts[,1], 
                                                color=ifelse(cur_counts[,"in_gate"] == TRUE, "red", "black")),
                           alpha=ifelse(cur_counts[,"in_gate"] == TRUE, 0.5,0.1)) +
          geom_hline(yintercept = cur_gate[1,1], color="black", alpha=1) +
          geom_hline(yintercept = cur_gate[1,2], color="black", alpha=1) +
          xlab(x_gate) +
          ylab("Expression (asinh)") +
          scale_color_identity() +
          theme_bw() +
          theme(legend.position = "none",
                axis.text.x = element_blank()) +
          coord_cartesian(xlim = c(0,8), ylim = c(0,8))
        p
      }
    })
  }
  
  n <- length(plots)
  nCol <- floor(sqrt(n))
  do.call("grid.arrange", c(plots, ncol=3, nrow=4, top=cur_celltype))
}
```
