---
title: "clinical data summary"
author: "daniels"
date: "9 4 2020"
output:
  html_document: 
    self_contained: no
  pdf_document: default
---

# Introduction
This script gives an overview over the clinical data and generates statistics used in the manuscript.

# Preparations

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message= FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Load libraries

```{r load libraries}
sapply(list.files("code/helper_functions", full.names = TRUE), source)
library(data.table)
library(survival)
library(ggplot2)
library(broom)
library(dplyr)
library(RColorBrewer)
library(ggalluvial)
library(tidyverse)
library(cowplot)
```

## Load Data

```{r}
# clinical data
dat <- read_csv("data/protein/clinical_data_protein.csv")

# SCE object
sce = readRDS(file = "data/sce_protein.rds")
```

# Statistics

## Number of Samples and Patients

```{r}
# Number of Samples without control samples
data.frame(colData(sce)) %>%
  filter(Location != "CTRL") %>%
  distinct(Description) %>%
  summarise(n=n())

# Number of Patients
data.frame(colData(sce)) %>%
  filter(Location != "CTRL") %>%
  distinct(PatientID) %>%
  summarise(n=n())
```


# Analysis

## Clinical features of the cohort

Note: as the cohort is very diverse, we are using the BlockID as the minimal unit since clinical parameters are described per BlockID. However, sometimes we do have patients of which we have multiple FFPE blocks (BlockIDs). Nonetheless, clinical parameters are not given per patient but per patient FFPE block and are therefore considered the minimial unit.

```{r metastasis location by BlockID, fig.height=8, fig.width=15}
dat[dat$BlockID %in% unique(sce[,sce$Location == "CTRL"]$BlockID),]$MM_location <- "Control"

# remove control samples
dat <- dat[dat$BlockID %in% unique(sce[,sce$Location != "CTRL"]$BlockID),]

p1 <- unique(dat[,c("BlockID","MM_location")]) %>%
  ggplot()+
  geom_bar(aes(y=MM_location),stat ="count") +
  xlab("BlockIDs per Location") +
  ylab("Metastasis Location") +
  theme(text = element_text(size=18))
  
p2 <- dat %>%
  ggplot()+
  geom_bar(aes(x=BlockID, fill=(MM_location)),stat="count")+
  theme(axis.text.x = element_text(angle = 90,vjust = 0.5)) + 
  ylab("Number of Samples") +
  guides(fill=guide_legend(title="Metasis Location")) +
  theme(text = element_text(size=18))

plot_grid(p1,p2,rel_widths = c(1,3))  
```

```{r MM_location with cancer stage and punch position, fig.height=10,fig.width=10}
cur_df <- dat %>%
  select(BlockID,Nr_treatments_before_surgery,MM_location,Treatment_after_surgery, Status_at_3m, Date_death,Location,Cancer_Stage, E_I_D)
cur_df <- unique(cur_df)
cur_df$Death <- "yes"
cur_df[which(is.na(cur_df$Date_death)),]$Death <- "no"

cur_df %>%
  dplyr::count(MM_location,Cancer_Stage,Location,E_I_D) %>%
  ggplot(aes(axis1 = MM_location, axis2 = Cancer_Stage, axis3 = Location,axis4 = E_I_D,y=n))+
  geom_alluvium(aes(fill=as.factor(MM_location)))+
  geom_stratum()+
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  theme_minimal()+
  scale_x_discrete(limits = c("met location", "Cancer Stage", "Punch position", "Global Infiltration"), expand = c(.2, .05))
```

```{r whole cohort colored by MM_location,fig.width=10,fig.height=7}
cur_df <- cur_df%>%
  select(!Location) %>%
  unique()

cur_df %>%
  dplyr::count(MM_location, Nr_treatments_before_surgery, Treatment_after_surgery,Status_at_3m) %>%
  ggplot(aes(axis1 = MM_location, axis2 = Nr_treatments_before_surgery, axis3 = Treatment_after_surgery,y=n))+
  geom_alluvium(aes(fill=as.factor(MM_location)))+
  geom_stratum()+
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  theme_minimal()+
  scale_x_discrete(limits = c("met location", "# prevous treatments", "Therapy"), expand = c(.2, .05))

```

```{r colored by death,fig.width=10,fig.height=7}
cur_df %>%
  dplyr::count(MM_location, Nr_treatments_before_surgery, Cancer_Stage,Death) %>%
  ggplot(aes(axis1 = MM_location, axis2 = Nr_treatments_before_surgery, axis3 = Cancer_Stage,axis4 = Death,y=n))+
  geom_alluvium(aes(fill=as.factor(Death)))+
  geom_stratum()+
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  theme_minimal()+
  scale_x_discrete(limits = c("met location", "# prevous treatments", "Therapy"), expand = c(.2, .05))
```

```{r whole cohort colored by MM_location sorted by Nr treatments,fig.width=10,fig.height=7}
cur_df %>%
  dplyr::count(MM_location, Nr_treatments_before_surgery, Treatment_after_surgery,Status_at_3m) %>%
  ggplot(aes(axis2 = MM_location, axis1 = Nr_treatments_before_surgery, axis3 = Treatment_after_surgery, axis4 = Status_at_3m,y=n))+
  geom_alluvium(aes(fill=as.factor(MM_location)))+
  geom_stratum()+
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  theme_minimal()+
  scale_x_discrete(limits = c( "# prevous treatments","met location", "Therapy", "response"), expand = c(.2, .05))

```


```{r whole cohort colored by Nr_treatments sorted by Nr treatments,fig.width=10,fig.height=7}
cur_df %>%
  dplyr::count(MM_location, Nr_treatments_before_surgery, Treatment_after_surgery,Status_at_3m) %>%
  ggplot(aes(axis2 = MM_location, axis1 = Nr_treatments_before_surgery, axis3 = Treatment_after_surgery, axis4 = Status_at_3m,y=n))+
  geom_alluvium(aes(fill=as.factor(Nr_treatments_before_surgery)))+
  geom_stratum()+
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  theme_minimal()+
  scale_x_discrete(limits = c( "# prevous treatments","met location", "Therapy", "reponse"), expand = c(.2, .05))

```

```{r untreated samples colored by MM_location,fig.width=10,fig.height=7}
cur_df %>%
  filter(Nr_treatments_before_surgery == 0)%>%
  dplyr::count(MM_location, Treatment_after_surgery, Status_at_3m,Death) %>%
  ggplot(aes(axis2 = Treatment_after_surgery, axis1 = MM_location,axis4 = Status_at_3m, axis5 =Death,y=n))+
  geom_alluvium(aes(fill=as.factor(MM_location)))+
  geom_stratum()+
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  theme_minimal()+
  scale_x_discrete(limits = c("met location", "Therapy", "response","Death"), expand = c(.2, .05))+
  ggtitle("subsetted to untreated samples")
```

```{r untreataed samples colored by Therapy,fig.width=10,fig.height=7}
cur_df %>%
  filter(Nr_treatments_before_surgery == 0)%>%
  dplyr::count(MM_location, Treatment_after_surgery, Status_at_3m,Death) %>%
  ggplot(aes(axis2 = Treatment_after_surgery, axis1 = MM_location,axis4 = Status_at_3m, axis5 =Death,y=n))+
  geom_alluvium(aes(fill=as.factor(Treatment_after_surgery)))+
  geom_stratum()+
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  theme_minimal()+
  scale_x_discrete(limits = c("met location", "Therapy", "response","Death"), expand = c(.2, .05))+
  ggtitle("subsetted to untreated samples")
```



