---
title: "XX_RNA_Protein_comparison"
author: "toobiwankenobi"
date: "2020-08-25"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction
Here, we define chemokine-expressing cells 

# Preparations

## Load libraries

```{r, message=F}
library(SingleCellExperiment)
library(ggplot2)
library(ggpmisc)
library(dplyr)
library(gridExtra)
```

## Load the single cell experiment object and the image metadata

```{r}
sce_rna <- readRDS(file = "data/sce_RNA.rds")
sce_prot <- readRDS(file = "data/sce_protein.rds")
```

## Comparison Plot 

```{r}
sub_sce_rna <- NULL
sub_sce_prot <- NULL
sub_sce_rna <- sce_rna[,colData(sce_rna)$celltype %in% c("Neutrophil")]
sub_sce_prot <- sce_prot[,colData(sce_prot)$celltype %in% c("Neutrophil")]

#sub_sce_rna <- sce_rna
#sub_sce_prot <- sce_prot

cur_rna <- data.frame(colData(sub_sce_rna))
cur_prot <- data.frame(colData(sub_sce_prot))

# cells per image - RNA
cells_image_rna <- cur_rna %>%
  group_by(Description) %>%
  mutate(cells=n()) %>%
  mutate(mean_cell_area = mean(Area)) %>%
  distinct(Description, cells, mean_cell_area)

# cells per image - Protein
cells_image_prot <- cur_prot %>%
  group_by(Description) %>%
  mutate(cells=n()) %>%
  mutate(mean_cell_area = mean(Area)) %>%
  distinct(Description, cells, mean_cell_area)

# data set id
cells_image_rna$data_set <- "RNA"
cells_image_prot$data_set <- "Protein"

cells_image <- rbind(cells_image_prot, cells_image_rna)

# measured area RNA and protein
image_mat_rna <- read.csv(file = "data/rna/Image.csv", stringsAsFactors = FALSE)
image_mat_prot <- read.csv(file = "data/protein/Image.csv", stringsAsFactors = FALSE)

# diff in ablated area
area_prot <- data.frame(image_area = image_mat_prot$Height_FullStack * image_mat_prot$Width_FullStack)
area_prot$Description <- image_mat_prot$Metadata_Description
area_prot$data_set <- "Protein"

area_rna <- data.frame(image_area = image_mat_rna$Height_FullStack * image_mat_rna$Width_FullStack)
area_rna$Description <- image_mat_rna$Metadata_Description
area_rna$data_set <- "RNA"

image_area <- rbind(area_rna, area_prot)

# long format summary
summary <- merge(cells_image, image_area)

# Plot Cell Numbers
a <- ggplot(data = summary, aes(x=interaction("cells",data_set),y=log10(cells))) + 
  geom_boxplot(aes(fill=data_set)) + 
  geom_point(aes(group=data_set), alpha =0.3) + 
  geom_line(aes(group=Description),alpha=0.2) + 
  scale_x_discrete(labels = rep("cells", each = 2))

# Plot Cell Numbers
b <- ggplot(data = summary, aes(x=interaction("cells",data_set),y=log10(mean_cell_area))) + 
  geom_boxplot(aes(fill=data_set)) + 
  geom_point(aes(group=data_set), alpha =0.3) + 
  geom_line(aes(group=Description),alpha=0.2) + 
  scale_x_discrete(labels = rep("mean_cell_area", each = 2))

c <- ggplot(data = summary, aes(x=interaction("cells",data_set),y=log10(image_area))) + 
  geom_boxplot(aes(fill=data_set)) + 
  geom_point(aes(group=data_set), alpha =0.3) + 
  geom_line(aes(group=Description),alpha=0.2) + 
  scale_x_discrete(labels = rep("total_image_size", each = 2))

grid.arrange(a,b,c, nrow=1)
```

## Compare cell density in both data sets (cells per 1000µm2)

```{r}
summary[,-1] %>%
  group_by(data_set) %>%
  summarise_each(funs(sum)) %>%
  mutate(cells_per_area = cells/image_area * (1000*1000))
```

## Cell Type Correlation between the two data sets
Which celltypes can we identify in both data sets?

- Tcytotoxic
- Macrophages
- Tumor cells
- Neutrophils

Other comparisons (HLA-DR/CD38 vs. B cells) might not be valid. 

```{r}
rna_sum <- cur_rna[cur_rna$celltype %in% c("Tcytotoxic", "Macrophage", "Tumor", "Neutrophil"),] %>%
  group_by(Description, celltype) %>%
  summarise(n = n()) %>%
  reshape2::dcast(Description ~ celltype, value.var = "n", fill = 0)

prot_sum <- cur_prot[cur_prot$celltype %in% c("Tcytotoxic", "Macrophage", "Tumor", "Neutrophil"),] %>%
  group_by(Description, celltype) %>%
  summarise(n = n()) %>%
  reshape2::dcast(Description ~ celltype, value.var = "n", fill = 0)

mean(c(cor(rna_sum$Macrophage, prot_sum$Macrophage, method = "pearson"), 
       cor(rna_sum$Neutrophil, prot_sum$Neutrophil, method = "pearson"), 
       cor(rna_sum$Tcytotoxic, prot_sum$Tcytotoxic, method = "pearson"), 
       cor(rna_sum$Tumor, prot_sum$Tumor, method = "pearson")))
```

## Correlation of all celltypes

What are CD38 cells in the RNA set? They don't really correlate with anything, hence, they don't seem to be of any value in the RNA cohort. 

```{r}
rna_sum <- cur_rna %>%
  group_by(Description, celltype) %>%
  summarise(n = n()) %>%
  reshape2::dcast(Description ~ celltype, value.var = "n", fill = 0)

prot_sum <- cur_prot %>%
  group_by(Description, celltype) %>%
  summarise(n = n()) %>%
  reshape2::dcast(Description ~ celltype, value.var = "n", fill = 0)

corrplot::corrplot(cor(rna_sum[,-1], prot_sum[,-1], method = "pearson"), method = "number")
```


